---
title: "Cellranger Mapping QC"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction
## Objective
In this Rmarkdown document, we are going to perform a quality control (QC) analysis of the mapping results obtained by running [cellranger version 6.1.1]("https://support.10xgenomics.com/single-cell-gene-expression/software/pipelines/latest/what-is-cell-ranger").

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(gt)
library(ggpubr)
library(ggrepel)
library(scales)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r warning = FALSE, message = FALSE}
# Paths
path_project_metadata <- here::here("01_cellranger_mapping/data/FIXnCUT_metadata.csv")
path_cellranger_metrics_gex <- here::here("01_cellranger_mapping/results/cellranger_mapping_metrics_GEX.csv")

# Functions
source(here::here("bin/utils.R"))
```

## Load data
The data used in this Rmarkdown document comes from **FIXnCUT** project.
```{r}
# Load metadata
metadata <- read.csv(path_project_metadata)
DT::datatable(metadata, options = list(scrollX = TRUE))

# Load cellranger metrics GEX
cellranger_metrics_gex <- read_csv(path_cellranger_metrics_gex, col_names = TRUE)
metadata_to_merge <- subset(metadata, type == "cDNA")
cellranger_metrics_gex <- merge(metadata_to_merge, cellranger_metrics_gex)
```

## Pre-process data
```{r}
cellranger_metrics_gex_df <- clean_cellranger_metrics_df(cellranger_metrics_gex)
cellranger_metrics_gex_df[,16:46] <- lapply(cellranger_metrics_gex_df[,16:46], FUN=as.numeric)
# cellranger_metrics_gex_df <- cellranger_metrics_gex_df[, order(cellranger_metrics_gex_df$tissue,
#                                                        cellranger_metrics_gex_df$library_name)]
```


# Gene Expression QC
We will start by showing the three most relevant metrics (number of reads, estimated number of recovered cells, fraction of reads in cells, mean reads per cell, fraction of reads mapped to exonic reads, and median genes per cell) obtained by cellranger for each of the working libraries. This information will give us an idea of the quality of the experiment as well as the sequencing and the mapping steps.

```{r fig.width=14}
table_cellranger_metrics_gex(metrics_gex_df = cellranger_metrics_gex_df,
                             cellranger_version = "CellRanger v6.1.1")
```


## Sequencing QC
First, we will assess the quality of the sequenced libraries before the mapping step. To so do, we will use the "Q30" variables in our dataset, as these report the fraction of the bases with a Q-score >30 for different sequences (barcodes, reads and UMIs).
Q-score is calculated in the following way:
$$Q = -10\log10(p)$$
where *p* is the probability of the base being wrongly called. Thus, bases with a high Q-score are reliable.
```{r fig.height = 11, fig.width = 14}
gg_lib_size <- cellranger_metrics_gex_df %>%
  mutate(Number_of_Reads_mil = lib__Number_of_reads / 1000000) %>%
  ggplot(aes(x = library_name, y = Number_of_Reads_mil, fill = sample_protocol)) +
  geom_bar(stat = "identity") +
  facet_grid(~tissue, scales = "free_x", space = "free") +
  theme_bw() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Library size",
    x = "Libraries", 
    y =  "Number Reads (in millions)") +
  theme(axis.title = element_text(size = 12),
        axis.text = element_text(size = 8),
          axis.text.x = element_text(hjust = 1, angle = 45),
        strip.placement = "outside",
        strip.background = element_rect(colour = NA),
        legend.position = "bottom") +
  guides(fill = guide_legend(override.aes = list(size = 5), ncol = 3))

qc_seq_vars <- c("lib__Q30_barcodes",
                 "lib__Q30_RNA_read",
                 "lib__Q30_UMI")

gg_qc_seq <- purrr::map(qc_seq_vars, function(var) {
  ggplot(cellranger_metrics_gex_df, aes_string(x = "library_name", y = var, fill = "sample_protocol")) +
    geom_col() +
    facet_grid(~tissue, scales = "free_x", space = "free") +
    theme_bw() +
    scale_fill_brewer(palette = "Set2") +
    ylim(0, 100) +
    labs(title = str_replace_all(var, "_", " "),
         x = "Libraries",
         y = "Q30 Bases (%)") +
    theme(axis.title = element_text(size = 12),
          axis.text = element_text(size = 8),
          axis.text.x = element_text(hjust = 1, angle = 45),
          strip.placement = "outside",
          strip.background = element_rect(colour = NA),
          legend.position = "none")
})

ggarrange(plotlist = c(list(gg_lib_size), gg_qc_seq), nrow = 2, ncol = 2)
```

## Mapping QC
Next, we will check the quality of the mapping step performed by `cellranger 5.0.0` across libraries. To do so, we will compare the percentage of reads mapped to the genome, and within these mapped reads, the amount of reads mapped to intergenic regions, intronic and exonic regions. We aim to obtain libraries with a high percentage of confidently mapped reads, and specially a high percentage of exonic reads, which correspond with gene expression or RNAs. The reads mapping to intergenic regions suggest contamination of ambient DNA, whereas reads mapping to intronic regions may come from pre-mRNAs or mature spliced isoforms that retain certain introns.

```{r fig.height = 11, fig.width = 14}
qc_map_vars <- c("lib__Confidently_mapped_to_genome",
                 "lib__Confidently_mapped_antisense",
                 "lib__Confidently_mapped_to_intergenic_regions",
                 "lib__Confidently_mapped_to_intronic_regions",
                 "lib__Confidently_mapped_to_exonic_regions",
                 "lib__Confidently_mapped_to_transcriptome"
                 )

gg_qc_map <- purrr::map(qc_map_vars, function(var) {
  ggplot(cellranger_metrics_gex_df, aes_string(x = "library_name", y = var, fill = "sample_protocol")) +
    geom_col() +
    facet_grid(~tissue, scales = "free_x", space = "free") +
    theme_bw() +
    scale_fill_brewer(palette = "Set2") +
    ylim(0, 100) +
    labs(title = str_replace_all(var, "_", " "),
         x = "Libraries",
         y = "Reads Mapped (%)") +
    theme(axis.title = element_text(size = 12),
          axis.text = element_text(size = 10),
          axis.text.x = element_text(hjust = 1, angle = 45),
          strip.placement = "outside",
          strip.background = element_rect(colour = NA),
          legend.position = "none")
})

ggarrange(plotlist = gg_qc_map, nrow = 2, ncol = 3)
```

## Sequencing saturation and depth
After assessing mapped reads, it is important to test which is the sequencing saturation and depth for each library. The sequencing saturation is dependent on the library complexity and sequencing depth. The library complexity is the total number of different transcripts present in the library and it varies between the cell types/tissues, whereas the sequencing depth is the number of Set2 reads per cell. For this reason, we will plot the number of detected genes as a function of depth (sequenced reads). As sequencing depth increases, more genes are detected, but this function reaches a plateau, whereby more sequenced reads does not result in more detected genes; therefore, at this point we assure we sequenced until saturation. More specifically, the sequencing saturation the fraction of confidently mapped, valid cell-barcode, valid UMI reads that had a non-unique (cell-barcode, UMI, gene).

```{r fig.height = 11, fig.width = 14}
gg_qc_samplesize_cells <- cellranger_metrics_gex_df %>%
  mutate(Number_of_Reads_mil = sample__Number_of_reads_assigned_to_the_sample / 1000000) %>%
  ggplot(aes(x = sample__Cells,
             y = Number_of_Reads_mil, color = sample_protocol)) +
  geom_point(aes(shape=tissue), size = 3) +
  theme_bw() +
  scale_color_brewer(palette = "Set2") +
  geom_text_repel(aes(label = library_name), size = 3) +
  labs(title = "Sample size",
       x = "Number of cells", 
       y = "Number of reads (in thousands)") +
  theme(axis.title = element_text(size = 12),
        axis.text = element_text(size = 8),
        #legend.position = "none"
        )

gg_qc_seq_sat <- cellranger_metrics_gex_df %>%
  mutate(Sequencing_Saturation_perc = lib__Sequencing_saturation / 100,
         Mean_Reads_per_Cell_tho = lib__Mean_reads_per_cell / 1000) %>%
  ggplot(aes(x = Mean_Reads_per_Cell_tho,
             y = Sequencing_Saturation_perc, color = sample_protocol)) +
  geom_point(aes(shape=tissue), size = 3) +
  theme_bw() +
  scale_color_brewer(palette = "Set2") +
  ylim(0, 1) +
  geom_text_repel(aes(label = library_name), size = 4) +
  labs(title = "Sequencing Saturation",
       x = "Mean Reads per Cell (in thousands)", 
       y = "") +
  theme(axis.title = element_text(size = 12),
        axis.text = element_text(size = 8),
        legend.position = "none"
        )

gg_qc_seq_depth_cell <- cellranger_metrics_gex_df %>%
  mutate(Median_Reads_per_Cell_tho = sample__Median_reads_per_cell / 1000) %>%
  ggplot(aes(x = Median_Reads_per_Cell_tho,
             y = sample__Median_UMI_counts_per_cell, color = sample_protocol)) +
  geom_point(aes(shape=tissue), size = 3) +
  theme_bw() +
  scale_color_brewer(palette = "Set2") +
  geom_text_repel(aes(label = library_name), size = 4) +
  labs(x = "Median Reads per Cell (in thousands)",
       y = "Median UMI counts per Cell") +
  theme(axis.title = element_text(size = 12),
        axis.text = element_text(size = 8),
        legend.position = "none"
        )


gg_qc_seq_depth <- cellranger_metrics_gex_df %>%
  mutate(Number_of_Reads_mil = sample__Number_of_reads_assigned_to_the_sample / 1000000) %>%
  ggplot(aes(x = Number_of_Reads_mil,
             y = sample__Total_genes_detected, color = sample_protocol)) +
  geom_point(aes(shape=tissue), size = 3) +
  theme_bw() +
  scale_color_brewer(palette = "Set2") +
  geom_text_repel(aes(label = library_name), size = 4) +
  labs(x = "Number of Reads (in millions)", 
       y = "Total Genes Detected") +
  theme(axis.title = element_text(size = 12),
        axis.text = element_text(size = 8),
        legend.position = "none"
        )

cowplot::plot_grid(gg_qc_samplesize_cells, gg_qc_seq_sat, 
                   gg_qc_seq_depth_cell, gg_qc_seq_depth, 
                   nrow = 2, ncol = 2)
```


# Library exploration of mapped/unmapped reads
```{r}
# Computing unmapped reads
cellranger_metrics_gex_df$lib_unmapped <- (100 - cellranger_metrics_gex_df$lib__Confidently_mapped_to_genome)
```


```{r fig.height=4, fig.width=7}
# PBMC
cellranger_metrics_gex_df_sub <- cellranger_metrics_gex_df[cellranger_metrics_gex_df$tissue == "PBMC",]
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "CNAG81_fixed"] <- "Fixed"
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "CNAG81_fresh"] <- "Fresh"

plot_library_pct_reads(cellranger_metrics_gex_df_sub)
ggsave("figs/HumanPBMC_LibReads.svg",
       width = 8, height = 5,
       dpi = 300)
```

```{r fig.height=4, fig.width=7}
# Mouse Lung
cellranger_metrics_gex_df_sub <- cellranger_metrics_gex_df[(cellranger_metrics_gex_df$tissue == "Lung" &
                                                            cellranger_metrics_gex_df$organism == "Mouse" &
                                                            (cellranger_metrics_gex_df$sample_protocol == "Fresh" |
                                                            cellranger_metrics_gex_df$sample_protocol == "Fix&Cut_24h")),]
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "Lung_fresh_0"] <- "Fresh_1"
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "Lung_fresh"] <- "Fresh_2"
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "Lung_fixed_24h_0"] <- "Fixed_1"
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "Lung_fixed_24h"] <- "Fixed_2"

plot_library_pct_reads(cellranger_metrics_gex_df_sub)
ggsave("figs/MouseLung_fixed_LibReads.svg",
       width = 8, height = 5,
       dpi = 300)
```

```{r fig.height=4, fig.width=8}
# Mouse Lung with Cryo
cellranger_metrics_gex_df_sub <- cellranger_metrics_gex_df[(cellranger_metrics_gex_df$subproject == "SCGTEST_33" &
                                                            (cellranger_metrics_gex_df$sample_protocol != "Fresh" &
                                                            cellranger_metrics_gex_df$sample_protocol != "Fix&Cut_7d")),]
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "Lung_fixed_24h"] <- "Fixed"
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "Lung_cryo"] <- "Cryopreserved"
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "Lung_fixed_cryo"] <- "Fixed+Cryopreserved"

cellranger_metrics_gex_df_sub$library_name <- factor(
  x = cellranger_metrics_gex_df_sub$library_name,
  levels = c("Fixed", "Cryopreserved", "Fixed+Cryopreserved"))

plot_library_pct_reads(cellranger_metrics_gex_df_sub)
ggsave("figs/MouseLung_cryo_LibReads.svg",
       width = 9, height = 4,
       dpi = 300)
```

```{r fig.height=4, fig.width=7}
# Mouse Lung Stress
cellranger_metrics_gex_df_sub <- cellranger_metrics_gex_df[(cellranger_metrics_gex_df$tissue == "Lung" &
                                                            cellranger_metrics_gex_df$organism == "Mouse" &
                                                            (cellranger_metrics_gex_df$sample_protocol == "Stress_Fresh" |
                                                            cellranger_metrics_gex_df$sample_protocol == "Stress_Fix&Cut")),]
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "Lung_fixed_stressed_1"] <- "Fresh+Stressed"
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "Lung_fresh_stressed_1"] <- "Fixed+Stressed"

plot_library_pct_reads(cellranger_metrics_gex_df_sub)
ggsave("figs/MouseLung_stress_LibReads.svg",
       width = 8, height = 5,
       dpi = 300)
```

```{r fig.height=4, fig.width=7}
# Mouse Colon
cellranger_metrics_gex_df_sub <- cellranger_metrics_gex_df[(cellranger_metrics_gex_df$tissue == "Colon" &
                                                            cellranger_metrics_gex_df$organism == "Mouse" &
                                                            (cellranger_metrics_gex_df$sample_protocol == "Fresh" |
                                                            cellranger_metrics_gex_df$sample_protocol == "Fix&Cut_24h")),]
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "Colon_fresh"] <- "Fresh"
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "Colon_fixed_24h"] <- "Fixed"


plot_library_pct_reads(cellranger_metrics_gex_df_sub)
ggsave("figs/MouseColon_LibReads.svg",
       width = 8, height = 5,
       dpi = 300)
```

```{r fig.height=4, fig.width=8}
# Human Colon
cellranger_metrics_gex_df_sub <- cellranger_metrics_gex_df[(cellranger_metrics_gex_df$tissue == "Colon" &
                                                            cellranger_metrics_gex_df$organism == "Human"),]
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "8277_Fix&Cut"] <- "Cryopreservation+Fixed"
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "8281_Kiel"] <- "Cryopreservation"


plot_library_pct_reads(cellranger_metrics_gex_df_sub)
```


```{r fig.height=4, fig.width=8}
# Human Colon
cellranger_metrics_gex_df_sub <- cellranger_metrics_gex_df[(cellranger_metrics_gex_df$tissue == "Colon" &
                                                            cellranger_metrics_gex_df$organism == "Human" &
                                                             cellranger_metrics_gex_df$subproject == "SALAS_47"),]
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "FIX-1"] <- "Fresh"
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "FIX-2"] <- "Fixed"
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "FIX-3"] <- "Fixed+Cryopreserved"

cellranger_metrics_gex_df_sub$library_name <- factor(
  x = cellranger_metrics_gex_df_sub$library_name,
  levels = c("Fresh", "Fixed", "Fixed+Cryopreserved"))

plot_library_pct_reads(cellranger_metrics_gex_df_sub)
```


```{r fig.height=4, fig.width=8}
# Human Colon
cellranger_metrics_gex_df_sub <- cellranger_metrics_gex_df[(cellranger_metrics_gex_df$tissue == "Colon" &
                                                            cellranger_metrics_gex_df$organism == "Human" &
                                                             cellranger_metrics_gex_df$subproject == "THREETR_27"),]
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "Human colon_fresh"] <- "Fresh"
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "Human colon_fixed"] <- "Fixed"
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "Human colon_cryopreserved"] <- "Cryopreserved"
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "Human colon_fixed_cryopreserved"] <- "Fixed+Cryopreserved"

cellranger_metrics_gex_df_sub$library_name <- factor(
  x = cellranger_metrics_gex_df_sub$library_name,
  levels = c("Fresh", "Fixed", "Cryopreserved", "Fixed+Cryopreserved"))

plot_library_pct_reads(cellranger_metrics_gex_df_sub)
ggsave("figs/HumanColon_LibReads.svg",
       width = 9, height = 4,
       dpi = 300)
```


```{r fig.height=3, fig.width=8}
# Human Lung
cellranger_metrics_gex_df_sub <- cellranger_metrics_gex_df[(cellranger_metrics_gex_df$tissue == "Lung" &
                                                            cellranger_metrics_gex_df$organism == "Human" &
                                                             cellranger_metrics_gex_df$subproject == "THREETR_26"),]
cellranger_metrics_gex_df_sub$library_name[cellranger_metrics_gex_df_sub$library_name == "ORIENT_127_fixed"] <- "Fixed+Cryopreservation"

cellranger_metrics_gex_df_sub$library_name <- factor(
  x = cellranger_metrics_gex_df_sub$library_name,
  levels = "Fixed+Cryopreservation")

plot_library_pct_reads(cellranger_metrics_gex_df_sub)
ggsave("figs/HumanLung_LibReads.svg",
       width = 8, height = 5,
       dpi = 300)
```

# Session Info
```{r}
sessionInfo()
```
