---
title: "scRNAseq QC parameters"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction

## Objective
In this Rmarkdown document, we are going to 

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
library(gt)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_project_metadata <- here::here("01_cellranger_mapping/data/FIXnCUT_metadata.csv")

# Functions
source(here::here("bin/utils.R"))

set.seed(1234) # for reproducibility 

palette_color_factor <- c("Fresh" = "#66C2A5", 
                          "Fixed" ="#FC8D62",
                          "Cryopreserved" ="#8DA0CB",
                          "Fixed+Cryopreserved" = "#E78AC3")
```

## Load metadata
The data used in this Rmarkdown document comes from **FIX&CUT** dataset.
```{r}
# Load all metadata information
metadata <- read.csv(path_project_metadata)

# Merging metadata tables
metadata <- metadata %>% 
  filter(subproject == "THREETR_27"
         & tissue == "Colon")


# Modify metadata
metadata$library_name[metadata$library_name == "Human colon_fresh"] <- "Fresh"
metadata$library_name[metadata$library_name == "Human colon_fixed"] <- "Fixed"
metadata$library_name[metadata$library_name == "Human colon_cryopreserved"] <- "Cryopreserved"
metadata$library_name[metadata$library_name == "Human colon_fixed_cryopreserved"] <- "Fixed+Cryopreserved"

metadata$library <- metadata$library_name
# sample_protocol

print("Libraries/Samples metadata")
DT::datatable(metadata, options = list(scrollX = TRUE))
```


# CELLRANGER OUTPUT -- Raw: Assessing sequencing depth

https://github.com/MarioniLab/DropletUtils/blob/master/vignettes/DropletUtils.Rmd

Downsampling on the reads: Given multiple batches of very different sequencing depths, it can be beneficial to downsample the deepest batches to match the coverage of the shallowest batches. This avoids differences in technical noise that can drive clustering by batch.

The `r Biocpkg("scuttle")` package provides some utilities to downsample count matrices, but technically speaking, downsampling on the reads is more appropriate as it recapitulates the effect of differences in sequencing depth per cell.

This can be achieved by applying the `downsampleReads` function to the molecule information file containing the read counts. Users should use `downsampleMatrix()` instead if they want to guarantee similar total counts after downsampling.


```{r}
seq_list <- c(10000000, 20000000, 30000000, 40000000, 50000000, 60000000, 70000000, 80000000)
#prop_list <- c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1)

seqreads_genes_df <- data.frame(matrix(ncol = 3, nrow = 0))
colnames(seqreads_genes_df) <- c("sequenced_reads", "detected_genes", "library")

for (library in metadata$library_name) {
  print(library)
  gem_id <- unique(metadata$gem_id[metadata$library_name == library])
  subproject <- unique(metadata$subproject[metadata$library_name == library])

  path_general <- here::here(paste0("01_cellranger_mapping/subprojects/", subproject, "/jobs"))
  mol.info.path <- paste(path_general, gem_id, gem_id, "outs/per_sample_outs", gem_id, "count/sample_molecule_info.h5", sep = "/")
  mol.info.file <- DropletUtils::read10xMolInfo(mol.info.path)
  
  all_reads <- sum(mol.info.file$data$reads)
  print(all_reads)
  prop_list <- seq_list / all_reads

  for (proportion in prop_list) {
    #print(proportion)
    
    sequenced_reads <- all_reads*proportion
    #print(sequenced_reads)
    
    with.sampling <- DropletUtils::downsampleReads(mol.info.path, prop=proportion)

    m <- rowSums(with.sampling)
    detected_genes <- length(m[m > 0])
    #print(detected_genes)

    seqreads_genes_df[nrow(seqreads_genes_df)+1, ] <- c(sequenced_reads, detected_genes, library)
  }
}

seqreads_genes_df$sequenced_reads <- as.integer(seqreads_genes_df$sequenced_reads)
seqreads_genes_df$detected_genes <- as.integer(seqreads_genes_df$detected_genes)

head(seqreads_genes_df)
```



```{r}
for (protocol in metadata$library_name) {
  fit1 <- lm(formula =  detected_genes ~ sequenced_reads, data = seqreads_genes_df[seqreads_genes_df$library == protocol, ])
  #summary(fit1)
  slope <- round(coef(fit1)[2], digits = 6)
  print(slope)

  index <- rownames(seqreads_genes_df[seqreads_genes_df$library == protocol, ][which.max(seqreads_genes_df[seqreads_genes_df$library == protocol, ]$sequenced_reads), ])
  seqreads_genes_df[index, "slope_gene"] <- slope
}

head(seqreads_genes_df)
write_csv(x = seqreads_genes_df,
          file = "tables/HumanColon_LibGenesvsReads.csv")
```

## # Genes / Seq reads

```{r fig.width=10, fig.height=6}
seqreads_genes_df <- read.csv("tables/HumanColon_LibGenesvsReads.csv")


seqreads_genes_df$library <- factor(x = seqreads_genes_df$library, levels = c("Fresh", "Fixed", "Cryopreserved","Fixed+Cryopreserved"))

seqreads_genes_df %>%
  ggplot(aes(x = sequenced_reads, y = detected_genes, color = library, fill=library, label=slope_gene)) +
  geom_point() +
  geom_smooth(se = TRUE) +
  scale_y_continuous(labels = scales::comma) + 
  scale_x_continuous(labels = scales::comma) +
  theme_classic() +
  scale_color_manual(values = palette_color_factor) +
  scale_fill_manual(values = palette_color_factor) +
  labs(x = "# Sequencing Reads",
     y = "# Detected Genes",
     color = "Protocol") +
    theme(axis.title = element_text(size = 14),
          axis.text = element_text(size = 10),
          strip.placement = "outside",
          strip.background = element_rect(colour = NA),
          legend.position = c(0.85, 0.2),
          legend.text = element_text(size = 12),
          legend.title=element_blank()) +
    guides(fill = guide_legend(override.aes = list(color = NA), reverse = TRUE), 
                      linetype = guide_legend(override.aes = list(color = "black", fill=NA)),
           color = FALSE
           )


seqreads_genes_df %>%
  ggplot(aes(x = sequenced_reads, y = detected_genes, color = library,  fill=library)) + # label = slope_gene,
  geom_point() +
  scale_y_continuous(labels = scales::comma) + 
  scale_x_log10(labels = scales::comma) +
  stat_smooth(method='lm', se = TRUE) +
  theme_classic() +
  scale_color_manual(values = palette_color_factor) +
  scale_fill_manual(values = palette_color_factor) +
  labs(x = "# Sequencing Reads",
     y = "# Detected Genes",
     color = "Protocol") +
    theme(axis.title = element_text(size = 14),
          axis.text = element_text(size = 10),
          strip.placement = "outside",
          strip.background = element_rect(colour = NA),
          legend.position = c(0.85, 0.2),
          legend.text = element_text(size = 12),
          legend.title=element_blank()) +
    guides(fill = guide_legend(override.aes = list(color = NA), reverse = TRUE), 
                      linetype = guide_legend(override.aes = list(color = "black", fill=NA)),
           color = FALSE
           )

ggsave("figs/HumanColon_LibGenesvsReads.svg",
       width = 8, height = 5,
       dpi = 300)
```


# CELLRANGER OUTPUT -- Unfiltered
```{r}
cell_info_df <- data.frame(matrix(ncol = 7, nrow = 0))
colnames(cell_info_df) <- c("cell", "reads", "umi", "gene", "library", "library", "replicate")


for (library in metadata$library_name) {
  #print(protocol)
  gem_id <- unique(metadata$gem_id[metadata$library_name == library])
  subproject <- unique(metadata$subproject[metadata$library_name == library])
  protocol <- unique(metadata$library[metadata$library_name == library])

  path_general <- here::here(paste0("01_cellranger_mapping/subprojects/", subproject, "/jobs"))
  mol.info.path <- paste(path_general, gem_id, gem_id, "outs/per_sample_outs", gem_id, "count/sample_molecule_info.h5", sep = "/")
  mol.info.file <- DropletUtils::read10xMolInfo(mol.info.path)
  
    # Get the number of reads / cell [Category, x]
  reads_cell_df <- aggregate(mol.info.file$data$reads, 
                             by= list(cell=(mol.info.file$data$cell)),
                             FUN=sum)
  names(reads_cell_df)[names(reads_cell_df) == "x"] <- "reads"
  
  # Get the number of umi / cell [Category, x]
  umi_cell_df <- aggregate(data = mol.info.file$data,
            umi ~ cell, 
            FUN=function(umi) length(unique(umi))
            )
  
  # Get the number of genes / cell [Category, x]
  gene_cell_df <- aggregate(data = mol.info.file$data,
            gene ~ cell, 
            FUN=function(gene) length(unique(gene))
            )
  
  df <- full_join(reads_cell_df, umi_cell_df, by = "cell")
  df <- full_join(df, gene_cell_df, by = "cell")
  df$library <- library
  df$library <- protocol


  # Information
  #sum(df$reads)
  #sum(df$umi)
  #No possible for genes, as they might be repeated across cells!
  
  cell_info_df <- rbind(cell_info_df, df)
}

head(cell_info_df)
```


```{r}
# Computing slope genes
for (library in metadata$library_name) {
  fit1 <- lm(formula =  gene ~ reads, data = cell_info_df[cell_info_df$library == library, ])
  #summary(fit1)
  slope <- round(coef(fit1)[2], digits = 5)
  print(slope)

  index <- rownames(cell_info_df[cell_info_df$library == library, ][which.max(cell_info_df[cell_info_df$library == library, ]$reads), ])
  cell_info_df[index, "slope_gene"] <- slope
}

# Computing slope UMIs
for (library in metadata$library_name) {
  fit1 <- lm(formula =  umi ~ reads, data = cell_info_df[cell_info_df$library == library, ])
  #summary(fit1)
  slope <- round(coef(fit1)[2], digits = 5)
  print(slope)

  index <- rownames(cell_info_df[cell_info_df$library == library, ][which.max(cell_info_df[cell_info_df$library == library, ]$reads), ])
  cell_info_df[index, "slope_umi"] <- slope
}

head(cell_info_df)
write_csv(x = cell_info_df,
          file = "tables/HumanColon_LibGenes-UMIsvsReadsCell.csv")
```


## # Genes / # Seq reads (per cell)
```{r fig.width=12, fig.height=8}
cell_info_df <- read.csv("tables/HumanColon_LibGenes-UMIsvsReadsCell.csv")

cell_info_df$library <- factor(x = cell_info_df$library, levels = c("Fresh", "Fixed", "Cryopreserved", "Fixed+Cryopreserved"))

gg <- cell_info_df %>%
  ggplot(aes(x = reads, y = gene, color = library, fill = library)) +
  geom_point(alpha=0.05) + 
  #geom_rug() +
  scale_color_manual(values = palette_color_factor) +
  scale_fill_manual(values = palette_color_factor) +
  stat_smooth(method='lm', se=TRUE) + # fitted linear regression model
    scale_y_log10(labels = scales::comma) +
  scale_x_log10(labels = scales::comma) +
  labs(x = "# Sequencing Reads",
     y = "# Detected Genes",
     color = "Protocol") +
  theme_classic() +
    theme(axis.title = element_text(size = 16),
          axis.text = element_text(size = 12),
          strip.placement = "outside",
          strip.background = element_rect(colour = NA),
          legend.position = c(0.8, 0.2),
          legend.text = element_text(size = 14),
          legend.title=element_blank()) +
    guides(fill = guide_legend(override.aes = list(color = NA), reverse = TRUE), 
           linetype = guide_legend(override.aes = list(color = "black", fill=NA)),
           color = FALSE
           )
ggExtra::ggMarginal(gg, groupColour = TRUE, groupFill = TRUE)

ggsave("figs/HumanColon_LibGenesvsReadsCell.svg",
       width = 8, height = 5,
       dpi = 300)
```

## # UMIs / # Seq reads (per cell)

```{r fig.width=12, fig.height=8}
cell_info_df <- read.csv("tables/HumanColon_LibGenes-UMIsvsReadsCell.csv")

cell_info_df$library <- factor(x = cell_info_df$library, levels = c("Fresh", "Fixed", "Cryopreserved","Fixed+Cryopreserved"))

gg <- cell_info_df %>%
  ggplot(aes(x = reads, y = umi, color = library, fill = library)) +
  geom_point(alpha=0.05) + 
  #geom_rug() +
  scale_color_manual(values = palette_color_factor) +
  scale_fill_manual(values = palette_color_factor) +
  stat_smooth(method='lm', se=TRUE) + # fitted linear regression model
  scale_x_log10(labels = scales::comma) +
  scale_y_log10(labels = scales::comma) +
  labs(x = "# Sequencing Reads",
     y = "# Detected UMIs",
     color = "Protocol") +
  theme_classic() +
    theme(axis.title = element_text(size = 16),
          axis.text = element_text(size = 12),
          strip.placement = "outside",
          strip.background = element_rect(colour = NA),
          legend.position = c(0.8, 0.2),
          legend.text = element_text(size = 14),
          legend.title=element_blank()) +
    guides(fill = guide_legend(override.aes = list(color = NA), reverse = TRUE), 
                      linetype = guide_legend(override.aes = list(color = "black", fill=NA)),
           color = FALSE
           )
           
ggExtra::ggMarginal(gg, groupColour = TRUE, groupFill = TRUE)

ggsave("figs/HumanColon_LibUMIsvsReadsCell.svg",
       width = 8, height = 5,
       dpi = 300)
```


# CELLRANGER OUTPUT -- Filtered (Cellranger)
```{r}
# Load Seurat object
seurat_obj <- readRDS(here::here("02_QC/results/R_objects/FIXnCUT_HumanColon_merged.rds"))
seurat_obj
```


## Cumulative # Genes / # cells
```{r}
size_list <- seq(1, 100, by=2)
seurat_obj$library_name <- as.factor(seurat_obj$library_name)

detectedgenes_df <- data.frame(matrix(
  ncol =  length(levels(seurat_obj$library_name)),
  nrow = length(size_list)), 
  row.names = as.character(size_list))
names(detectedgenes_df) <- levels(seurat_obj$library_name)


Idents(seurat_obj) <- "library_name"
for (size in size_list) {
  #print(size)
  for (library in levels(seurat_obj$library_name)) {
  #print(library)
    seurat_obj_sub <- seurat_obj[, seurat_obj$library_name == library]
    detectedgenes_list <- list()

    for (var in 1:50) {
      seurat_obj_sub_n <- subset(x = seurat_obj_sub, downsample = size)
      #print(table(seurat_obj_sub$library))
  
      m <- rowSums(seurat_obj_sub_n@assays$RNA@counts != 0)
      detected_genes <- length(m[m > 0])
      #print(detected_genes)
      detectedgenes_list[var] <- detected_genes
    }
  
    mean_detectedgenes <- mean(unlist(detectedgenes_list))
    #print(mean_detectedgenes)
    detectedgenes_df[as.character(size), library] <- mean_detectedgenes  
    
  }
}

detectedgenes_df
detectedgenes_df$num_cells <- as.integer(rownames(detectedgenes_df))

detectedgenes <- detectedgenes_df %>% 
  gather(key = "library_name", value = "detected_genes", `Cryopreserved`:`Fresh`)

head(detectedgenes)
write_csv(x = detectedgenes,
          file = "tables/HumanColon_CumulativeGenes.csv")
```


```{r fig.width=10, fig.height=6}
detectedgenes <- read.csv("tables/HumanColon_CumulativeGenes.csv")

detectedgenes$library_name <- factor(x = detectedgenes$library_name, levels = c("Fresh", "Fixed", "Cryopreserved","Fixed+Cryopreserved"))


detectedgenes %>% 
  ggplot(aes(x = num_cells, y = detected_genes, color = library_name)) +
  geom_point() +
  scale_fill_manual(values = palette_color_factor) +
  scale_color_manual(values = palette_color_factor) +
  labs(x = "Number of sampled cells",
     y = "Detected genes") +
  theme_classic() +
  theme(axis.title = element_text(size = 16),
          axis.text = element_text(size = 12),
          strip.placement = "outside",
          strip.background = element_rect(colour = NA),
          legend.position = c(0.85, 0.2),
          legend.text = element_text(size = 14),
          legend.title=element_blank()) +
    guides(color = guide_legend(override.aes = list(fill = NA), reverse = TRUE))

ggsave("figs/HumanColon_CumulativeGenes.svg",
       width = 8, height = 5,
       dpi = 300)
```


## QC metrics
Here, we show an overview of the scRNA-seq data obtained after the quality control.
```{r}
table_qc_gex(seurat_obj@meta.data, subtitle = "After cell QC filtering out")
```


```{r fig.height=6, fig.width=20}
VlnPlot(seurat_obj,
        features = c("nCount_RNA", "nFeature_RNA", "pct_mt", "pct_rb"),
        group.by = "library_name",
        pt.size = 0.01,
        log = T,
        ncol = 4
        ) &
  scale_fill_manual(values = palette_color_factor) &
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          axis.title = element_blank())
  #theme_classic() &
  #ggpubr::stat_compare_means(ref.group = "Fresh", method = "wilcox.test",
  #                               label = "p.format", label.y.npc = 0.95)

VlnPlot(seurat_obj,
        features = c("nCount_RNA", "nFeature_RNA", "pct_mt", "pct_rb"),
        group.by = "library_name",
        pt.size = 0.01,
        log = F,
        ncol = 4
        ) &
  scale_fill_manual(values = palette_color_factor) &
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          axis.title = element_blank())
  #theme_classic() &
  #ggpubr::stat_compare_means(ref.group = "Fresh", method = "wilcox.test",
  #                               label = "p.format", label.y.npc = 0.95)
```


# Session Info
```{r}
sessionInfo()
```