---
title: "Differential Expression Analysis (DEA)"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction
We aim to find the transcriptional signatures associated with a particular condition.

## Objective
In this Rmarkdown document, we will explore the results of the Differential Expression Analysis (DEA) between the different levels associated to a particular condition.

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(ggrepel)
library(enrichR)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_r_objects <- here::here("04_GEX_analysis/DE_GO_analysis/results/R_objects")
path_r_tables <- here::here("04_GEX_analysis/DE_GO_analysis/results/tables")
path_r_figs <- here::here("04_GEX_analysis/DE_GO_analysis/results/figs")

# Functions
source(here::here("bin/utils.R"))

# Define comparison parameters
comparison_name_list <- c("sample_protocol")
group_1_list <- c("Fix&Cut")
group_2_list <- c("Fresh")
```


# DEA across all cell-types
```{r fig.width=14, fig.height=10}
for (index in 1:length(comparison_name_list)) {
  comparison_name <- comparison_name_list[[index]]
  group_1 <- group_1_list[[index]]
  group_2 <- group_2_list[[index]]
  
  # Define comparison paths
  #path_volcanoplot <- paste0(path_r_figs, "/scbatlle30_DEgenes_", comparison_name, ".png")
  #path_GO_results <- paste0(path_r_tables, "/scbatlle30_GOBP21_", comparison_name, ".xlsx")
  
  # Load data
  path_object <- paste0(path_r_objects, "/scgtest32_m_COLON_healthy_downs_DEgenes_", comparison_name, ".rds")
  DE_genes <- readRDS(path_object)
  
  
  # Data exploration
  print(paste0(comparison_name, ": ", group_1, " vs ", group_2))
  print("Number of significant DE genes (padj < 0.05)")
  print(table(DE_genes$is_significant))
  print("Number of sDEgenes UP-regulated (& Log2FC > |0.5|)")
  print(sum(DE_genes$avg_log2FC[DE_genes$is_significant == TRUE] > 0))
  print(sum(DE_genes$avg_log2FC[DE_genes$is_significant == TRUE] > 0.5))
  print("Number of sDEgenes DOWN-regulated (& Log2FC > |0.5|)")
  print(sum(DE_genes$avg_log2FC[DE_genes$is_significant == TRUE] < 0))
  print(sum(DE_genes$avg_log2FC[DE_genes$is_significant == TRUE] < -0.5))

  # Volcano Plot
  gg_volcanoplot <- gg_volcano_DEgenes(
  DEgenes_df = DE_genes, 
  title = comparison_name, 
  ident_1 = group_1,
  ident_2 = group_2,
  threshold_log2FC = 0.5,
  threshold_adjpval = -log10(0.001)
  )
  print(gg_volcanoplot)
  
  # Save results
  #ggsave(filename = path_volcanoplot, plot = gg_volcanoplot)
}
```


# GO analysis
```{r fig.width=14, fig.height=10}
for (index in 1:length(comparison_name_list)) {
  comparison_name <- comparison_name_list[[index]]
  group_1 <- group_1_list[[index]]
  group_2 <- group_2_list[[index]]
  
  # Define comparison paths
  #path_GO_results <- paste0(path_r_tables, "/scbatlle30_GOBP21_", comparison_name)
  
  # Load data
  path_object <- paste0(path_r_objects, "/scgtest32_m_COLON_healthy_downs_DEgenes_", comparison_name, ".rds")
  DE_genes <- readRDS(path_object)

  print(paste0(comparison_name, ": ", group_1, " vs ", group_2))
  ## Pre-processing
  DE_genes_up <- DE_genes %>% 
    dplyr::filter(is_significant & avg_log2FC > 0.5) %>% 
    dplyr::select(gene, avg_log2FC, p_val_adj)
  DE_genes_down <- DE_genes %>% 
    dplyr::filter(is_significant & avg_log2FC < -0.5) %>% 
    dplyr::select(gene, avg_log2FC, p_val_adj)  

  database <- "GO_Biological_Process_2021" # GO_Biological_Process_2021, KEGG_2021_Human, Reactome_2016
  ## EnrichR
  enrichR_results_up <- run_enrichR_forGO(
    database = database,
    DE_genes = DE_genes_up$gene,
    adj_pval = 0.05,
    number_enriched_genes = 5,
    num_GO_size_min = 10,
    num_GO_size_max = 300,
    OR = 2
  )
  enrichR_results_down <- run_enrichR_forGO(
  database = database,
  DE_genes = DE_genes_down$gene,
  adj_pval = 0.05,
  number_enriched_genes = 5,
  num_GO_size_min = 10,
  num_GO_size_max = 300,
  OR = 2
)
  
  ## Filter reliable results
  enrichR_results_up_filtered <- enrichR_results_up %>% 
    dplyr::filter(is_reliable == TRUE)
  enrichR_results_down_filtered <- enrichR_results_down %>% 
  dplyr::filter(is_reliable == TRUE)
  
  ## Plot
  gg_enrichR_GO_up <- gg_enrichR_GO(paste0(comparison_name, " UP-regulated"), 
               database,
               enrichR_results_up_filtered, 
               n_top = 20)
  print(gg_enrichR_GO_up)
  
  gg_enrichR_GO_down <- gg_enrichR_GO(paste0(comparison_name, " DOWN-regulated"),  
             database,
             enrichR_results_down_filtered, 
             n_top = 20)
  print(gg_enrichR_GO_down)
  
  ## Export list of GO pathways
  #name_group_1 <- paste0(group_1, " UP")
  #name_group_2 <- paste0(group_2, " DOWN")
  #enrichR_results_split <- list(enrichR_results_up_filtered, enrichR_results_down_filtered)
  #names(enrichR_results_split) <- c(name_group_1, name_group_2)
  
  # Save results
  ## Export list of DE genes
  #openxlsx::write.xlsx(enrichR_results_split, file = path_GO_results)  
}
```

# Hallmark enrichment analysis
```{r fig.width=14, fig.height=10}
DE_genes_list <- list(DE_genes_up$gene, DE_genes_down$gene)
names(DE_genes_list) <- c("FixnCut", "Fresh")
myList <- list()

for (index in 1:length(DE_genes_list)) {
  # Define comparison paths
  #path_GO_results <- paste0(path_r_tables, "/scbatlle30_GOBP21_", comparison_name)
  condition <- names(DE_genes_list[index])
  print(condition)
  
  DE_genes <- DE_genes_list[[index]]

  database <- "MSigDB_Hallmark_2020" # GO_Biological_Process_2021, KEGG_2021_Human, Reactome_2016
  ## EnrichR
  enrichR_results <- run_enrichR_forGO(
    database = database,
    DE_genes = DE_genes,
    adj_pval = 0.05,
    number_enriched_genes = 5,
    num_GO_size_min = 10,
    num_GO_size_max = 300,
    OR = 2
  )
  
  ## Filter reliable results
  enrichR_results_filtered <- enrichR_results %>% 
    dplyr::filter(is_reliable == TRUE)
  
  ## Plot
  gg_enrichR_hallmark <- gg_enrichR_GO(condition, 
               database,
               enrichR_results_filtered, 
               n_top = 15)
  print(gg_enrichR_hallmark)
  myList[[length(myList)+1]] <- enrichR_results_filtered
}
names(myList) <- names(DE_genes_list)


# Save results
path_hallmark <- paste0(path_r_tables, "/scgtest32_m_COLON_healthy_downs_hallmark_enrichment.xlsx")
openxlsx::write.xlsx(myList, file = path_hallmark)  
```


# Pre-processing
```{r}
library(Seurat)

path_r_objects_in <- here::here("03_clustering_annotation/results/R_objects")
## Load data
seurat_obj <- readRDS(paste0(path_r_objects_in, "/scgtest32_m_COLON_healthy_downs_cells1_clustering_annotation.rds"))


# Excluding Low quality cells
seurat_obj <- subset(seurat_obj, celltypes_1 != "Low quality cells high MT%" &
                       celltypes_1 != "Low quality cells high RB% 1" &
                       celltypes_1 != "Low quality cells high RB% 2" &
                       celltypes_1 != "Low quality cells Hba+Hbb+" &
                       celltypes_1 != "Immune cells ?")

# Color palette
set.seed(1234) # for reproducibility
color_palette <- Polychrome::createPalette(35, c("#fc6060", "#74f774", "#7c7cfc"))
names(color_palette) <- NULL
#Polychrome::swatch(color_palette)
```


## VlnPlot hallmarks
```{r}
HALLMARK_APOPTOSIS <- msigdbr::msigdbr(species = "mouse", category = "H") %>%
  filter(gs_name == "HALLMARK_APOPTOSIS")
HALLMARK_APOPTOSIS <- HALLMARK_APOPTOSIS$gene_symbol

HALLMARK_HYPOXIA <- msigdbr::msigdbr(species = "mouse", category = "H") %>%
  filter(gs_name == "HALLMARK_HYPOXIA")
HALLMARK_HYPOXIA <- HALLMARK_HYPOXIA$gene_symbol

HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY <- msigdbr::msigdbr(species = "mouse", category = "H") %>%
  filter(gs_name == "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY")
HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY <- HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY$gene_symbol

seurat_obj <- AddModuleScore(seurat_obj,
                               features = list(HALLMARK_APOPTOSIS,
                                               HALLMARK_HYPOXIA,
                                               HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY),
                               name = c("HALLMARK_APOPTOSIS",
                                        "HALLMARK_HYPOXIA",
                                        "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY")
                              )
```


```{r fig.width=21, fig.height=7}
VlnPlot(seurat_obj,
        features = c("HALLMARK_APOPTOSIS1", "HALLMARK_HYPOXIA2", "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY3"),
        group.by = "celltypes_1",
        split.by = "sample_protocol",
        fill.by = "sample_protocol",
        cols = color_palette,
        same.y.lims = TRUE,
        pt.size = 0
        ) &
  #scale_fill_brewer(palette = "Paired", direction = -1) +
      stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median,
                 geom='crossbar', colour = "black",
                 width=0.25, position = position_dodge(width = .75)) #&
    #ggpubr::stat_compare_means(ref.group = "Universal Fibroblasts", method = "wilcox.test",
    #                             label = "p.signif")  + # p.format)
  #ylim(-0.2, 0.7)
```


# Session info
```{r}
sessionInfo()
```
