---
title: "Differential Expression Analysis (DEA)"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction
We aim to find the transcriptional signatures associated with a particular condition.

## Objective
In this Rmarkdown document, we will explore the results of the Differential Expression Analysis (DEA) between the different levels associated to a particular condition.

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(ggrepel)
#library(enrichR)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_r_objects <- here::here("04_GEX_analysis/DE_GO_analysis/results/R_objects")
path_r_tables <- here::here("04_GEX_analysis/DE_GO_analysis/results/tables")
path_r_figs <- here::here("04_GEX_analysis/DE_GO_analysis/results/figs")

# Functions
source(here::here("bin/utils.R"))

# Define comparison parameters
comparison_name_list <- c("Fixed24h_vs_Fresh", "Fixed7d_vs_Fresh", "Cryo_vs_Fresh", "FixedCryo_vs_Fresh")
group_1_list <- c("Fix&Cut_24h", "Fix&Cut_7d", "Cryopreservation", "Fix&Cut+Cryopreservation")
group_2_list <- c("Fresh", "Fresh", "Fresh", "Fresh")
```

```{r, echo=FALSE,include = FALSE}
# You need this code to conduct the magic dependences attaching...
DT::datatable(matrix())
```

# DEA across each cell-types
```{r fig.height=20, fig.width=20, results='asis'}
for (index in 1:length(comparison_name_list)) {
  comparison_name <- comparison_name_list[[index]]
  group_1 <- group_1_list[[index]]
  group_2 <- group_2_list[[index]]
  
  # Define comparison paths
  #path_volcanoplot <- paste0(path_r_figs, "/post12_cd45neg_DEgenes_", comparison_name, ".png")
  #path_GO_results <- paste0(path_r_tables, "/post12_cd45neg_GOBP21_", comparison_name, ".xlsx")
  
  # Load data
  path_object <- paste0(path_r_objects, "/scgtest33_m_LUNG_healthy_DEgenes_cells1_", comparison_name, ".rds")
  DE_genes_bycell <- readRDS(path_object)
  
  
  # Data exploration
  print(paste0("COMPARISON: ", comparison_name, ": ", group_1, " vs ", group_2))
  
  df <- data.frame(matrix(ncol = 4, nrow = length(DE_genes_bycell)))
  colnames(df) <- c('CELLTYPES',
                    'Number of significant DE genes (padj < 0.05)',
                    'Number of sDEgenes UP-regulated (& Log2FC > 0.25)',
                    'Number of sDEgenes DOWN-regulated (& Log2FC < -0.25)')
  for (index in 1:length(DE_genes_bycell)) {
    celltype <- names(DE_genes_bycell[index])
    # Get values
    sDE <- sum(DE_genes_bycell[[celltype]]$is_significant == TRUE)
    sDE_up <- sum(DE_genes_bycell[[celltype]]$avg_log2FC[DE_genes_bycell[[celltype]]$is_significant == TRUE] > 0)
    sDE_up_FC <- sum(DE_genes_bycell[[celltype]]$avg_log2FC[DE_genes_bycell[[celltype]]$is_significant == TRUE] > 0.25)
    sDE_down <- sum(DE_genes_bycell[[celltype]]$avg_log2FC[DE_genes_bycell[[celltype]]$is_significant == TRUE] < 0)
    sDE_down_FC <- sum(DE_genes_bycell[[celltype]]$avg_log2FC[DE_genes_bycell[[celltype]]$is_significant == TRUE] < -0.25)   
    # Fill df
    df[index,]$CELLTYPES <- celltype
    df[index,]$`Number of significant DE genes (padj < 0.05)` <- sDE
    df[index,]$`Number of sDEgenes UP-regulated (& Log2FC > 0.25)` <- paste0(sDE_up, " (", sDE_up_FC, ")") 
    df[index,]$`Number of sDEgenes DOWN-regulated (& Log2FC < -0.25)` <- paste0(sDE_down, " (", sDE_down_FC, ")")
  }
  cat(knitr::knit_print(DT::datatable(df, height = "100%", width = "100%", options = list(scrollX = TRUE))))  
  
  ## Volcano Plot
  gg_volcanoplot_list <- purrr::map2(DE_genes_bycell, names(DE_genes_bycell),  function(DE_genes, celltype) {
    gg <- gg_volcano_DEgenes(
      DEgenes_df = DE_genes, 
      title = celltype, 
      ident_1 = group_1,
      ident_2 = group_2,
      threshold_log2FC = 0.25,
      threshold_adjpval = -log10(0.05)
    ) + theme(legend.position = "none")
    gg
  })
  
  gg_volcanoplot_bycell <- cowplot::plot_grid(plotlist = gg_volcanoplot_list, ncol = 4)
  print(gg_volcanoplot_bycell)

  # Save results
  #ggsave(filename = path_volcanoplot, plot = gg_volcanoplot_bycell)
}
```


## Comparison across preservation protocols
```{r}
abs_log2FC <- 0.25
```

### UP-sDE
```{r fig.height=5, fig.width=7}
celltypes_names <- c("Alveolar macrophages", "B cells", "Endothelial cells", "Fibroblasts", "Monocytes/Macrophages/DCs", "Myofibroblasts",
                     "Neutrophils", "NK cells", "Pheumocytes Type I", "Pneumocytes Type II", "Pulmonary Endothelial cells", "T cells")

for (celltype in celltypes_names) {

  print(celltype)
  DE_genes_list <- list()

  for (index in 1:length(comparison_name_list)) {
    comparison_name <- comparison_name_list[[index]]
    group_1 <- group_1_list[[index]]
    group_2 <- group_2_list[[index]]
    
    # Load data
    path_object <- paste0(path_r_objects, "/scgtest33_m_LUNG_healthy_DEgenes_cells1_", comparison_name, ".rds")
    DE_genes_bycell <- readRDS(path_object)
    
    print(paste0(comparison_name, ": ", group_1, " vs ", group_2))
    ## Pre-processing
    DE_genes <- DE_genes_bycell[[celltype]]
    
    DE_genes_filt <- DE_genes %>% 
      dplyr::filter(is_significant & avg_log2FC > abs_log2FC) %>% 
      dplyr::select(gene)
    print(length(DE_genes_filt$gene))
    
    DE_genes_list[index] <- DE_genes_filt
  }
  
  names(DE_genes_list) <- group_1_list

  sDE_UP_intersect <- Reduce(intersect,
       list(DE_genes_list$`Fix&Cut_24h`,
            DE_genes_list$`Fix&Cut_7d`,
            DE_genes_list$Cryopreservation,DE_genes_list$`Fix&Cut+Cryopreservation`
            )
       )
  print(sDE_UP_intersect)  
  
  # Plot UP-regulated
  p <- UpSetR::upset(
    UpSetR::fromList(list(
      "Fix&Cut_24h" = DE_genes_list$`Fix&Cut_24h`,
      "Fix&Cut_7d" = DE_genes_list$`Fix&Cut_7d`,
      "Cryopreservation" = DE_genes_list$Cryopreservation,
      "Fix&Cut+Cryopreservation" = DE_genes_list$`Fix&Cut+Cryopreservation`
    )),
    nsets = 6,
    order.by = "freq",
    #group.by = "sets",
    mainbar.y.label = "# Genes", sets.x.label = "# Genes / sample",
    point.size = 4, line.size = 1,
    empty.intersections = "on",
    queries = list(
       list(query = UpSetR::intersects, 
            params = list("Fix&Cut_24h", "Fix&Cut_7d", "Cryopreservation", "Fix&Cut+Cryopreservation"), 
            color = "orange", active = T)
    )
  )
  
  print(p)
}
```

### DOWN-sDE
```{r fig.height=5, fig.width=7}
celltypes_names <- c("Alveolar macrophages", "B cells", "Endothelial cells", "Fibroblasts", "Monocytes/Macrophages/DCs", "Myofibroblasts",
                     "Neutrophils", "NK cells", "Pheumocytes Type I",  "Pulmonary Endothelial cells", "T cells") #"Pneumocytes Type II",

for (celltype in celltypes_names) {

  print(celltype)
  DE_genes_list <- list()

  for (index in 1:length(comparison_name_list)) {
    comparison_name <- comparison_name_list[[index]]
    group_1 <- group_1_list[[index]]
    group_2 <- group_2_list[[index]]
    
    # Load data
    path_object <- paste0(path_r_objects, "/scgtest33_m_LUNG_healthy_DEgenes_cells1_", comparison_name, ".rds")
    DE_genes_bycell <- readRDS(path_object)
    
    print(paste0(comparison_name, ": ", group_1, " vs ", group_2))
    ## Pre-processing
    DE_genes <- DE_genes_bycell[[celltype]]
    
    DE_genes_filt <- DE_genes %>% 
      dplyr::filter(is_significant & avg_log2FC < -abs_log2FC) %>% 
      dplyr::select(gene)
    print(length(DE_genes_filt$gene))
    
    DE_genes_list[index] <- DE_genes_filt
  }
  
  names(DE_genes_list) <- group_1_list

  sDE_DOWN_intersect <- Reduce(intersect,
       list(DE_genes_list$`Fix&Cut_24h`,
            DE_genes_list$`Fix&Cut_7d`,
            DE_genes_list$Cryopreservation,DE_genes_list$`Fix&Cut+Cryopreservation`
            )
       )
  print(sDE_DOWN_intersect)
  
  
  # Plot DOWN-regulated
  p <- UpSetR::upset(
    UpSetR::fromList(list(
      "Fix&Cut_24h" = DE_genes_list$`Fix&Cut_24h`,
      "Fix&Cut_7d" = DE_genes_list$`Fix&Cut_7d`,
      "Cryopreservation" = DE_genes_list$Cryopreservation,
      "Fix&Cut+Cryopreservation" = DE_genes_list$`Fix&Cut+Cryopreservation`
    )),
    nsets = 6,
    order.by = "freq",
    #group.by = "sets",
    mainbar.y.label = "# Genes", sets.x.label = "# Genes / sample",
    point.size = 4, line.size = 1,
    empty.intersections = "on",
    queries = list(
       list(query = UpSetR::intersects, 
            params = list("Fix&Cut_24h", "Fix&Cut_7d", "Cryopreservation", "Fix&Cut+Cryopreservation"), 
            color = "orange", active = T)
    )
  )
  
  print(p)
}
```

# Session info
```{r}
sessionInfo()
```
