---
title: "Differential Expression Analysis (DEA)"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction
We aim to find the transcriptional signatures associated with a particular condition.

## Objective
In this Rmarkdown document, we will explore the results of the Differential Expression Analysis (DEA) between the different levels associated to a particular condition.

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(ggrepel)
library(Seurat)
library(fgsea)
library(enrichplot)
library(gridExtra)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_r_objects <- here::here("04_GEX_analysis/4_DE_GO_GSEA/results/R_objects")
path_r_tables <- here::here("04_GEX_analysis/4_DE_GO_GSEA/results/tables")
path_r_figs <- here::here("04_GEX_analysis/4_DE_GO_GSEA/results/figs")

# Functions
source(here::here("bin/utils.R"))

# Define comparison parameters
comparison_name_list <- c("Fixed_vs_Fresh")
group_1_list <- c("Fixed")
group_2_list <- c("Fresh")
```

# DEA
```{r fig.width=10, fig.height=7}
for (index in 1:length(comparison_name_list)) {
  comparison_name <- comparison_name_list[[index]]
  group_1 <- group_1_list[[index]]
  group_2 <- group_2_list[[index]]
  
  # Load data
  path_object <- paste0(path_r_objects, "/FIXnCUT_MouseColon_DEgenes_", comparison_name, ".rds")
  DE_genes <- readRDS(path_object)
  
  # Data exploration
  print(paste0(comparison_name, ": ", group_1, " vs ", group_2))
  print("Number of significant DE genes (padj < 0.05)")
  print(table(DE_genes$is_significant))
  print("Number of sDEgenes UP-regulated (& Log2FC > 1)")
  #print(sum(DE_genes$avg_log2FC[DE_genes$is_significant == TRUE] > 0))
  print(sum(DE_genes$avg_log2FC[DE_genes$is_significant == TRUE] > 1))
  print("Number of sDEgenes DOWN-regulated (& Log2FC < -1)")
  #print(sum(DE_genes$avg_log2FC[DE_genes$is_significant == TRUE] < 0))
  print(sum(DE_genes$avg_log2FC[DE_genes$is_significant == TRUE] < -1))

  # Modifying p_val_adj=0 to avoid Inf values when -Log10()
  DE_genes["p_val_adj"][DE_genes["p_val_adj"] == 0] <- 1e-300

  # Volcano Plot
  gg_volcanoplot <- gg_volcano_DEgenes(
  DEgenes_df = DE_genes, 
  title = comparison_name, 
  ident_1 = group_1,
  ident_2 = group_2,
  threshold_log2FC = 1,
  threshold_adjpval = -log10(0.05)
  )
  print(gg_volcanoplot)
  
  # Save results
  path_volcanoplot <- paste0(path_r_figs, "/FIXnCUT_MouseColon_DEA_", comparison_name, ".svg")
  ggsave(filename = path_volcanoplot, plot = gg_volcanoplot, width = 12, height = 8, dpi = 300)
}
```


# GSEA
Gene set enrichment analysis (GSEA) is a rank-based approach that determines whether predefined groups of genes/proteins/etc. are primarily up or down in one condition relative to another (Vamsi K. Mootha et al., 2003; Subramanian et al., 2005). It is typically performed as a follow-up to differential analysis, and is preferred to ORA.

Fast GSEA (FGSEA) in R, which is based on the gene permutation approach (Korotkevich et al., 2016). To perform FGSEA with the fgsea package, we need a list of gene sets/pathways and the ranking metric vector.

fgsea is an R-package for fast preranked gene set enrichment analysis (GSEA). This package allows to quickly and accurately calculate arbitrarily low GSEA P-values for a collection of gene sets. P-value estimation is based on an adaptive multi-level split Monte-Carlo scheme. See the [preprint](https://www.biorxiv.org/content/10.1101/060012v2) for algorithmic details.


```{r}
# Retrieve GENE SETS pathways - Pathways of interest
# C2-REACTOME --> CP:REACTOME
# C2-WP --> CP:WIKIPATHWAYS
# C5-GOBP --> GO:BP 
# C5-GOMF --> GO:MF
category_db <- "MSigDBsubset"
canonical_pathways_db <- msigdbr::msigdbr(species = "mouse") %>% 
  filter(gs_subcat %in% c("GO:BP", "CP:KEGG", "CP:REACTOME", "CP:WIKIPATHWAYS"))
#, category = "C5", subcategory = "GO:MF")

# Fixing format to work with fgsea
canonical_pathways <- split(x = canonical_pathways_db$gene_symbol, f = canonical_pathways_db$gs_name)
# Explore pathway
# print(summary(sapply(canonical_pathways, length)))
```


```{r fig.width=15, fig.height=10}
for (index_1 in 1:length(comparison_name_list)) {
  comparison_name <- comparison_name_list[[index_1]]
  group_1 <- group_1_list[[index_1]]
  group_2 <- group_2_list[[index_1]]
  
  # Load data
  path_object <- paste0(path_r_objects, "/FIXnCUT_MouseColon_DEgenes_", comparison_name, ".rds")
  DE_genes <- readRDS(path_object)

  # GSEA analysis
  DE_genes_df <- DE_genes %>% 
    dplyr::filter(is_significant) %>% # & avg_log2FC > 0) %>% 
    dplyr::select(gene, avg_log2FC, p_val_adj) %>% 
    dplyr::arrange(desc(avg_log2FC, p_val_adj))

  
  # Ranking DE genes for fgsea
  ranked_genes <- DE_genes_df$avg_log2FC
  names(ranked_genes) <- DE_genes_df$gene
  
  # Running FGSEA
  fgseaRes <- fgsea(pathways = canonical_pathways, 
                    stats    = ranked_genes,
                    scoreType = "std", # "pos",
                    #eps0.0, # set to 0 to estimate Pvalue more accurately
                    minSize  = 10,
                    maxSize  = 300)
  
  # Filtering FGSEA results
  # Formatting results
  fgseaRes <- fgseaRes %>% 
    filter(padj < 0.05) 
  fgseaRes <- fgseaRes %>% 
    mutate(leadingEdge_num = lengths(fgseaRes$leadingEdge),
           genes_size_pct = round(leadingEdge_num/size*100,1)) %>% 
    filter(leadingEdge_num >= 5) %>% 
    filter(genes_size_pct >= 15) %>% 
    arrange(padj)
  
  fgseaRes_UP <- fgseaRes %>% filter(fgseaRes$ES > 0) %>% 
    dplyr::arrange(desc(NES, padj))
  fgseaRes_DOWN <- fgseaRes %>% filter(fgseaRes$ES < 0) %>% 
    dplyr::arrange(NES, padj)
  fgseaRes_split <- list(fgseaRes_UP, fgseaRes_DOWN)
  names(fgseaRes_split) <- c("UP-regulated", "DOWN-regulated")
  
  # Exploring results
  # Plot the top 10 pathways enriched at the top and bottom of the ranked list, respectively.
  topPathwaysUp <- fgseaRes[ES > 0][head(order(padj), n=20), pathway]
  topPathwaysDown <- fgseaRes[ES < 0][head(order(padj), n=20), pathway]
  topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
  
  print(paste0(comparison_name, ": ", group_1, " vs ", group_2))
  gg_GSEAresults <- ggpubr::as_ggplot(plotGseaTable(canonical_pathways[topPathways], ranked_genes, fgseaRes, render = F))
  print(gg_GSEAresults)
  
  # Save results
  path_GSEArds <- paste0(path_r_objects, "/GSEA/FIXnCUT_MouseColon_GSEA_", comparison_name, "_", category_db ,".rds")
  saveRDS(fgseaRes_split, file = path_GSEArds)
  
  path_GSEAtable <- paste0(path_r_figs, "/GSEA/FIXnCUT_MouseColon_GSEAtop_", comparison_name, "_", category_db , ".svg")
  ggsave(plot = gg_GSEAresults, filename = path_GSEAtable, width = 15, height = 10, dpi = 300)
  
  results_path <- paste0(path_r_tables, "/GSEA/FIXnCUT_MouseColon_GSEA_", comparison_name, "_", category_db ,".xlsx")
  openxlsx::write.xlsx(fgseaRes_split, file = results_path)
}
```

## Exploring GSEA
```{r}
# #DE_genesdf <- DE_genes_df %>% arrange(desc(avg_log2FC, p.adjust()))
# DE_genes_df$rank_num <-1:nrow(DE_genes_df)
# 
# #p1 <- plotEnrichment(canonical_pathways[[fgseaRes[1]$pathway]], ranked_genes) +
# p1 <- plotEnrichment(canonical_pathways[["GOBP_RESPONSE_TO_TOXIC_SUBSTANCE"]], ranked_genes) +
#   #labs(title=fgseaRes[1]$pathway, x = "", y = "Enrichment Score") +
#   theme(plot.margin = unit(c(1,1,0,1), "cm"),
#         axis.text.x = element_blank(),
#         axis.ticks.x = element_blank()
#         )
# 
# p2 <- ggplot() +
#   geom_segment(data=DE_genes_df, aes_(x=~rank_num, xend=~rank_num, y=~avg_log2FC, yend=0), color="grey") +
#   ylab("Ranked Metric") +
#   xlab("Rank in Ordered Dataset") +
#   scale_y_continuous(limits = c(-3,3), labels = scales::comma) +
#   theme_minimal() +
#   theme(plot.margin = unit(c(0,1,1,1), "cm"),
#         panel.grid.minor = element_blank())
# 
# cowplot::plot_grid(p1, p2, ncol=1, nrow = 2, rel_heights = c(2, 1))
```

# Session info
```{r}
sessionInfo()
```
