---
title: "scRNAseq QC parameters"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction

## Objective
In this Rmarkdown document, we are going to 

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
library(gt)
library(grid)
library(RColorBrewer)
library(parallelDist)
library(scales)
library(gtable)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_project_metadata <- here::here("01_cellranger_mapping/data/FIXnCUT_metadata.csv")
path_r_figs_out <- here::here("04_GEX_analysis/results/figs")

# Functions
source(here::here("bin/utils.R"))

# Parameters
confounder_variables <- c("library_name", "sample_protocol", "fixation_time")
confounder_names <- c("Library", "Sample Protocol", "Fixation time")

metadata_variables_confounder <- c("nCount_RNA", "nFeature_RNA", "gem_id", 
                                   "pct_mt", "pct_rb", "Phase",
                                   "library_name", "sample_protocol", "fixation_time")

palette_color_factor <- c("Fresh" = "#66C2A5", 
                          "Fixed" ="#FC8D62")

palette_color_cells <- c("Naive CD4+ T cells" = "#184E77",
                         "Transitional Memory CD4+ T cells" = "#1E6091",
                         "Central Memory CD4+ T cells" = "#1A759F",
                         "Effector Memory CD4+ T cells" = "#168AAD",
                         "Tregs" = "#34A0A4",
                         "Naive CD8+ T cells" = "#52B69A",
                         "Cytotoxic CD8+ T cells" = "#76C893",
                         "gdT cells" = "#99D98C",
                         "MAIT" = "#D9ED92",
                         "NK cells" = "#FFB627",

                         "Classical Monocytes" = "#590D22", 
                         "Intermediate Monocytes" = "#A4133C",
                         "Non-classical Monocytes" = "#FF4D6D",
                         "DC1/DC2" = "#FF8FA3",
                         "pDCs" = "#FFCCD5",

                         "Naive B cells" = "#7371FC",
                         "Memory B cells" = "#CDC1FF",
                         
                         "Progenitor cells" = "#595959",
                         "Cycling cells" = "#A5A5A5"
                         )
set.seed(1234) # for reproducibility 
```

## Load metadata
The data used in this Rmarkdown document comes from **SCGTEST_34** dataset.
```{r}
# Load all metadata information
metadata <- read.csv(path_project_metadata)

# Merging metadata tables
metadata <- metadata %>% 
  filter(subproject == "SCGTEST_34" & tissue == "PBMC")

print("Libraries/Samples metadata")
DT::datatable(metadata, options = list(scrollX = TRUE))
```


```{r}
# Load Seurat object
seurat_obj <- readRDS(here::here("03_clustering_annotation/results/R_objects/FIXnCUT_HumanPBMC_clustering_annotation_cleaned.rds"))
seurat_obj

seurat_obj$celltypes <- factor(x = seurat_obj$celltypes,
                                    levels = c("Naive CD4+ T cells", "Transitional Memory CD4+ T cells",
                         "Central Memory CD4+ T cells", "Effector Memory CD4+ T cells", "Tregs", 
                         "Naive CD8+ T cells", "Cytotoxic CD8+ T cells", "gdT cells", "MAIT", "NK cells",
                         "Classical Monocytes", "Intermediate Monocytes", "Non-classical Monocytes",
                         "DC1/DC2","pDCs",
                         "Naive B cells", "Memory B cells",
                         "Progenitor cells","Cycling cells"))
```


```{r}
table_qc_gex(seurat_obj@meta.data, subtitle = "After cell QC filtering out")
```


```{r fig.width=7, fig.height=6}
DimPlot(
  seurat_obj,
  reduction = "pca",
  group.by = "condition",
  shuffle = TRUE,
  pt.size = 0.25) +
  scale_color_manual(values = palette_color_factor) +
  labs(title = "", color = "")

ggsave(paste0(path_r_figs_out, "/HumanPBMC_PCA.svg"),
       width = 7, height = 6,
       dpi = 300)
```

```{r fig.width=7, fig.height=6}
DimPlot(
  seurat_obj,
  reduction = "umap",
  group.by = "condition",
  shuffle = TRUE,
  pt.size = 0.25) +
  scale_color_manual(values = palette_color_factor) +
  labs(title = "", color = "") +
  theme(legend.text = element_text(size = 15))

ggsave(paste0(path_r_figs_out, "/HumanPBMC_UMAP.svg"),
       width = 7, height = 6,
       dpi = 300)
```

```{r fig.width=10, fig.height=5}
DimPlot(
  seurat_obj,
  reduction = "umap",
  group.by = "condition",
  split.by = "condition") +
  scale_color_manual(values = palette_color_factor) &
  labs(title = "", color = "") + NoLegend()

ggsave(paste0(path_r_figs_out, "/HumanPBMC_UMAPsplit.svg"),
       width = 10, height = 5,
       dpi = 300)
```

# QC metrics
```{r fig.height=5, fig.width=15}
VlnPlot(seurat_obj,
        features = c("nCount_RNA", "nFeature_RNA", "pct_mt", "pct_rb"),
        group.by = "condition",
        pt.size = 0,
        log = TRUE,
        ncol = 4
        ) &
  scale_fill_manual(values = palette_color_factor) &
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 20),
          axis.title = element_blank(),
          axis.text.y = element_text(size = 15)
        ) &
  stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median,
                 geom='crossbar', colour = "black",
                 width=0.4, position = position_dodge(width = .75))
  #ggpubr::stat_compare_means(ref.group = "Fresh", method = "wilcox.test",
  #                               label = "p.format", label.y.npc = 0.95)

ggsave(paste0(path_r_figs_out, "/HumanPBMC_QCmetrics.svg"),
       width = 20, height = 6,
       dpi = 300)
```


# QC metrics by cell-type
```{r}
seurat_metadata <-  seurat_obj@meta.data %>% 
  gather(key = "qc_metric", value = "qc_metric_value", c("nCount_RNA", "nFeature_RNA", "pct_mt"))
```


```{r fig.height=10, fig.width=20}
ggplot(seurat_metadata, 
       aes(x = celltypes, y = qc_metric_value, fill = condition)) +
  introdataviz::geom_split_violin() +
  theme_classic() +
#    ggpubr::stat_compare_means(ref.group = "Fresh", method = "wilcox.test",
#                                 label = "p.signif", label.y.npc = 0.95) + # p.format)
  #scale_y_log10() +
  scale_fill_manual(values = palette_color_factor) +
  scale_color_manual(values = palette_color_factor) +
  labs(x = "Cell-types", y="") +
  facet_wrap(~qc_metric, scales = "free_y", nrow = 3) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text = element_text(size=18),
        axis.title = element_blank(),
        strip.text = element_text(size=20),
          legend.position = "right",
          legend.text = element_text(size = 20),
          legend.title=element_blank()) +
    guides(fill = guide_legend(override.aes = list(color = NA), reverse = TRUE), 
           color = FALSE
           )

ggsave(paste0(path_r_figs_out, "/HumanPBMC_QCmetricsCell.svg"),
       width = 20, height = 10,
       dpi = 300)
```


# Cell-types UMAP
```{r fig.width=10, fig.height=7}
# Clustering overview
DimPlot(object = seurat_obj,
        group.by = "celltypes",
        repel = TRUE,
        cols = palette_color_cells,
        shuffle = TRUE,
        pt.size = 0.25
        ) +
  labs(title = "Cell types") +
  guides(color = guide_legend(override.aes = list(size = 5), ncol = 1))

ggsave(paste0(path_r_figs_out, "/HumanPBMC_UMAPcelltypes.svg"),
       width = 10, height = 7,
       dpi = 300)
```

# Celltypes Proportions
```{r fig.width=10, fig.height=5}
dittoSeq::dittoBarPlot(seurat_obj, 
                             "celltypes",
                             group.by = "condition",
                             scale = "percent",
                            retain.factor.levels = TRUE,
                       legend.show  = FALSE,
                             color.panel = palette_color_cells) +
  coord_flip() + 
  labs(title = "") +
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(angle = 0, hjust = 0.5, size=14),
        axis.title.x = element_text(size=20),
        axis.title.y = element_blank())

ggsave(paste0(path_r_figs_out, "/HumanPBMC_CellProp.svg"),
       width = 10, height = 5,
       dpi = 300)
```

# Dotplot gene markers
```{r fig.width=14, fig.height=7}
genes_of_interest <- c("CD3E",	"CD4", # Tcells
                       "CCR7",	"LEF1",	"TCF7", # Naive CD4+ T cells
                       "CRIP1", "ITGB1", "CCR10", # Transitional CD4+ T cells
                       "IL7R",	"PASK", #	Central Memory CD4+ T cells	 
                       "MAF",	"S100A4",	"IL32", # Effector Memory CD4+ T cells
                       "FOXP3",	"DUSP",	"CTLA4", # Tregs
                       "CD8A",	"CD8B", # CD8+
                       "GLYN",	"CCL5",	"NKG7", # Cytotoxic CD8+ T cells
                       "TRDC",	"TRGC1",	"KLRC1", # gd T cells
                       "KLRB1",	"IL7R", "GZMK", # MAIT
                       "FCGR3A",	"GNLY",	"KLRD1", # NK cells
                       "CD14",	"LYZ",	"FCGR3A", # Monocytes
                       "VCAN",	# Classical
                       "HAVCR2",	"HLA-DRA", "HLA-DPA1", # Intermediate
                       "MS4A7",	"LST1", # Non-classical
                       "CLEC9A", "HLA-DPA1", "CD1C",	"CLEC10A", # DC1/DC2
                       "JCHAIN",	"IL3RA",	"IRF8", # pDCs
                       "CD79A",	"CD19",	"MS4A1", # B cells
                       "IGHD",	"IGHM",	"CCR7", # Naive
                        # Memory
                       "CD34", "SOX4", # Progenitors
                       "STMN1",	"TOP2A", "MKI67" # Cycling
                       )
genes_of_interest <- unique(genes_of_interest)

DotPlot(seurat_obj,
        features = genes_of_interest,
        group.by = "celltypes",
        #scale = TRUE,
        #col.max = 3,
        #col.min = 0,
        cols = "RdYlBu",
        dot.scale = 8) +
  theme(axis.text.x = element_text(angle=90, hjust = 1, vjust = 0.5),
        axis.title = element_blank(),
        legend.position = "bottom")

ggsave(paste0(path_r_figs_out, "/HumanPBMC_DotplotMarkers.svg"),
       width = 14, height = 7,
       dpi = 300)
```


```{r fig.width=15, fig.height=6}
avgexpr_mat <- AverageExpression(
  features = genes_of_interest,
  seurat_obj,
  assays = "RNA",
  return.seurat = F,
  group.by = "celltypes",
  slot = "counts")

cell_types <- names(table(seurat_obj$celltypes))
mycolors <- list(cell_type = palette_color_cells)
mat <- as.data.frame(avgexpr_mat$RNA)
colnames(mat) <- cell_types
annotation_col = data.frame(
                    cell_type = cell_types) 
rownames(annotation_col) <- cell_types

input_mat <- t(apply(mat, MARGIN = 1, FUN = function(X) (X - min(X))/diff(range(X))))

#pdf(paste(figures_folder, "M6B_reduced_genes_heatmap.pdf", sep=""),width=5,height=7,paper='special') 
pheatmap2::pheatmap2(t(input_mat),
                annotation_names_col = F,
                annotation_legend = F,
                show_rownames=T, show_colnames = T, 
                border_color = NA,
                cluster_rows = F,
                cluster_cols = F,
                fontsize_row = 12,
                gaps_row = c(10,15,17), 
                gaps_col = c(27,38,46), 
)

ggsave(paste0(path_r_figs_out, "/HumanPBMC_HeatmapMarkers.svg"),
       width = 15, height = 6,
       dpi = 300)
```

# Session Info
```{r}
sessionInfo()
```