---
title: "scRNAseq QC parameters"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction

## Objective
In this Rmarkdown document, we are going to 

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
library(gt)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_project_metadata <- here::here("01_cellranger_mapping/data/FIXnCUT_metadata.csv")
path_r_figs_out <- here::here("04_GEX_analysis/results/figs")

# Functions
source(here::here("bin/utils.R"))

# Parameters
confounder_variables <- c("library_name", "sample_protocol", "fixation_time")
confounder_names <- c("Library", "Sample Protocol", "Fixation time")

metadata_variables_confounder <- c("nCount_RNA", "nFeature_RNA", "gem_id", 
                                   "pct_mt", "pct_rb", "Phase",
                                   "library_name", "sample_protocol", "fixation_time")
palette_color_factor <- c("Fresh" = "#66C2A5", 
                          "Fixed" ="#FC8D62")

palette_color_cells <- c("T-NK cells" = "#E2711D", 
                       "Myeloid cells" = "#590D22", 
                       "Immature B cells" = "#7371FC", 
                       "Mature B cells" = "#c8b6ff",
                       "Distal Colonocytes" = "#22577A", 
                       "Distal Colonocytes Emp1+" = "#38A3A5",
                       "Proximal Colonocytes" = "#57CC99",
                       "Transit-amplifying (TA) cells" = "#C7F9CC",
                       "Goblet cells" = "#590d22", 
                       "Paneth cells" = "#a4133c", 
                       "Tuft cells" = "#ff758f",
                       "Enteroendocrine (EEC) cells" = "#ffccd5" ,  
                       "Endothelium" = "#BB9457", 
                       "Lymphatic Endothelial cells" = "#FFB627",
                       "Myofibroblasts" = "#432818", 
                       "Stromal cells" = "#99582A"
                         )

set.seed(1234) # for reproducibility 
```

## Load metadata
The data used in this Rmarkdown document comes from **SCGTEST_32** dataset.
```{r}
# Load all metadata information
metadata <- read.csv(path_project_metadata)

# Merging metadata tables
metadata <- metadata %>% 
  filter(subproject == "SCGTEST_32" & tissue == "Colon")

print("Libraries/Samples metadata")
DT::datatable(metadata, options = list(scrollX = TRUE))
```


```{r}
# Load Seurat object
seurat_obj <- readRDS(here::here("03_clustering_annotation/results/R_objects/FIXnCUT_MouseColon_filtered_post_clustering_annotation.rds"))
seurat_obj

seurat_obj$celltypes <- factor(x = seurat_obj$celltypes,
                                    levels = c("Immature B cells", "Mature B cells", "T-NK cells", "Myeloid cells", 
                                               "Distal Colonocytes", "Distal Colonocytes Emp1+", "Proximal Colonocytes",
                                               "Transit-amplifying (TA) cells", "Goblet cells", "Paneth cells", "Tuft cells",
                                               "Enteroendocrine (EEC) cells", "Endothelium", "Lymphatic Endothelial cells",
                                               "Myofibroblasts", "Stromal cells"
                                               ))
```


```{r}
table_qc_gex(seurat_obj@meta.data, subtitle = "After cell QC filtering out")
```


```{r fig.width=7, fig.height=6}
DimPlot(
  seurat_obj,
  reduction = "pca",
  group.by = "library_name",
  shuffle = TRUE,
  pt.size = 0.25) +
  scale_color_manual(values = palette_color_factor) +
  labs(title = "", color = "")

ggsave(paste0(path_r_figs_out, "/MouseColon_PCA.svg"),
       width = 7, height = 6,
       dpi = 300)
```

```{r fig.width=7, fig.height=6}
DimPlot(
  seurat_obj,
  reduction = "umap",
  group.by = "library_name",
  shuffle = TRUE,
  pt.size = 0.25) +
  scale_color_manual(values = palette_color_factor) +
  labs(title = "", color = "") +
  theme(legend.text = element_text(size = 15))

ggsave(paste0(path_r_figs_out, "/MouseColon_UMAP.svg"),
       width = 7, height = 6,
       dpi = 300)
```

```{r fig.width=10, fig.height=5}
DimPlot(
  seurat_obj,
  reduction = "umap",
  group.by = "sample_protocol",
  split.by = "library_name") +
  scale_color_manual(values = palette_color_factor) &
  labs(title = "", color = "") + NoLegend()

ggsave(paste0(path_r_figs_out, "/MouseColon_UMAPsplit.svg"),
       width = 15, height = 5,
       dpi = 300)
```


# QC metrics
```{r fig.height=5, fig.width=10}
VlnPlot(seurat_obj,
        features = c("nCount_RNA", "nFeature_RNA", "pct_mt"),
        group.by = "library_name",
        pt.size = 0,
        log = TRUE,
        ncol = 3
        ) &
  scale_fill_manual(values = palette_color_factor) &
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 20),
          axis.title = element_blank(),
          axis.text.y = element_text(size = 15)
        ) &
  stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median,
                 geom='crossbar', colour = "black",
                 width=0.4, position = position_dodge(width = .75))
  #ggpubr::stat_compare_means(ref.group = "Fresh", method = "wilcox.test",
  #                               label = "p.format", label.y.npc = 0.95)

ggsave(paste0(path_r_figs_out, "/MouseColon_QCmetrics.svg"),
       width = 10, height = 6,
       dpi = 300)
```


# QC metrics by cell-type
```{r}
seurat_metadata <-  seurat_obj@meta.data %>% 
  gather(key = "qc_metric", value = "qc_metric_value", c("nCount_RNA", "nFeature_RNA", "pct_mt"))
```


```{r fig.height=7, fig.width=15}
ggplot(seurat_metadata, 
       aes(x = celltypes, y = qc_metric_value, fill = library_name)) +
  introdataviz::geom_split_violin() +
  theme_classic() +
#    ggpubr::stat_compare_means(ref.group = "Fresh", method = "wilcox.test",
#                                 label = "p.signif", label.y.npc = 0.95) + # p.format)
  scale_y_log10() +
  scale_fill_manual(values = palette_color_factor) +
  scale_color_manual(values = palette_color_factor) +
  labs(x = "Cell-types", y="") +
  facet_wrap(~qc_metric, scales = "free_y", nrow = 3) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text = element_text(size=15),
        axis.title = element_blank(),
        strip.text = element_text(size=18),
          legend.position = "none")

ggsave(paste0(path_r_figs_out, "/MouseColon_QCmetricsCell.svg"),
       width = 15, height = 7,
       dpi = 300)
```



# Cell-types UMAP
```{r fig.width=10, fig.height=7}
# Clustering overview
DimPlot(object = seurat_obj,
        group.by = "celltypes",
        repel = TRUE,
        cols = palette_color_cells,
        shuffle = TRUE,
        pt.size = 0.25
        ) +
  labs(title = "Cell types") +
  guides(color = guide_legend(override.aes = list(size = 5), ncol = 1))

ggsave(paste0(path_r_figs_out, "/MouseColon_UMAPcelltypes.svg"),
       width = 10, height = 7,
       dpi = 300)
```



# Celltypes Proportions
```{r fig.width=10, fig.height=4}
dittoSeq::dittoBarPlot(seurat_obj, 
                             "celltypes",
                             group.by = "library_name",
                             scale = "percent",
                            retain.factor.levels = TRUE,
                       legend.show  = FALSE,
                             color.panel = palette_color_cells) +
  coord_flip() + 
  labs(title = "") +
  theme(axis.text.y = element_text(size=20, hjust = 0.5, angle = 90),
        axis.text.x = element_text(angle = 0, hjust = 0.5, size=14),
        axis.title.x = element_text(size=20),
        axis.title.y = element_blank())

ggsave(paste0(path_r_figs_out, "/MouseColon_CellProp_simple.svg"),
       width = 10, height = 4,
       dpi = 300)
```


```{r fig.width=10, fig.height=7}
gg_c <- dittoSeq::dittoBarPlot(seurat_obj, 
                             "celltypes",
                             group.by = "library_name",
                             scale = "count",
                            retain.factor.levels = TRUE,
                       legend.show  = FALSE,
                             color.panel = palette_color_cells) +
  coord_flip() + 
  labs(title = "") +
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(angle = 0, hjust = 0.5, size=14),
        axis.title.x = element_text(size=20),
        axis.title.y = element_blank())

gg_p <- dittoSeq::dittoBarPlot(seurat_obj, 
                             "celltypes",
                             group.by = "library_name",
                             scale = "percent",
                            retain.factor.levels = TRUE,
                       legend.show  = FALSE,
                             color.panel = palette_color_cells) +
  coord_flip() + 
  labs(title = "") +
  theme(axis.text.y = element_text(size=20),
        axis.text.x = element_text(angle = 0, hjust = 0.5, size=14),
        axis.title.x = element_text(size=20),
        axis.title.y = element_blank())

gg <- gg_c / gg_p

print(gg)

ggsave(paste0(path_r_figs_out, "/MouseColon_CellProp.svg"),
       width = 10, height = 5,
       dpi = 300)
```



# Dotplot gene markers
```{r fig.width=5, fig.height=9}
genes_of_interest <- c("Ebf1", "Cd79a", "Ms4a1", "H2-Eb1", # Immature B cells 
                       # Mature B cells",
                       "Cd3e", "Cd3d", "Klrd1", # T-NK cells, 
                       "Cd14", "Apoe", "C1qa", # Myeloid cells
                       "Mep1a", "Emp1", "Guca2a", "Aqp8", # Distal Colonocytes"
                       "Cbr1", # Proximal Colonocytes
                       "Tubb5", "Stmn1", "Cenpf", # Transit-amplifying (TA) cells
                       "Tff3", "Manf", # Goblet cells", 
                       "Mptx1", # Paneth cells", 
                       "Lrmp", "Dclk1", "Cd24a", # "Tuft cells",
                       "Chgb", "Cpe", "Gfra3", # Enteroendocrine (EEC) cells", 
                       "Fabp4", "Pecam1", "Plvap", # Endothelium", 
                       "Lyve1", # Lymphatic Endothelial cells
                       "Postn", "Col6a2", #Myofibroblasts", 
                       "Dcn" #Stromal cells"
                       )
                       
genes_of_interest <- unique(genes_of_interest)

DotPlot(seurat_obj,
        features = genes_of_interest,
        group.by = "celltypes",
        #scale = TRUE,
        #col.max = 3,
        #col.min = 0,
        cols = "RdYlBu",
        dot.scale = 6) +
  theme(axis.text.x = element_text(angle=90, hjust = 1, vjust = 0.5),
        axis.title = element_blank(),
        legend.position = "bottom") +
  coord_flip() +
  guides(color = guide_legend(ncol = 2)) +
  labs(color = "Average Expression", size = "Percent Expressed") +
  guides(
    color = guide_colorbar(ncol = 4, title.position = "top", title.hjust = 0.5, label.hjust = 0.5, order = 1),
    size = guide_legend(ncol = 4, title.position = "top", title.hjust = 0.5, label.hjust = 0.5, order = 2)
    )

ggsave(paste0(path_r_figs_out, "/MouseColon_DotplotMarkers.svg"),
       width = 5, height = 9,
       dpi = 300)
```

# Session Info
```{r}
sessionInfo()
```