---
title: "scRNAseq QC parameters"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction

## Objective
In this Rmarkdown document, we are going to 

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
library(gt)
library(grid)
library(RColorBrewer)
library(parallelDist)
library(scales)
library(gtable)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_project_metadata <- here::here("01_cellranger_mapping/data/FIXnCUT_metadata.csv")
path_r_figs_out <- here::here("04_GEX_analysis/results/figs")

# Functions
source(here::here("bin/utils.R"))

# Parameters
confounder_variables <- c("library_name", "sample_protocol", "fixation_time")
confounder_names <- c("Library", "Sample Protocol", "Fixation time")

metadata_variables_confounder <- c("nCount_RNA", "nFeature_RNA", "gem_id", 
                                   "pct_mt", "pct_rb", "Phase",
                                   "library_name", "sample_protocol", "fixation_time")
palette_color_factor <- c("Fresh" = "#66C2A5", 
                          "Fixed" ="#FC8D62")

palette_color_cells <- c("Arterial cells" = "#184E77",
                         "Bipotent Capillary cells" = "#1E6091",
                         "Capillary aerocyte (aCap) cells" = "#1A759F",
                         "Capillary general (gCap) cells" = "#168AAD",
                         "Vein cells" = "#34A0A4",
                         "Pneumocytes Type I" = "#52B69A",
                         "Pneumocytes Type II" = "#76C893",
                         "Club cells" = "#99D98C",
                         
                         "Fibroblasts" = "#432818", 
                         "Pericytes" = "#99582A", 
                         "Lymphatic Endothelial cells" = "#BB9457",
                         
                         "Monocytes" = "#590D22", 
                         "Macrophages Apoe+" = "#A4133C",
                         "Alveolar Macrophages" = "#FF4D6D",
                         "Monocyte-derived DCs" = "#FFCCD5",
                         "Neutrophils" = "#595959",
                         
                         "T cells" = "#E2711D",
                         "NK cells" = "#FFB627",

                         "B cells" = "#7371FC",
                         
                         "Cycling cells" = "#A5A5A5"
                         )

set.seed(1234) # for reproducibility 
```

## Load metadata
The data used in this Rmarkdown document comes from **SCGTEST_32+33** dataset.
```{r}
# Load all metadata information
metadata <- read.csv(path_project_metadata)

# Merging metadata tables
metadata <- metadata %>% 
  filter((subproject == "SCGTEST_32" | subproject == "SCGTEST_33")
         & tissue == "Lung"
         & ( sample_protocol == "Fresh" | sample_protocol == "Fix&Cut_24h"))

print("Libraries/Samples metadata")
DT::datatable(metadata, options = list(scrollX = TRUE))
```


```{r}
# Load Seurat object
seurat_obj <- readRDS(here::here("03_clustering_annotation/results/R_objects/FIXnCUT_MouseLung_fixed_clustering_annotation_cleaned.rds"))
seurat_obj

seurat_obj$sample_protocol <- factor(x = seurat_obj$sample_protocol,
                                    levels = c("Fresh", "Fixed"))

seurat_obj$celltypes <- factor(x = seurat_obj$celltypes,
                                    levels = c("T cells", "NK cells", "Monocytes", "Macrophages Apoe+", 
                                               "Alveolar Macrophages", "Monocyte-derived DCs", "B cells", 
                                               "Fibroblasts", "Pericytes", "Lymphatic Endothelial cells",
                                               "Arterial cells", "Vein cells",  "Capillary aerocyte (aCap) cells",
                                               "Capillary general (gCap) cells", "Bipotent Capillary cells", "Pneumocytes Type I", "Pneumocytes Type II",
                                               "Club cells", "Neutrophils","Cycling cells"))

# Remove Neutrophils

cells_to_remove <- c("Neutrophils")
seurat_obj_c <- subset(seurat_obj, idents = cells_to_remove, invert = TRUE)
seurat_obj_c$celltypes <- as.factor(as.character(seurat_obj_c$celltypes))

seurat_obj_c$celltypes <- factor(x = seurat_obj_c$celltypes,
                                    levels = c("T cells", "NK cells", "Monocytes", "Macrophages Apoe+", 
                                               "Alveolar Macrophages", "Monocyte-derived DCs", "B cells",
                                                "Fibroblasts", "Pericytes", "Lymphatic Endothelial cells",
                                               "Arterial cells", "Vein cells",  "Capillary aerocyte (aCap) cells",
                                               "Capillary general (gCap) cells", "Bipotent Capillary cells", "Pneumocytes Type I", "Pneumocytes Type II",
                                               "Club cells", "Cycling cells"))
```


```{r}
table_qc_gex(seurat_obj@meta.data, subtitle = "After cell QC filtering out")
```

```{r fig.width=7, fig.height=6}
DimPlot(
  seurat_obj,
  reduction = "pca",
  group.by = "library_name",
  shuffle = TRUE,
  pt.size = 0.25) +
    scale_color_brewer(palette = "Spectral") +
  labs(title = "", color = "")

ggsave(paste0(path_r_figs_out, "/MouseLung_fixed_PCA.svg"),
       width = 7, height = 6,
       dpi = 300)
```


```{r fig.width=7, fig.height=5}
DimPlot(
  seurat_obj,
  reduction = "umap",
  group.by = "library_name",
  shuffle = TRUE) +
    scale_color_brewer(palette = "Spectral") +
  labs(title = "", color = "")

ggsave(paste0(path_r_figs_out, "/MouseLung_fixed_UMAP.svg"),
       width = 7, height = 6,
       dpi = 300)
```

```{r fig.width=12, fig.height=4}
seurat_obj$library_name <- factor(x = seurat_obj$library_name,
                                    levels = c("Fresh_1", "Fresh_2","Fixed_1", "Fixed_2"))

DimPlot(
  seurat_obj,
  reduction = "umap",
  group.by = "sample_protocol",
  split.by = "library_name") +
  scale_color_manual(values = palette_color_factor) &
  labs(title = "", color = "") & NoLegend()

ggsave(paste0(path_r_figs_out, "/MouseLung_fixed_UMAPsplit.svg"),
       width = 15, height = 5,
       dpi = 300)
```


# QC metrics
```{r fig.height=4, fig.width=10}
VlnPlot(seurat_obj,
        features = c("nCount_RNA", "nFeature_RNA", "pct_mt"),
        split.by = "sample_protocol",
        group.by = "library_name",
        pt.size = 0,
        log = TRUE,
        ncol = 3
        ) &
  scale_fill_manual(values = palette_color_factor) &
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          axis.title = element_blank()) &
  stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median,
                 geom='crossbar', colour = "black",
                 width=0.25, position = position_dodge(width = .75))
  #ggpubr::stat_compare_means(ref.group = "Fresh", method = "wilcox.test",
  #                               label = "p.format", label.y.npc = 0.95)

ggsave(paste0(path_r_figs_out, "/MouseLung_fixed_QCmetrics.svg"),
       width = 10, height = 4,
       dpi = 300)
```

```{r fig.height=4, fig.width=10}
seurat_obj_c$library_name <- factor(x = seurat_obj_c$library_name,
                                    levels = c("Fresh_1", "Fresh_2","Fixed_1", "Fixed_2"))

VlnPlot(seurat_obj_c,
        features = c("nCount_RNA", "nFeature_RNA", "pct_mt"),
        split.by = "sample_protocol",
        group.by = "library_name",
        pt.size = 0,
        log = TRUE,
        ncol = 3
        ) &
  scale_fill_manual(values = palette_color_factor) &
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          axis.title = element_blank()) &
  stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median,
                 geom='crossbar', colour = "black",
                 width=0.25, position = position_dodge(width = .75))
  #ggpubr::stat_compare_means(ref.group = "Fresh", method = "wilcox.test",
  #                               label = "p.format", label.y.npc = 0.95)

ggsave(paste0(path_r_figs_out, "/MouseLung_fixed_QCmetrics-NoNeu.svg"),
       width = 10, height = 4,
       dpi = 300)
```

# QC metrics by cell-type
```{r}
seurat_metadata <-  seurat_obj@meta.data %>%
  gather(key = "qc_metric", value = "qc_metric_value", c("nCount_RNA", "nFeature_RNA", "pct_mt"))
```


```{r fig.height=7, fig.width=15}
ggplot(seurat_metadata, 
       aes(x = celltypes, y = qc_metric_value, fill = sample_protocol)) +
  introdataviz::geom_split_violin() +
  theme_classic() +
    #ggpubr::stat_compare_means(ref.group = "Fresh", method = "wilcox.test",
     #                            label = "p.signif", label.y.npc = 0.95) + # p.format)
  scale_y_log10() +
  scale_fill_manual(values = palette_color_factor) +
  labs(x = "Cell-types", y="") +
  facet_wrap(~qc_metric, scales = "free_y", nrow = 3) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text = element_text(size=15),
        axis.title = element_blank(),
        strip.text = element_text(size=15),
          legend.position = "right",
          legend.text = element_text(size = 18),
          legend.title=element_blank())

ggsave(paste0(path_r_figs_out, "/MouseLung_fixed_QCmetricsCell.svg"),
       width = 15, height = 7,
       dpi = 300)
```



# Cell-types UMAP
```{r fig.width=10, fig.height=7}
# Clustering overview
DimPlot(object = seurat_obj,
        group.by = "celltypes",
        repel = TRUE,
        cols = palette_color_cells,
        shuffle = TRUE,
        pt.size = 0.25
        ) +
  labs(title = "Cell types") +
  guides(color = guide_legend(override.aes = list(size = 5), ncol = 1))

ggsave(paste0(path_r_figs_out, "/MouseLung_fixed_UMAPcelltypes.svg"),
       width = 10, height = 7,
       dpi = 300)
```

```{r fig.width=10, fig.height=7}
# Clustering overview
DimPlot(object = seurat_obj_c,
        group.by = "celltypes",
        repel = TRUE,
        cols = palette_color_cells,
        shuffle = TRUE,
        pt.size = 0.25
        ) +
  labs(title = "Cell types") +
  guides(color = guide_legend(override.aes = list(size = 5), ncol = 1))

ggsave(paste0(path_r_figs_out, "/MouseLung_fixed_UMAPcelltypes-NoNeu.svg"),
       width = 10, height = 7,
       dpi = 300)
```


# Celltypes Proportions
```{r fig.width=12, fig.height=6}
dittoSeq::dittoBarPlot(seurat_obj_c, 
                             "celltypes",
                             group.by = "library_name",
                             scale = "percent",
                            retain.factor.levels = TRUE,
                       legend.show  = FALSE,
                             color.panel = palette_color_cells) +
  coord_flip() + 
  labs(title = "") +
  theme(axis.text.y = element_text(size=20, hjust = 0.5, angle = 90),
        axis.text.x = element_text(angle = 0, hjust = 0.5, size=14),
        axis.title.x = element_text(size=20),
        axis.title.y = element_blank())

ggsave(paste0(path_r_figs_out, "/MouseLung_fixed_CellProp-NoNeu.svg"),
       width = 12, height = 6,
       dpi = 300)
```



# Dotplot gene markers
```{r fig.width=12, fig.height=6}
genes_of_interest <- c("Cd3e",	"Cd4", "Cd8b1", # Tcells# CD8+
                       "Gzmb",	"Klrd1", # NK cells
                       "Plac8",	"Ccr2", # Monocytes
                       "Cd68", "Apoe",  # macros
                       "Ear2", "Itgax", "Siglecf", #Alveolar macro
                       "Clec9a", "Clec10a", # DCs
                       "Ms4a1", "Cd79a", "Cd79b", # B cells
                       "Col1a1", "Col13a1", # epithelial_cells
                       "Postn", "Gucy1a1", 
                       "Ccl21a", "Mmrn1", # Lymph
                       "Mgp", "Eln", # Art
                       "Vwf", "Slc6a2", # Vein
                       "Emp2", "Car4",
                       "Kit", # endothelial_cells
                       "Gpihbp1", "Itga6",
                       "Rtkn2", "Ager", #PneumoI
                       "Sftpd", "Sftpa1", # pneumocytesII
                       "Scgb1a1", "Sftpd", # club_goblet_cells
                       "S100a8", "S100a9", # Neutrophils
                       "Stmn1",	"Top2a", "Mki67" # Cycling
                       )
genes_of_interest <- unique(genes_of_interest)

DotPlot(seurat_obj,
        features = genes_of_interest,
        group.by = "celltypes",
        #scale = TRUE,
        #col.max = 3,
        #col.min = 0,
        cols = "RdYlBu",
        dot.scale = 8) +
  theme(axis.text.x = element_text(angle=90, hjust = 1, vjust = 0.5),
        axis.title = element_blank(),
        legend.position = "bottom")

ggsave(paste0(path_r_figs_out, "/MouseLung_fixed_DotplotMarkers.svg"),
       width = 12, height = 6,
       dpi = 300)
```


# Session Info
```{r}
sessionInfo()
```