---
title: "scRNAseq QC parameters"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction

## Objective
In this Rmarkdown document, we are going to 

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
library(gt)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_project_metadata <- here::here("01_cellranger_mapping/data/FIXnCUT_metadata.csv")
path_r_figs_out <- here::here("04_GEX_analysis/results/figs")

# Functions
source(here::here("bin/utils.R"))

# Parameters
confounder_variables <- c("library_name", "sample_protocol", "fixation_time")
confounder_names <- c("Library", "Sample Protocol", "Fixation time")

metadata_variables_confounder <- c("nCount_RNA", "nFeature_RNA", "gem_id", 
                                   "pct_mt", "pct_rb", "Phase",
                                   "library_name", "sample_protocol", "fixation_time")

palette_color_factor <- c("Fresh" = "#66C2A5", 
                          "Fixed" ="#FC8D62",
                          "Cryo" ="#8DA0CB",
                          "Fixed+Cryo" = "#E78AC3")

set.seed(1234) # for reproducibility 
```

## Load metadata
The data used in this Rmarkdown document comes from **THREETR_27** dataset.
```{r}
# Load all metadata information
metadata <- read.csv(path_project_metadata)

# Merging metadata tables
metadata <- metadata %>% 
  filter(subproject == "THREETR_27" & tissue == "Colon")

print("Libraries/Samples metadata")
DT::datatable(metadata, options = list(scrollX = TRUE))
```


```{r}
# Load Seurat object
seurat_obj <- readRDS(here::here("03_clustering_annotation/results/R_objects/FIXnCUT_HumanColon_filtered_clustering_annotation_cleaned.rds"))
seurat_obj

seurat_obj$library_name <- as.character(seurat_obj$library_name)
seurat_obj$library_name[seurat_obj$library_name == "Cryopreserved"] <- "Cryo"
seurat_obj$library_name[seurat_obj$library_name == "Fixed+Cryopreserved"] <- "Fixed+Cryo"

seurat_obj$library_name <- factor(x = seurat_obj$library_name, 
                                  levels = c("Fresh", "Fixed", "Cryo", "Fixed+Cryo"))

seurat_obj$celltypes <- factor(x = seurat_obj$celltypes,
                                    levels = c("CD4+ T cells", "CD8+ T cells", "NK cells", "M0 macrophages", "M2 macrophages", 
                                               "B cells", "B cells cycling", "Plasma cells", "Plasmablasts", "Mast cells",
                                                "Colonocytes 1", "Colonocytes 2", "Colonocytes PLCG2+", "Colonocytes BEST4+OTOP2+",
                                               "Transit-amplifying (TA) cells", "Goblet cells", "Tuft cells", 
                                               "Endothelium", "Myofibroblasts", "Stromal cells", "Glia", "Cycling cells"))
```


```{r}
table_qc_gex(seurat_obj@meta.data, subtitle = "After cell QC filtering out")
```

```{r fig.width=7, fig.height=5}
DimPlot(
  seurat_obj,
  reduction = "pca",
  group.by = "library_name",
  shuffle = TRUE,
  pt.size = 0.25) +
  scale_color_manual(values = palette_color_factor) &
  labs(title = "", color = "")

ggsave(paste0(path_r_figs_out, "/HumanColon_PCA.svg"),
       width = 7, height = 6,
       dpi = 300)
```

```{r fig.width=7, fig.height=5}
DimPlot(
  seurat_obj,
  reduction = "umap",
  group.by = "library_name",
  shuffle = TRUE) +
  scale_color_manual(values = palette_color_factor) &
  labs(title = "", color = "")

ggsave(paste0(path_r_figs_out, "/HumanColon_UMAP.svg"),
       width = 7, height = 6,
       dpi = 300)
```

```{r fig.width=15, fig.height=5}
DimPlot(
  seurat_obj,
  reduction = "umap",
  group.by = "library_name",
  split.by = "library_name") +
  scale_color_manual(values = palette_color_factor) &
  labs(title = "", color = "") + NoLegend()

ggsave(paste0(path_r_figs_out, "/HumanColon_UMAPsplit.svg"),
       width = 15, height = 5,
       dpi = 300)
```

# QC metrics
```{r fig.height=4, fig.width=10}
VlnPlot(seurat_obj,
        features = c("nCount_RNA", "nFeature_RNA", "pct_mt"),
        split.by = "library_name",
        group.by = "library_name",
        pt.size = 0,
        log = TRUE,
        ncol = 3
        ) &
  scale_fill_manual(values = palette_color_factor) &
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          axis.title = element_blank()) &
  stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median,
                 geom='crossbar', colour = "black",
                 width=0.25, position = position_dodge(width = .75))
  #ggpubr::stat_compare_means(ref.group = "Fresh", method = "wilcox.test",
  #                               label = "p.format", label.y.npc = 0.95)

ggsave(paste0(path_r_figs_out, "/HumanColon_QCmetrics.svg"),
       width = 10, height = 4,
       dpi = 300)
```


# QC metrics by cell-type
```{r}
seurat_metadata <-  seurat_obj@meta.data %>% 
  gather(key = "qc_metric", value = "qc_metric_value", c("nCount_RNA", "nFeature_RNA", "pct_mt"))
```


```{r fig.height=8, fig.width=16}
ggplot(seurat_metadata, 
       aes(x = celltypes, y = qc_metric_value, fill = library_name)) +
  geom_boxplot() +
  theme_classic() +
    #ggpubr::stat_compare_means(ref.group = "Fresh", method = "wilcox.test",
     #                            label = "p.signif", label.y.npc = 0.95) + # p.format)
  scale_y_log10() +
  scale_fill_manual(values = palette_color_factor) +
  labs(x = "Cell-types", y="") +
  facet_wrap(~qc_metric, scales = "free_y", nrow = 3) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text = element_text(size=18),
        axis.title = element_blank(),
        strip.text = element_text(size=20),
          legend.position = "right",
          legend.text = element_text(size = 20),
          legend.title=element_blank()) +
    guides(fill = guide_legend(override.aes = list(color = NA), reverse = TRUE), 
           color = FALSE
           )

ggsave(paste0(path_r_figs_out, "/HumanColon_QCmetricsCell.svg"),
       width = 16, height = 8,
       dpi = 300)
```


# Cell-types UMAP
```{r fig.width=10, fig.height=7}
palette_color_cells <- c("CD4+ T cells" = "#E2711D",
                         "CD8+ T cells" = "#e09f3e",
                         "NK cells" = "#FFB627",
                         
                         "M0 macrophages" = "#590D22",
                         "M2 macrophages" = "#A4133C",
                         "B cells" = "#7371FC", 
                         "B cells cycling" = "#ffd6ff",
                         "Plasma cells" = "#c8b6ff",
                         "Plasmablasts" = "#bbd0ff",
                         "Mast cells" = "#FF4D6D",
                         
                         
                         "Colonocytes 1" = "#2f3e46",
                         "Colonocytes 2" = "#184E77",
                         "Colonocytes PLCG2+" = "#1A759F",
                         "Colonocytes BEST4+OTOP2+" = "#34A0A4",
                         "Transit-amplifying (TA) cells"= "#52B69A",
                         "Goblet cells"  = "#76C893",
                         "Tuft cells" = "#99D98C",
                         "Endothelium" = "#BB9457",
                         "Myofibroblasts" = "#432818",
                         "Stromal cells" = "#99582A",
                         "Glia" = "#595959",
                         "Cycling cells" = "#A5A5A5"
                         )

# Clustering overview
DimPlot(object = seurat_obj,
        group.by = "celltypes",
        repel = TRUE,
        cols = palette_color_cells,
        shuffle = TRUE,
        pt.size = 0.25
        ) +
  labs(title = "Cell types") +
  guides(color = guide_legend(override.aes = list(size = 5), ncol = 1))

ggsave(paste0(path_r_figs_out, "/HumanColon_UMAPcelltypes.svg"),
       width = 10, height = 7,
       dpi = 300)
```


# Celltypes Proportions
```{r fig.width=12, fig.height=6}
dittoSeq::dittoBarPlot(seurat_obj, 
                             "celltypes",
                             group.by = "library_name",
                             scale = "percent",
                            retain.factor.levels = TRUE,
                       legend.show  = FALSE,
                             color.panel = palette_color_cells) +
  coord_flip() + 
  labs(title = "") +
  theme(axis.text.y = element_text(size=20, hjust = 0.5, angle = 90),
        axis.text.x = element_text(angle = 0, hjust = 0.5, size=14),
        axis.title.x = element_text(size=20),
        axis.title.y = element_blank())

ggsave(paste0(path_r_figs_out, "/HumanColon_CellProp.svg"),
       width = 12, height = 6,
       dpi = 300)
```

# Gene markers

```{r fig.width=12, fig.height=7}
genes_of_interest <- c("CD3D", "CD4", # cd4+ T cells
                       "CD8A", # CD8+ T cells
                       "KLRD1", # NK cells
                       "CD68", "C1QA", "TYROBP", "FTL", # M0 macrophages
                       "STAB1", "FOLR2", # M2 macrophages
                       "CD79A", "MS4A1", "HLA-DRA", # B cells
                       "IGKC", "IGHA1", "JCHAIN",# Plasma cells
                       "PRDM1", "CD27", # Plasmablasts
                       "TPSB2", "CPA3", # Mast cells
                       "FABP1", "CA1", # Colonocytes 1
                       "AQP8", "GUCA2A", # Colonocytes 2
                       "PLCG2", "MUC4", "ELF3", # Colonocytes PLCG2+
                       "BEST4", "OPOD2", # Intestinal BEST4+OTOP2+ cells
                       "CENPF", "NUSAP1", # Transit-amplifying (TA) cells cycling
                       "ZG16", "TFF3", "MUC2", # "Mature Goblet"
                       "SH2D6", "TRPM5", # Tuft cells
                       "SPARCL1", "PLVAP", # Endothelium
                       "POSTN", "COL6A2", # Myofibroblasts
                       "ADAMDEC", "CCL11", # Stromal cells
                       "NRXN1", "CRYAB", # Glia
                       "STMN1",	"TOP2A", "MKI67" # Cycling
                       )


genes_of_interest <- unique(genes_of_interest)

DotPlot(seurat_obj,
        features = genes_of_interest,
        group.by = "celltypes",
        #scale = TRUE,
        #col.max = 3,
        #col.min = 0,
        cols = "RdYlBu",
        dot.scale = 8) +
  theme(axis.text.x = element_text(angle=90, hjust = 1, vjust = 0.5),
        axis.title = element_blank(),
        legend.position = "bottom")

ggsave(paste0(path_r_figs_out, "/HumanColon_DotplotMarkers.svg"),
       width = 12, height = 7,
       dpi = 300)
```

# Session Info
```{r}
sessionInfo()
```