---
title: "scRNAseq finding stress signature"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction
Following with the results obtained in the previous notebook, we will now perform other pre-processing steps, such as data normalization, feature selection, scaling, dimensionality reduction, and data visualization prior to batch-effect correction (data integration). To do so, we will follow the current best practices for scRNA-seq QC described in [Luecken al. Mol Syst Biol (2018)](https://doi.org/10.15252/msb.20188746) and adapt some workflows from [Satija Lab](https://satijalab.org/seurat/vignettes.html).

## Objective
In this Rmarkdown document, we are going to perform the previous pre-processing steps.

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(svglite)
library(Seurat)
library(UCell)
library(Nebulosa)
library(viridis)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_project_metadata <- here::here("01_cellranger_mapping/data/FIXnCUT_metadata.csv")
path_r_objects <- here::here("03_clustering_annotation/results/R_objects")
path_r_figs_out <- here::here("04_GEX_analysis/results/figs")

# Functions
source(here::here("bin/utils.R"))

# Color palette
palette_color_factor <- c("Fresh" = "#66C2A5", 
                          "Fixed" ="#FC8D62",
                          "Cryo" ="#8DA0CB",
                          "Fixed+Cryo" = "#E78AC3")

# Comparisons between means
my_comparisons <- list( c("Fresh", "Fixed"), c("Fresh", "Cryo"),  
                        c("Cryo", "Fixed"), c("Cryo", "Fixed+Cryo"), c("Fixed", "Fixed+Cryo"))
```

## Load data
```{r}
# Load Seurat object
seurat_obj <- readRDS(str_c(path_r_objects, "/FIXnCUT_HumanColon_filtered_clustering_annotation_cleaned.rds"))

seurat_obj$library_name <- as.character(seurat_obj$library_name)
seurat_obj$library_name[seurat_obj$library_name == "Cryopreserved"] <- "Cryo"
seurat_obj$library_name[seurat_obj$library_name == "Fixed+Cryopreserved"] <- "Fixed+Cryo"


seurat_obj$library_name <- as.factor(seurat_obj$library_name)
seurat_obj$library_name <- factor(x = seurat_obj$library_name, 
                                  levels = c("Fresh", "Fixed", "Cryo", "Fixed+Cryo"))

seurat_obj$celltypes <- factor(x = seurat_obj$celltypes,
                                    levels = c("CD4+ T cells", "CD8+ T cells", "NK cells", "M0 macrophages", "M2 macrophages", 
                                               "B cells", "B cells cycling", "Plasma cells", "Plasmablasts", "Mast cells",
                                                "Colonocytes 1", "Colonocytes 2", "Colonocytes PLCG2+", "Colonocytes BEST4+OTOP2+",
                                               "Transit-amplifying (TA) cells", "Goblet cells", "Tuft cells", 
                                               "Endothelium", "Myofibroblasts", "Stromal cells", "Glia", "Cycling cells"))
seurat_obj
```

# Housekeeping markers

Housekeeping genes used from: 
https://www.cell.com/trends/genetics/text/S0168-9525(13)00089-9?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS0168952513000899%3Fshowall%3Dtrue

```{r fig.width=20, fig.height=8}
housekeeping_genes <- c("ACTB", "C1orf43", "CHMP2A", "EMC7", "GPI", "PSMB2", "PSMB4", "RAB7A", "REEP5", "SNRPD3", "VCP", "VPS29")

VlnPlot(
  seurat_obj,
  features = housekeeping_genes,
  group.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0, 
  ncol = 6) &
  stat_summary(fun = median, fun.min = median, fun.max = median,
               geom='crossbar', colour = "black", width=0.25, position = position_dodge(width = .25)) &
  ggpubr::stat_compare_means(method = "wilcox.test", comparisons = my_comparisons, 
                             aes(label = "..p.adj.."), label.y.npc = 0.95, label.x.npc = 0.5, size=4)&
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5),
        axis.text = element_text(size=18),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20)) &
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1)))
```


```{r fig.width=20, fig.height=20}
VlnPlot(
  seurat_obj,
  features = housekeeping_genes,
  group.by = "celltypes",
  split.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0, 
  ncol = 4) &
  theme(axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1),
        axis.text = element_text(size=12),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20)) 
```


# Low quality markers

```{r fig.width=15, fig.height=4}
lowquality_genes <- c("HBB", "HBA1", "HBA2", "HBD")

VlnPlot(
  seurat_obj,
  features = lowquality_genes,
  group.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0, 
  ncol = 4) &
  stat_summary(fun = median, fun.min = median, fun.max = median,
               geom='crossbar', colour = "black", width=0.25, position = position_dodge(width = .25)) &
  ggpubr::stat_compare_means(method = "wilcox.test", comparisons = my_comparisons, 
                             aes(label = "..p.adj.."), label.y.npc = 0.95, label.x.npc = 0.5, size=4)&
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5),
        axis.text = element_text(size=18),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20)) &
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1)))
```

```{r fig.width=20, fig.height=5}
VlnPlot(
  seurat_obj,
  features = lowquality_genes,
  group.by = "celltypes",
  split.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0.01, 
  ncol = 4) &
  theme(axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1),
        axis.text = element_text(size=12),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20)) 
```


# Heat shock protein (HSP) markers
```{r fig.width=20, fig.height=8}
HSP_genes <- c("HSP90AA1", "HSP90AB1", "HSPA1A", "HSPA1B", "HSPA5", "HSPA6", "HSPA8", "HSPB1", "HSPD1", "HSPE1", "HSPH1")

VlnPlot(
  seurat_obj,
  features = HSP_genes,
  group.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0, 
  ncol = 6) &
  stat_summary(fun = median, fun.min = median, fun.max = median,
               geom='crossbar', colour = "black", width=0.25, position = position_dodge(width = .25)) &
  ggpubr::stat_compare_means(method = "wilcox.test", comparisons = my_comparisons, 
                             aes(label = "..p.adj.."), label.y.npc = 0.95, label.x.npc = 0.5, size=4)&
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5),
        axis.text = element_text(size=18),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20)) &
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) 
```


```{r fig.width=20, fig.height=15}
VlnPlot(
  seurat_obj,
  features = HSP_genes,
  group.by = "celltypes",
  split.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0, 
  ncol = 4) &
  theme(axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1),
        axis.text = element_text(size=12),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20)) 
```

# Stress-sensitive genes
```{r fig.width=20, fig.height=8}
hypoxiaadaptation_genes <- c("DDIT3", "XBP1", "DDIT4", "P4HB", # ER stress (Rashid et al, 2015)
                             "ENO", "HK2", "PGK1", "GAPDH", # glycolysis
                             "PDK1", "PHD", "SLC2A1" # hypoxia; https://pubmed.ncbi.nlm.nih.gov/32144406/
                             )

VlnPlot(
  seurat_obj,
  features = hypoxiaadaptation_genes,
  group.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0, 
  ncol = 5) &
  stat_summary(fun = median, fun.min = median, fun.max = median,
               geom='crossbar', colour = "black", width=0.25, position = position_dodge(width = .25)) &
  ggpubr::stat_compare_means(method = "wilcox.test", comparisons = my_comparisons, 
                             aes(label = "..p.adj.."), label.y.npc = 0.95, label.x.npc = 0.5, size=4)&
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5),
        axis.text = element_text(size=18),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20)) &
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1)))
```

```{r fig.width=20, fig.height=15}
VlnPlot(
  seurat_obj,
  features = hypoxiaadaptation_genes,
  group.by = "celltypes",
  split.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0, 
  ncol = 3) &
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text = element_text(size=12),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20)) 
```


# SIGNATURE: dissociation-associated on mouse Muscle Stem Cells (MSC) by van den Brink et al.
```{r fig.width=20, fig.height=8}
path_signature_vandenBrinketal <- here::here("04_GEX_analysis/5_gene_signatures/published_signatures/signature_vandenBrinketal_dissociation-associated_mMuscleSC.csv")
vandenBrinketal__genes <- read_csv(path_signature_vandenBrinketal)

vandenBrinketal_up_genes <- vandenBrinketal__genes %>% 
  filter(direction == "up-regulated") %>% 
  dplyr::select(gene)

vandenBrinketal_up_genes_list <- vandenBrinketal_up_genes$gene
vandenBrinketal_up_genes_list <- str_to_upper(vandenBrinketal_up_genes_list)
print(length(vandenBrinketal_up_genes_list))
print(vandenBrinketal_up_genes_list)
```


### Visualizing genes

```{r fig.width=14, fig.height=8}
DotPlot(seurat_obj,
        features = vandenBrinketal_up_genes_list,
        group.by = "celltypes",
        split.by = "library_name",
        cols = "RdBu",
        dot.scale = 5
        ) +
   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### Computing signature
```{r fig.width=8, fig.height=5}
# UCELL GENE SIGNATURE SCORE
## Using positive and negative gene sets, GENE+ or GENE-
seurat_obj <- AddModuleScore_UCell(seurat_obj, 
                                   features=list("vandenBrinketal_up_genes" = vandenBrinketal_up_genes_list)
                                   )
```

### Visualizing Ucell
```{r fig.width=8, fig.height=5}
VlnPlot(
  seurat_obj,
  features = "vandenBrinketal_up_genes_UCell",
  group.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0) &
  stat_summary(fun = median, fun.min = median, fun.max = median,
               geom='crossbar', colour = "black", width=0.25, position = position_dodge(width = .25)) &
  ggpubr::stat_compare_means(method = "wilcox.test", comparisons = my_comparisons, 
                             aes(label = "..p.adj.."), label.y.npc = 0.95, label.x.npc = 0.5, size=4)&
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5),
        axis.text = element_text(size=18),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20),
        legend.position = "none") &
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1)))
```

```{r fig.width=12, fig.height=8}
VlnPlot(
  seurat_obj,
  features = "vandenBrinketal_up_genes_UCell",
  group.by = "celltypes",
  split.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0) &
  stat_summary(fun = median, fun.min = median, fun.max = median,
               geom='crossbar', colour = "black", width=0.25, position = position_dodge(width = .25)) &
  ggpubr::stat_compare_means(method = "wilcox.test", comparisons = my_comparisons, 
                             aes(label = "..p.adj.."), label.y.npc = 0.95, label.x.npc = 0.5, size=4)&
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text = element_text(size=18),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20)) &
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1)))


ggsave(paste0(path_r_figs_out, "/HumanColon_vandenBrinketalUPgenes_Ucell_Vlnbycells.svg"),
       width = 12, height = 8,
       dpi = 300)
```

```{r fig.width=10, fig.height=5}
plot_density(seurat_obj,
            features = 'vandenBrinketal_up_genes_UCell') + 
    facet_grid(.~seurat_obj$library_name)


ggsave(paste0(path_r_figs_out, "/HumanColon_vandenBrinketalUPgenes_Ucelldensity.svg"),
       width = 10, height = 5,
       dpi = 300)
```

<!-- ## Comparing signature with Fresh sDEgenes -->
<!-- ```{r} -->
<!-- path_r_objects <- here::here("04_GEX_analysis/4_DE_GO_GSEA/results/R_objects") -->
<!-- path_object_DEgenes <- paste0(path_r_objects, "/FIXnCUT_HumanColonpreserved_DEgenes_FixedCryo_vs_Cryo.rds") -->
<!-- DE_genes <- readRDS(path_object_DEgenes) -->

<!-- DE_genes_DOWN_df <- DE_genes %>%  -->
<!--     dplyr::filter(is_significant & avg_log2FC < 0) %>%  -->
<!--     dplyr::select(gene, avg_log2FC, p_val_adj) -->
<!-- DE_genes_DOWN_list <- DE_genes_DOWN_df$gene -->
<!-- ``` -->

<!-- ```{r fig.height=4, fig.width=4} -->
<!-- venndiagram <- VennDiagram::venn.diagram( -->
<!--   x = list(DE_genes_DOWN_list, vandenBrinketal_up_genes_list), -->
<!--   category.names = c(" ", " "), -->
<!--   filename = NULL, -->
<!--   fill = c("#023047", "#ffb703"), -->
<!--   lty = 'blank', -->

<!--   cex = 1.5, -->
<!--   fontfamily = "sans", -->

<!--   cat.cex = 1,   -->
<!--   cat.fontface = "bold", -->
<!--   cat.default.pos = "text", -->
<!--   cat.fontfamily = "sans" -->
<!-- ) -->

<!-- grid::grid.newpage() -->
<!-- grid::grid.draw(venndiagram) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- signature_genes_intersection <- intersect(vandenBrinketal_up_genes_list, DE_genes_DOWN_list) -->
<!-- print(signature_genes_intersection) -->
<!-- ``` -->


<!-- ```{r fig.width=12, fig.height=9} -->
<!-- DotPlot(seurat_obj, -->
<!--         features = signature_genes_intersection, -->
<!--         group.by = "celltypes", -->
<!--         split.by = "library_name", -->
<!--         cols = "RdBu", -->
<!--         dot.scale = 5 -->
<!--         ) + -->
<!--   coord_flip() + -->
<!--   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) + -->
<!--   theme(axis.title = element_blank(), -->
<!--         legend.position = "bottom") -->

<!-- ggsave(paste0(path_r_figs_out, "/HumanColon_vandenBrinketalUPgenes_DEgenes.svg"), -->
<!--        width = 12, height = 9, -->
<!--        dpi = 300) -->
<!-- ``` -->


# SIGNATURE: Warm collagenase-associated on  mouse primary tumors and patient-derived mouse xenografts by O'Flanagan et al.
```{r fig.width=20, fig.height=8}
path_signature_OFlanaganetal <- here::here("04_GEX_analysis/5_gene_signatures/published_signatures/signature_OFlanaganetal_collagenase-associated_xCancer_WARMfig3.csv")
OFlanaganetal__genes <- read_csv(path_signature_OFlanaganetal)

OFlanaganetal_up_genes <- OFlanaganetal__genes %>% 
  filter(direction == "up-regulated") %>% 
  dplyr::select(gene)

OFlanaganetal_up_genes_list <- OFlanaganetal_up_genes$gene
print(length(OFlanaganetal_up_genes_list))
print(OFlanaganetal_up_genes_list)
```


### Visualizing genes

```{r fig.width=14, fig.height=8}
DotPlot(seurat_obj,
        features = OFlanaganetal_up_genes_list,
        group.by = "celltypes",
        split.by = "library_name",
        cols = "RdBu",
        dot.scale = 5
        ) +
   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### Computing signature
```{r fig.width=8, fig.height=5}
# UCELL GENE SIGNATURE SCORE
## Using positive and negative gene sets, GENE+ or GENE-
seurat_obj <- AddModuleScore_UCell(seurat_obj, 
                                   features=list("OFlanaganetal_up_genes" = OFlanaganetal_up_genes_list)
                                   )
```

### Visualizing Ucell
```{r fig.width=8, fig.height=5}
VlnPlot(
  seurat_obj,
  features = "OFlanaganetal_up_genes_UCell",
  group.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0) &
  stat_summary(fun = median, fun.min = median, fun.max = median,
               geom='crossbar', colour = "black", width=0.25, position = position_dodge(width = .25)) &
  ggpubr::stat_compare_means(method = "wilcox.test", comparisons = my_comparisons, 
                             aes(label = "..p.adj.."), label.y.npc = 0.95, label.x.npc = 0.5, size=4)&
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5),
        axis.text = element_text(size=18),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20),
        legend.position = "none") &
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1)))
```

```{r fig.width=12, fig.height=8}
VlnPlot(
  seurat_obj,
  features = "OFlanaganetal_up_genes_UCell",
  group.by = "celltypes",
  split.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0) &
  stat_summary(fun = median, fun.min = median, fun.max = median,
               geom='crossbar', colour = "black", width=0.25, position = position_dodge(width = .25)) &
  ggpubr::stat_compare_means(method = "wilcox.test", comparisons = my_comparisons, 
                             aes(label = "..p.adj.."), label.y.npc = 0.95, label.x.npc = 0.5, size=4)&
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text = element_text(size=18),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20)) &
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1)))


ggsave(paste0(path_r_figs_out, "/HumanColon_OFlanaganetalUPgenes_Ucell_Vlnbycells.svg"),
       width = 12, height = 8,
       dpi = 300)
```

```{r fig.width=10, fig.height=5}
plot_density(seurat_obj,
            features = 'OFlanaganetal_up_genes_UCell') + 
    facet_grid(.~seurat_obj$library_name)


ggsave(paste0(path_r_figs_out, "/HumanColon_OFlanaganetalUPgenes_Ucelldensity.svg"),
       width = 10, height = 5,
       dpi = 300)
```


<!-- ## Comparing signature with Fresh sDEgenes -->
<!-- ```{r} -->
<!-- path_r_objects <- here::here("04_GEX_analysis/4_DE_GO_GSEA/results/R_objects") -->
<!-- path_object_DEgenes <- paste0(path_r_objects, "/FIXnCUT_HumanColonpreserved_DEgenes_FixedCryo_vs_Cryo.rds") -->
<!-- DE_genes <- readRDS(path_object_DEgenes) -->

<!-- DE_genes_DOWN_df <- DE_genes %>%  -->
<!--     dplyr::filter(is_significant & avg_log2FC < 0) %>%  -->
<!--     dplyr::select(gene, avg_log2FC, p_val_adj) -->
<!-- DE_genes_DOWN_list <- DE_genes_DOWN_df$gene -->
<!-- ``` -->

<!-- ```{r fig.height=4, fig.width=4} -->
<!-- venndiagram <- VennDiagram::venn.diagram( -->
<!--   x = list(DE_genes_DOWN_list, OFlanaganetal_up_genes_list), -->
<!--   category.names = c(" ", " "), -->
<!--   filename = NULL, -->
<!--   fill = c("#023047", "#ffb703"), -->
<!--   lty = 'blank', -->

<!--   cex = 1.5, -->
<!--   fontfamily = "sans", -->

<!--   cat.cex = 1,   -->
<!--   cat.fontface = "bold", -->
<!--   cat.default.pos = "text", -->
<!--   cat.fontfamily = "sans" -->
<!-- ) -->

<!-- grid::grid.newpage() -->
<!-- grid::grid.draw(venndiagram) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- signature_genes_intersection <- intersect(OFlanaganetal_up_genes_list, DE_genes_DOWN_list) -->
<!-- print(signature_genes_intersection) -->
<!-- ``` -->


<!-- ```{r fig.width=12, fig.height=9} -->
<!-- DotPlot(seurat_obj, -->
<!--         features = signature_genes_intersection, -->
<!--         group.by = "celltypes", -->
<!--         split.by = "library_name", -->
<!--         cols = "RdBu", -->
<!--         dot.scale = 5 -->
<!--         ) + -->
<!--   coord_flip() + -->
<!--   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) + -->
<!--   theme(axis.title = element_blank(), -->
<!--         legend.position = "bottom") -->

<!-- ggsave(paste0(path_r_figs_out, "/HumanColon_OFlanaganetalUPgenes_DEgenes.svg"), -->
<!--        width = 12, height = 9, -->
<!--        dpi = 300) -->
<!-- ``` -->


# SIGNATURE: Warm-dissociation on mouse kidney samples by Denisenko et al.
```{r fig.width=20, fig.height=8}
path_signature_Denisenkoatal <- here::here("04_GEX_analysis/5_gene_signatures/published_signatures/signature_Denisenkoetal_warm-dissociation_mKidney_TableS1.csv")
Denisenkoatal__genes <- read_csv(path_signature_Denisenkoatal)

Denisenkoatal_up_genes <- Denisenkoatal__genes %>% 
  filter(direction == "up-regulation") %>% 
  dplyr::select(gene)

Denisenkoatal_up_genes_list <- Denisenkoatal_up_genes$gene
Denisenkoatal_up_genes_list <- str_to_upper(Denisenkoatal_up_genes_list)
print(length(Denisenkoatal_up_genes_list))
print(Denisenkoatal_up_genes_list)
```


### Visualizing genes

```{r fig.width=14, fig.height=8}
DotPlot(seurat_obj,
        features = Denisenkoatal_up_genes_list,
        group.by = "celltypes",
        split.by = "library_name",
        cols = "RdBu",
        dot.scale = 5
        ) +
   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### Computing signature
```{r fig.width=8, fig.height=5}
# UCELL GENE SIGNATURE SCORE
## Using positive and negative gene sets, GENE+ or GENE-
seurat_obj <- AddModuleScore_UCell(seurat_obj, 
                                   features=list("Denisenkoatal_up_genes" = Denisenkoatal_up_genes_list)
                                   )
```

### Visualizing Ucell
```{r fig.width=8, fig.height=5}
VlnPlot(
  seurat_obj,
  features = "Denisenkoatal_up_genes_UCell",
  group.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0) &
  stat_summary(fun = median, fun.min = median, fun.max = median,
               geom='crossbar', colour = "black", width=0.25, position = position_dodge(width = .25)) &
  ggpubr::stat_compare_means(method = "wilcox.test", comparisons = my_comparisons, 
                             aes(label = "..p.adj.."), label.y.npc = 0.95, label.x.npc = 0.5, size=4)&
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5),
        axis.text = element_text(size=18),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20),
        legend.position = "none") &
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1)))
```

```{r fig.width=12, fig.height=8}
VlnPlot(
  seurat_obj,
  features = "Denisenkoatal_up_genes_UCell",
  group.by = "celltypes",
  split.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0) &
  stat_summary(fun = median, fun.min = median, fun.max = median,
               geom='crossbar', colour = "black", width=0.25, position = position_dodge(width = .25)) &
  ggpubr::stat_compare_means(method = "wilcox.test", comparisons = my_comparisons, 
                             aes(label = "..p.adj.."), label.y.npc = 0.95, label.x.npc = 0.5, size=4)&
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text = element_text(size=18),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20)) &
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1)))


ggsave(paste0(path_r_figs_out, "/HumanColon_DenisenkoatalUPgenes_Ucell_Vlnbycells.svg"),
       width = 12, height = 8,
       dpi = 300)
```

```{r fig.width=10, fig.height=5}
plot_density(seurat_obj,
            features = 'Denisenkoatal_up_genes_UCell') + 
    facet_grid(.~seurat_obj$library_name)


ggsave(paste0(path_r_figs_out, "/HumanColon_DenisenkoatalUPgenes_Ucelldensity.svg"),
       width = 10, height = 5,
       dpi = 300)
```



<!-- ## Comparing signature with Fresh sDEgenes -->
<!-- ```{r} -->
<!-- path_r_objects <- here::here("04_GEX_analysis/4_DE_GO_GSEA/results/R_objects") -->
<!-- path_object_DEgenes <- paste0(path_r_objects, "/FIXnCUT_HumanColonpreserved_DEgenes_FixedCryo_vs_Cryo.rds") -->
<!-- DE_genes <- readRDS(path_object_DEgenes) -->

<!-- DE_genes_DOWN_df <- DE_genes %>%  -->
<!--     dplyr::filter(is_significant & avg_log2FC < 0) %>%  -->
<!--     dplyr::select(gene, avg_log2FC, p_val_adj) -->
<!-- DE_genes_DOWN_list <- DE_genes_DOWN_df$gene -->
<!-- ``` -->

<!-- ```{r fig.height=4, fig.width=4} -->
<!-- venndiagram <- VennDiagram::venn.diagram( -->
<!--   x = list(DE_genes_DOWN_list, Denisenkoatal_up_genes_list), -->
<!--   category.names = c(" ", " "), -->
<!--   filename = NULL, -->
<!--   fill = c("#023047", "#ffb703"), -->
<!--   lty = 'blank', -->

<!--   cex = 1.5, -->
<!--   fontfamily = "sans", -->

<!--   cat.cex = 1,   -->
<!--   cat.fontface = "bold", -->
<!--   cat.default.pos = "text", -->
<!--   cat.fontfamily = "sans" -->
<!-- ) -->

<!-- grid::grid.newpage() -->
<!-- grid::grid.draw(venndiagram) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- signature_genes_intersection <- intersect(Denisenkoatal_up_genes_list, DE_genes_DOWN_list) -->
<!-- print(signature_genes_intersection) -->
<!-- ``` -->


<!-- ```{r fig.width=12, fig.height=9} -->
<!-- DotPlot(seurat_obj, -->
<!--         features = signature_genes_intersection, -->
<!--         group.by = "celltypes", -->
<!--         split.by = "library_name", -->
<!--         cols = "RdBu", -->
<!--         dot.scale = 5 -->
<!--         ) + -->
<!--   coord_flip() + -->
<!--   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) + -->
<!--   theme(axis.title = element_blank(), -->
<!--         legend.position = "bottom") -->

<!-- ggsave(paste0(path_r_figs_out, "/HumanColon_DenisenkoetalUPgenes_DEgenes.svg"), -->
<!--        width = 12, height = 9, -->
<!--        dpi = 300) -->
<!-- ``` -->


# OTHER SIGNATURES: PBMC-related
## SIGNATURE: Sampling-time on single-cell PBMC by Massoni-Badosa et al.
```{r fig.width=20, fig.height=8}
path_signature_Massonietal <- here::here("04_GEX_analysis/5_gene_signatures/published_signatures/signature_Massonietal_samplingtime_hPBMC.csv")
Massonietal__genes <- read_csv(path_signature_Massonietal)

Massonietal_up_genes <- Massonietal__genes %>% 
  filter(direction == "up-regulated") %>% 
  dplyr::select(gene)

Massonietal_up_genes_list <- Massonietal_up_genes$gene
print(length(Massonietal_up_genes_list))
print(Massonietal_up_genes_list)
```


### Visualizing genes

```{r fig.width=14, fig.height=10}
DotPlot(seurat_obj,
        features = Massonietal_up_genes_list,
        group.by = "celltypes",
        split.by = "library_name",
        cols = "RdBu",
        dot.scale = 5
        ) +
   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### Computing signature
```{r fig.width=8, fig.height=5}
# UCELL GENE SIGNATURE SCORE
## Using positive and negative gene sets, GENE+ or GENE-
seurat_obj <- AddModuleScore_UCell(seurat_obj, 
                                   features=list("Massonietal_up_genes" = Massonietal_up_genes_list)
                                   )
```

### Visualizing Ucell
```{r fig.width=8, fig.height=5}
VlnPlot(
  seurat_obj,
  features = "Massonietal_up_genes_UCell",
  group.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0) &
  stat_summary(fun = median, fun.min = median, fun.max = median,
               geom='crossbar', colour = "black", width=0.25, position = position_dodge(width = .25)) &
  ggpubr::stat_compare_means(method = "wilcox.test", comparisons = my_comparisons, 
                             aes(label = "..p.adj.."), label.y.npc = 0.95, label.x.npc = 0.5, size=4)&
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5),
        axis.text = element_text(size=18),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20),
        legend.position = "none") &
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1)))
```

```{r fig.width=12, fig.height=8}
VlnPlot(
  seurat_obj,
  features = "Massonietal_up_genes_UCell",
  group.by = "celltypes",
  split.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0) &
  stat_summary(fun = median, fun.min = median, fun.max = median,
               geom='crossbar', colour = "black", width=0.25, position = position_dodge(width = .25)) &
  ggpubr::stat_compare_means(method = "wilcox.test", comparisons = my_comparisons, 
                             aes(label = "..p.adj.."), label.y.npc = 0.95, label.x.npc = 0.5, size=4)&
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text = element_text(size=18),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20)) &
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1)))


ggsave(paste0(path_r_figs_out, "/HumanColon_MassonietalUPgenes_Ucell_Vlnbycells.svg"),
       width = 12, height = 8,
       dpi = 300)
```

```{r fig.width=10, fig.height=5}
plot_density(seurat_obj,
            features = 'Massonietal_up_genes_UCell') + 
    facet_grid(.~seurat_obj$library_name)


ggsave(paste0(path_r_figs_out, "/HumanColon_MassonietalUPgenes_Ucelldensity.svg"),
       width = 10, height = 5,
       dpi = 300)
```


<!-- ### Comparing Massonietal. signature with Fresh sDEgenes -->
<!-- ```{r} -->
<!-- path_r_objects <- here::here("04_GEX_analysis/4_DE_GO_GSEA/results/R_objects") -->
<!-- path_object_DEgenes <- paste0(path_r_objects, "/FIXnCUT_HumanColonpreserved_DEgenes_FixedCryo_vs_Cryo.rds") -->
<!-- DE_genes <- readRDS(path_object_DEgenes) -->

<!-- DE_genes_DOWN_df <- DE_genes %>%  -->
<!--     dplyr::filter(is_significant & avg_log2FC < 0) %>%  -->
<!--     dplyr::select(gene, avg_log2FC, p_val_adj) -->
<!-- DE_genes_DOWN_list <- DE_genes_DOWN_df$gene -->
<!-- ``` -->

<!-- ```{r fig.height=4, fig.width=4} -->
<!-- venndiagram <- VennDiagram::venn.diagram( -->
<!--   x = list(DE_genes_DOWN_list, Massonietal_up_genes_list), -->
<!--   category.names = c(" ", " "), -->
<!--   filename = NULL, -->
<!--   fill = c("#023047", "#ffb703"), -->
<!--   lty = 'blank', -->

<!--   cex = 1.5, -->
<!--   fontfamily = "sans", -->

<!--   cat.cex = 1,   -->
<!--   cat.fontface = "bold", -->
<!--   cat.default.pos = "text", -->
<!--   cat.fontfamily = "sans" -->
<!-- ) -->

<!-- grid::grid.newpage() -->
<!-- grid::grid.draw(venndiagram) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- signature_genes_intersection <- intersect(Massonietal_up_genes_list, DE_genes_DOWN_list) -->
<!-- print(signature_genes_intersection) -->
<!-- ``` -->


<!-- ```{r fig.width=12, fig.height=9} -->
<!-- DotPlot(seurat_obj, -->
<!--         features = signature_genes_intersection, -->
<!--         group.by = "celltypes", -->
<!--         split.by = "library_name", -->
<!--         cols = "RdBu", -->
<!--         dot.scale = 5 -->
<!--         ) + -->
<!--   coord_flip() + -->
<!--   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) + -->
<!--   theme(axis.title = element_blank(), -->
<!--         legend.position = "bottom") -->

<!-- ggsave(paste0(path_r_figs_out, "/HumanColon_MassonietalUPgenes_DEgenes.svg"), -->
<!--        width = 12, height = 8, -->
<!--        dpi = 300) -->
<!-- ``` -->



## SIGNATURE: Ex-vivo blood handling on microarrays by Baechler et al.
```{r fig.width=20, fig.height=8}
path_signature_Baechleretal <- here::here("04_GEX_analysis/5_gene_signatures/published_signatures/signature_Baechleretal_storagetime_hPBMC_microarrays.csv")
Baechleretal__genes <- read_csv(path_signature_Baechleretal)

Baechleretal__genes <- distinct(Baechleretal__genes)
Baechleretal_up_genes <- Baechleretal__genes %>% 
  filter(direction == "up-regulated") %>% 
  dplyr::select(gene)

Baechleretal_up_genes_list <- Baechleretal_up_genes$gene
print(length(Baechleretal_up_genes_list))
print(Baechleretal_up_genes_list)
```


### Visualizing genes
```{r fig.width=14, fig.height=10}
DotPlot(seurat_obj,
        features = Baechleretal_up_genes_list,
        group.by = "celltypes",
        split.by = "library_name",
        cols = "RdBu",
        dot.scale = 5
        ) +
   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### Computing signature
```{r fig.width=8, fig.height=5}
# UCELL GENE SIGNATURE SCORE
## Using positive and negative gene sets, GENE+ or GENE-
seurat_obj <- AddModuleScore_UCell(seurat_obj, 
                                   features=list("Baechleretal_up_genes" = Baechleretal_up_genes_list)
                                   )
```

### Visualizing Ucell
```{r fig.width=8, fig.height=5}
VlnPlot(
  seurat_obj,
  features = "Baechleretal_up_genes_UCell",
  group.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0) &
  stat_summary(fun = median, fun.min = median, fun.max = median,
               geom='crossbar', colour = "black", width=0.25, position = position_dodge(width = .25)) &
  ggpubr::stat_compare_means(method = "wilcox.test", comparisons = my_comparisons, 
                             aes(label = "..p.adj.."), label.y.npc = 0.95, label.x.npc = 0.5, size=4)&
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5),
        axis.text = element_text(size=18),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20),
        legend.position = "none") &
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1)))
```

```{r fig.width=12, fig.height=8}
VlnPlot(
  seurat_obj,
  features = "Baechleretal_up_genes_UCell",
  group.by = "celltypes",
  split.by = "library_name",
  cols = palette_color_factor, 
  pt.size = 0) &
  stat_summary(fun = median, fun.min = median, fun.max = median,
               geom='crossbar', colour = "black", width=0.25, position = position_dodge(width = .25)) &
  ggpubr::stat_compare_means(method = "wilcox.test", comparisons = my_comparisons, 
                             aes(label = "..p.adj.."), label.y.npc = 0.95, label.x.npc = 0.5, size=4)&
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text = element_text(size=18),
        axis.title.x = element_blank(),
        strip.text = element_text(size=20)) &
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.1)))


ggsave(paste0(path_r_figs_out, "/HumanColon_BaechleretalUPgenes_Ucell_Vlnbycells.svg"),
       width = 12, height = 8,
       dpi = 300)
```

```{r fig.width=10, fig.height=5}
plot_density(seurat_obj,
            features = 'Baechleretal_up_genes_UCell') + 
    facet_grid(.~seurat_obj$library_name)


ggsave(paste0(path_r_figs_out, "/HumanColon_BaechleretalUPgenes_Ucelldensity.svg"),
       width = 10, height = 5,
       dpi = 300)
```


<!-- ### Comparing Baechleretal. signature with Fresh sDEgenes -->
<!-- ```{r} -->
<!-- path_r_objects <- here::here("04_GEX_analysis/4_DE_GO_GSEA/results/R_objects") -->
<!-- path_object_DEgenes <- paste0(path_r_objects, "/FIXnCUT_HumanColonpreserved_DEgenes_FixedCryo_vs_Cryo.rds") -->
<!-- DE_genes <- readRDS(path_object_DEgenes) -->

<!-- DE_genes_DOWN_df <- DE_genes %>%  -->
<!--     dplyr::filter(is_significant & avg_log2FC < 0) %>%  -->
<!--     dplyr::select(gene, avg_log2FC, p_val_adj) -->
<!-- DE_genes_DOWN_list <- DE_genes_DOWN_df$gene -->
<!-- ``` -->

<!-- ```{r fig.height=4, fig.width=4} -->
<!-- venndiagram <- VennDiagram::venn.diagram( -->
<!--   x = list(DE_genes_DOWN_list, Baechleretal_up_genes_list), -->
<!--   category.names = c(" ", " "), -->
<!--   filename = NULL, -->
<!--   fill = c("#023047", "#ffb703"), -->
<!--   lty = 'blank', -->

<!--   cex = 1.5, -->
<!--   fontfamily = "sans", -->

<!--   cat.cex = 1,   -->
<!--   cat.fontface = "bold", -->
<!--   cat.default.pos = "text", -->
<!--   cat.fontfamily = "sans" -->
<!-- ) -->

<!-- grid::grid.newpage() -->
<!-- grid::grid.draw(venndiagram) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- signature_genes_intersection <- intersect(Baechleretal_up_genes_list, DE_genes_DOWN_list) -->
<!-- print(signature_genes_intersection) -->
<!-- ``` -->


<!-- ```{r fig.width=12, fig.height=8} -->
<!-- DotPlot(seurat_obj, -->
<!--         features = signature_genes_intersection, -->
<!--         group.by = "celltypes", -->
<!--         split.by = "library_name", -->
<!--         cols = "RdBu", -->
<!--         dot.scale = 5 -->
<!--         ) + -->
<!--   coord_flip() + -->
<!--   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) + -->
<!--   theme(axis.title = element_blank(), -->
<!--         legend.position = "bottom") -->

<!-- ggsave(paste0(path_r_figs_out, "/HumanColon_BaechleretalUPgenes_DEgenes.svg"), -->
<!--        width = 12, height = 8, -->
<!--        dpi = 300) -->
<!-- ``` -->


# Session Info
```{r}
sessionInfo()
```