---
title: "GEX Quality control"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction
As shown in the previous notebook, there are no major differences among all libraries processed and sequenced at CNAG in terms of sequencing depth, number of recovered cells, mean reads per cell, median genes per cell, among others. For this reason, we will merge all libraries together and we will perform a single quality control (QC) including all libraries. Therefore, unique QC thresholds will be determined and applied to all libraries.

## Objective
In this Rmarkdown document, we are going to perform a quality control (QC) in order to filter out low-quality cells and genes. To do so, we will follow the current best practices for scRNA-seq QC described in [Luecken al. Mol Syst Biol (2018)](https://doi.org/10.15252/msb.20188746).

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
library(ggrepel)
library(UpSetR)
library(gt)
library(grid)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
sub_folder <- "SCGTEST_32"
path_project_metadata <- here::here("01_cellranger_mapping/data/FIXnCUT_metadata.csv")
path_cellranger_data <- here::here(paste0("01_cellranger_mapping/subprojects/", sub_folder, "/jobs"))
#path_scrublet_data <- here::here("02_QC/2_doublet_detection/results/tables")
path_r_objects <- here::here("02_QC/results/R_objects")

# Functions
source(here::here("bin/utils.R"))
```


## Load data
The data used in this Rmarkdown document comes from **SCGTEST_32** dataset, filtered matrices were processed with `cellranger v6.1.1`, and the doublet predictions were obtained using `scrublet`.
```{r}
# Load all metadata information
metadata <- read.csv(path_project_metadata)

# Merging metadata_all tables
metadata <- metadata %>% 
  filter(subproject == "SCGTEST_32" & type == "cDNA")

print("Libraries metadata")
DT::datatable(metadata, options = list(scrollX = TRUE))
```


```{r}
# Getting list of gem_ids
gemid_list <- sort(unique(metadata$gem_id))

# Create Seurat object
## Counts data
seurat_obj_list <- purrr::map(gemid_list, function(gem_id) {
  filtered_matrix_path <- paste(path_cellranger_data, gem_id, gem_id, 
                                    "outs/per_sample_outs", gem_id, "count/sample_feature_bc_matrix", sep = "/")
  expression_matrix <- Read10X(data.dir = filtered_matrix_path)
  seurat = CreateSeuratObject(counts = expression_matrix, project = gem_id)
  # Add gem_id to each barcode (cell)
  seurat$gem_id <- as.factor(gem_id)
  seurat
})
names(seurat_obj_list) <- gemid_list
```


## Merge seurat objects
```{r}
# Merge Seurat objects
seurat_obj <- merge(x = seurat_obj_list[[1]],
                    y = seurat_obj_list[2:length(seurat_obj_list)],
                    add.cell.ids = gemid_list)

# Remove Seurat object list
rm(seurat_obj_list)
```

## Add library metadata
Now, we will add metadata_all information of the different samples.
```{r}
# Create dataframe with seurat metadata_all (by barcodes) and sample metadata_all
seurat_obj_metadata <- left_join(seurat_obj@meta.data, metadata, by = "gem_id")

# Add only relevant metadata_all variables & convert into factors
seurat_obj$library_name <- as.factor(seurat_obj_metadata$library_name)
seurat_obj$donor_id <- as.factor(seurat_obj_metadata$organism)
seurat_obj$tissue <- as.factor(seurat_obj_metadata$tissue)
seurat_obj$disease <- as.factor(seurat_obj_metadata$disease)
seurat_obj$sample_protocol <- as.factor(seurat_obj_metadata$sample_protocol)
seurat_obj$fixation_time <- as.factor(seurat_obj_metadata$fixation_time)
```

## Save merged data
```{r}
# Save Merged Seurat object
saveRDS(seurat_obj, str_c(path_r_objects, "/scgtest32_merged.rds"))
#seurat_obj <- readRDS(str_c(path_r_objects, "/scgtest32_merged.rds"))
```


# Exclusive genes
```{r}
seurat_obj_list <- SplitObject(seurat_obj, split.by = "library_name")
```

```{r}
seurat_obj_list <- purrr::map(seurat_obj_list, function(seurat) {
  # Number of cells expressing each gene, excluding 0 counts
  n_cells <- Matrix::rowSums(seurat[["RNA"]]@counts > 0)
  
  # Subset genes that are expressed in data
  keep_genes <- rownames(seurat)[n_cells > 0]

  # Filtering out genes expressed in few cells
  seurat_obj <- subset(seurat, features = keep_genes)
  seurat_obj  
})
seurat_obj_list
```


## Upset Plot
https://github.com/hms-dbmi/UpSetR

```{r fig.height=5, fig.width=8}
UpSetR::upset(
  UpSetR::fromList(list(
    "Colon_fixed_24h" = rownames(seurat_obj_list$Colon_fixed_24h),
    "Colon_fresh" = rownames(seurat_obj_list$Colon_fresh),
    "Lung_fixed_24h" = rownames(seurat_obj_list$Lung_fixed_24h),
    "Lung_fresh" = rownames(seurat_obj_list$Lung_fresh)
  )),
  order.by = "freq",
  #group.by = "sets",
  mainbar.y.label = "# Genes", sets.x.label = "# Genes / sample",
  point.size = 4, line.size = 1,
  empty.intersections = "on",
   queries = list(
     list(query = intersects, params = list("Colon_fixed_24h", "Lung_fixed_24h"), color = "orange", active = T),
     list(query = intersects, params = list("Colon_fresh", "Lung_fresh"), color = "blue", active = T),
     list(query = intersects, params = list("Colon_fixed_24h"), color = "red", active = T),
     list(query = intersects, params = list("Lung_fixed_24h"), color = "red", active = T),
     list(query = intersects, params = list("Colon_fresh"), color = "green", active = T),
     list(query = intersects, params = list("Lung_fresh"), color = "green", active = T)
     )
  )
```

## Identifyng genes & GSEA
```{r}
fromList_togetdata <- function (input) {
  # Same as original fromList()...
  elements <- unique(unlist(input))
  data <- unlist(lapply(input, function(x) {
      x <- as.vector(match(elements, x))
      }))
  data[is.na(data)] <- as.integer(0)
  data[data != 0] <- as.integer(1)
  data <- data.frame(matrix(data, ncol = length(input), byrow = F))
  data <- data[which(rowSums(data) != 0), ]
  names(data) <- names(input)
  # ... Except now it conserves your original value names!
  row.names(data) <- elements
  return(data)
  }

upset_data <- fromList_togetdata(list(
    "Colon_fixed_24h" = rownames(seurat_obj_list$Colon_fixed_24h),
    "Colon_fresh" = rownames(seurat_obj_list$Colon_fresh),
    "Lung_fixed_24h" = rownames(seurat_obj_list$Lung_fixed_24h),
    "Lung_fresh" = rownames(seurat_obj_list$Lung_fresh)
  ))


# TECHNOLOGY UNIQUE
genes_fresh_unique <- upset_data %>% 
  filter(Colon_fresh == 1 & Lung_fresh == 1 & Colon_fixed_24h == 0 & Lung_fixed_24h == 0)
genes_fresh_unique <- rownames(genes_fresh_unique)
genes_fresh_unique

genes_fixncut_unique <- upset_data %>% 
  filter(Colon_fresh == 0 & Lung_fresh == 0 & Colon_fixed_24h == 1 & Lung_fixed_24h == 1)
genes_fixncut_unique <- rownames(genes_fixncut_unique)
genes_fixncut_unique


# TISSUE-TECH UNIQUE
genes_lungfresh_unique <- upset_data %>% 
  filter(Colon_fresh == 0 & Lung_fresh == 1 & Colon_fixed_24h == 0 & Lung_fixed_24h == 0)
genes_lungfresh_unique <- rownames(genes_lungfresh_unique)
genes_lungfresh_unique

genes_lungfixncut_unique <- upset_data %>% 
  filter(Colon_fresh == 0 & Lung_fresh == 0 & Colon_fixed_24h == 0 & Lung_fixed_24h == 1)
genes_lungfixncut_unique <- rownames(genes_lungfixncut_unique)
genes_lungfixncut_unique


genes_colonfresh_unique <- upset_data %>% 
  filter(Colon_fresh == 1 & Lung_fresh == 0 & Colon_fixed_24h == 0 & Lung_fixed_24h == 0)
genes_colonfresh_unique <- rownames(genes_colonfresh_unique)
genes_colonfresh_unique

genes_colonfixncut_unique <- upset_data %>% 
  filter(Colon_fresh == 0 & Lung_fresh == 0 & Colon_fixed_24h == 1 & Lung_fixed_24h == 0)
genes_colonfixncut_unique <- rownames(genes_colonfixncut_unique)
genes_colonfixncut_unique
```

```{r}
library(enrichR)
```


```{r}
# Define comparison paths
#path_GO_results <- paste0(path_r_tables, "/scbatlle30_GOBP21_", comparison_name)
database <- "GO_Biological_Process_2021" # GO_Biological_Process_2021, KEGG_2021_Human, Reactome_2016

## EnrichR
enrichR_results <- run_enrichR_forGO(
  database = database,
  DE_genes = genes_lungfixncut_unique,
  adj_pval = 0.05,
  number_enriched_genes = 2,
  num_GO_size_min = 5,
  num_GO_size_max = 300,
  OR = 2
)

## Filter reliable results
enrichR_results_filtered <- enrichR_results %>% 
  dplyr::filter(is_reliable == TRUE)

## Plot
gg_enrichR_hallmark <- gg_enrichR_GO("", 
             database,
             enrichR_results_filtered, 
             n_top = 20)
print(gg_enrichR_hallmark)

# Save results
## Export list of DE genes
#path_hallmark <- paste0(path_r_tables, "/stromal_hallmark.xlsx")
#openxlsx::write.xlsx(myList, file = path_hallmark)  
```


# Session Info
```{r}
sessionInfo()
```