---
title: "Annotation"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction
After the pre-processing steps done in previous notebooks, we will now start with the downstream data analysis in order to extract biological insights and describe the underlying biology.

## Objective
In this Rmarkdown document, we are going to assign cell type clusters into different cell-types clusters based on the similarity of their gene expression profiles. Then, we will perform again process each object to define cell-type specific clusters with more resolution.

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
library(DT)
library(dittoSeq)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_project_metadata <- here::here("01_cellranger_mapping/data/FIXnCUT_metadata.csv")
path_r_objects <- here::here("03_clustering_annotation/results/R_objects")
path_r_tables <- here::here("03_clustering_annotation/results/tables")
path_r_figs <- here::here("03_clustering_annotation/results/figs")

# Functions
source(here::here("bin/utils.R"))

## Define resolution of choice
resolution <- 0.9
resolution_name <- paste0("RNA_snn_res.", resolution)

# Color palette
set.seed(1234) # for reproducibility
color_palette <- Polychrome::createPalette(35, c("#fc6060", "#74f774", "#7c7cfc"))
names(color_palette) <- NULL
#Polychrome::swatch(color_palette)
```

## Load data
The data used in this Rmarkdown document comes from **SCGTEST_34** dataset.
```{r}
# Load all metadata information
metadata <- read.csv(path_project_metadata)
metadata <- metadata %>% 
  filter(subproject == "SCGTEST_34" & tissue == "PBMC")

print("Libraries/Samples metadata")
DT::datatable(metadata, options = list(scrollX = TRUE))

# Load Seurat object
seurat_obj <- readRDS(paste0(path_r_objects, "/FIXnCUT_HumanPBMC_clustering_resolutions.rds"))
DefaultAssay(seurat_obj) <- "RNA"
Idents(seurat_obj) <- "RNA_snn_res.0.9"
seurat_obj

# Load biomarkers
seurat_obj_markers <- readRDS(paste0(path_r_objects, "/biomarkers_FIXnCUT_HumanPBMC_resolution0.9.rds"))
```


# Explore cell clusters
To visualize the cell clusters, there are a few different dimensionality reduction techniques, such as t-SNE and UMAP. These methods require to input a number of PCA dimensions (same number as the one used for the clustering). We will not run UMAP again as it was computed in the previous notebook.
```{r fig.width=14, fig.height=10}
# Clustering overview
gg_umap_clustering <- DimPlot(object = seurat_obj,
        group.by = resolution_name,
        label = TRUE,
        label.size = 6,
        cols = color_palette,
        pt.size=0.5
        ) & NoLegend()
gg_umap_clustering
```

Now, we will also explore clusters across conditions.
```{r fig.width=14, fig.height=5}
# Clusters across tissues
gg_clustering_by_umap <- DimPlot(
  object = seurat_obj,
  group.by = resolution_name,
  split.by = "sample_protocol",
  label = FALSE,
  cols = color_palette,
  ncol = 3
  )

gg_clustering_by_umap
```


## Cell clusters data exploration
```{r}
# Number of cells / cluster
table(seurat_obj@active.ident)

# Number of cells per cluster vs confounder variables
table(seurat_obj$library_name, seurat_obj@active.ident)

# Creating new annotation variable
seurat_obj$celltypes <- as.character(seurat_obj@active.ident)
Idents(seurat_obj) <- "celltypes"
```


## Assessing sources of uninteresting variation
Next, we will explore additional metrics, such as the number of UMIs and genes per cell, mitochondrial / ribosomal gene expression and scrublet doublet scores on the UMAP embedding. 

```{r fig.width=14, fig.height=10}
qc_metrics <-  c("nCount_RNA", "nFeature_RNA", "pct_mt", "pct_rb")

gg_qc_umap <- FeaturePlot(
  seurat_obj,
  reduction = "umap",
  features = qc_metrics,
  cols = c("lightgrey", "darkblue"), 
  order = TRUE,
  ncol = 2)

gg_qc_umap
```

```{r fig.width=10, fig.height=15}
gg_qc_vln <- VlnPlot(
  seurat_obj,
  features = qc_metrics,
  cols = color_palette, 
  pt.size = 0,
  ncol = 1)

gg_qc_vln
```

Finally, we will also check the cell-cycle phase by assessing S-phase and G2M-phase markers signature.
```{r fig.width=10, fig.height=10}
gg_phase_vln <- VlnPlot(
  seurat_obj,
  features = c("S.Score", "G2M.Score"),
  cols = color_palette, 
  pt.size = 0,
  ncol = 1)

gg_phase_vln
```

# Explore cluster biomarkers
```{r}
DefaultAssay(seurat_obj) <- "RNA"
```

## Table
```{r}
# Visualize top markers per cluster
top_all_markers <- seurat_obj_markers %>%
  group_by(cluster) %>%
  slice_max(n = 10, order_by = avg_log2FC)

top_all_markers
DT::datatable(top_all_markers, options = list(scrollX = TRUE))
```

## Dotplot
```{r fig.width=20, fig.height=6}
# Visualize top markers per cluster
top_all_markers <- seurat_obj_markers %>%
  group_by(cluster) %>%
  slice_max(n = 4, order_by = avg_log2FC)

top_genemarkers_list <- unique(top_all_markers$gene)

DotPlot(
  seurat_obj,
  features = top_genemarkers_list,
  dot.scale = 6) +
  RotatedAxis()
```

<!-- ## FeaturePlot and ViolinPlot -->
<!-- With the cells clustered and based on the previous cell-type markers, we will select the most informative gene markers for each cluster and some well-known cell markers and project them in the UMAP. Basically, we are looking for consistency of expression of the markers across the clusters (e.g. if there are two markers for a cell type and only one of them is expressed in a cluster, we cannot reliably assign that cluster to the cell type). -->
<!-- ```{r} -->
<!-- # Selected gene markers -->
<!-- selected_markers <- seurat_obj_markers %>% -->
<!--   group_by(cluster) %>% -->
<!--   slice_max(n = 4, order_by = avg_log2FC) %>% -->
<!--   group_split() -->
<!-- ``` -->

<!-- ```{r fig.width=7, fig.height=5} -->
<!-- # Feature plots -->
<!-- purrr::map(selected_markers, function(cluster) { -->
<!--   FeaturePlot(seurat_obj, -->
<!--               features = cluster$gene, -->
<!--               label = TRUE, -->
<!--               label.size = 3, -->
<!--               ncol = 2) -->
<!-- }) -->
<!-- ``` -->


## Gene expression

```{r fig.width=25, fig.height=20}
lineage_genes <- c("PTPRC", # CD45+, immune cells
                   "CD3D", "CD3E", # T cells
                   "CD4",  "IRF7", "FOXP3", # CD4+ T cells, Treg
                   "CD8A", "CD8B", "NKG7", # CD8+ T cells, cytotoxic
                   "KLRD1", "GZMA", # NK cells
                   "CD79A", # B cells
                   "MS4A1", "JCHAIN", # Naive vs Plasma cells
                   "LYZ", "CD14", "FCGR3A", "PLAC8", "CD209", # Mono/macro/DCs
                   "S100A9", "CD200R3", # Neutrophils, Mast cells
                   "MKI67", "TOP2A", "HLA-DRA" # Cycling, MHC-antigen presenting
                   )

FeaturePlot(seurat_obj,
            features = lineage_genes,
            label = T,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 5)

VlnPlot(
  seurat_obj,
  features = lineage_genes,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 5)
```

### Myeloid

```{r fig.width=20, fig.height=15}
mono_genes <- c("CD14", "ITGAM", "CCR2", "SELL", "CD36",
                "FCGR3A", "CX3CR1", "ITGAL", "HLA-DRA", "AIF1",
                "CD68", "CD163") 
# classical CD14/CCR2/Cd11b == ITGAM, non-classical CD16 == FCGR3A, intermediate 
FeaturePlot(seurat_obj,
            features = mono_genes,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 4
)

VlnPlot(
  seurat_obj,
  features = mono_genes,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 4)
```


#### Subclustering
```{r fig.width=7, fig.height=5}
# Find subclusters
seurat_obj <- FindSubCluster(
  seurat_obj,
  cluster = "7",
  graph.name = "RNA_snn",
  subcluster.name = "RNA_snn_res.0.9_sub7",
  resolution = 0.1
)

# Plotting
DimPlot(
  seurat_obj,
  group.by =  "RNA_snn_res.0.9_sub7",
  label = TRUE,
  label.size = 5,
  cols = color_palette
  ) + labs(title="")
```


```{r fig.width=20, fig.height=15}
VlnPlot(
  seurat_obj,
  features = mono_genes,
  group.by = "RNA_snn_res.0.9_sub7",
  cols = color_palette, 
  pt.size = 0, 
  ncol = 4)
```


```{r fig.width=7, fig.height=5}
# Replacing new subclusters labels
seurat_obj$celltypes[seurat_obj$celltypes == "7"] <- as.character(
  seurat_obj$RNA_snn_res.0.9_sub7[seurat_obj$celltypes == "7"])
```

```{r fig.width=15, fig.height=10}
cDC1_genes <- c("CLEC9A", "C1orf54", "HLA-DPA1", "CADM1", "CAMK2D", "XCR1")
FeaturePlot(seurat_obj,
            features = cDC1_genes,
            #split.by = "therapy",
            pt.size = 0.25,
            ncol = 3
)

VlnPlot(
  seurat_obj,
  features = cDC1_genes,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 3)
```

```{r fig.width=15, fig.height=5}
cDC2_genes <- c("CD1C", "FCER1A", "CLEC10A")
FeaturePlot(seurat_obj,
            features = cDC2_genes,
            #split.by = "tissue",
            pt.size = 0.25,
            ncol = 3
)

VlnPlot(
  seurat_obj,
  features = cDC2_genes,
  pt.size = 0, 
  ncol = 3)
```

```{r fig.width=15, fig.height=10}
DC3_genes <- c("S100A9", "S100A8", "VCAN", "LYZ", "ANXA1")
FeaturePlot(seurat_obj,
            features = DC3_genes,
            #split.by = "tissue",
            pt.size = 0.25,
            ncol = 3
)

VlnPlot(
  seurat_obj,
  features = DC3_genes,
  pt.size = 0, 
  ncol = 3)
```

```{r fig.width=15, fig.height=10}
DC4_genes <- c("FCGR3A", "FTL", "SERPINA1", "LST1", "IFITM2")
FeaturePlot(seurat_obj,
            features = DC4_genes,
            #split.by = "tissue",
            pt.size = 0.25,
            ncol = 3
)

VlnPlot(
  seurat_obj,
  features = DC4_genes,
  pt.size = 0, 
  ncol = 3)
```


```{r fig.width=15, fig.height=10}
pDC_genes <- c("GZMB", "JCHAIN", "PTGDS", "SERPINF1", "IL3RA", "ITM2C", "CLEC4C", "NRP1", "IRF7")
FeaturePlot(seurat_obj,
            features = pDC_genes,
            #split.by = "tissue",
            pt.size = 0.25,
            ncol = 3
)

VlnPlot(
  seurat_obj,
  features = pDC_genes,
  pt.size = 0, 
  ncol = 3)
```



### B Lymphocytes
```{r fig.width=20, fig.height=10}
Bnaive_genes <- c("MS4A1", "CD79A", "CD79B", "HLA-DRA",
                  "CD69", "FCER2", "CD1C", "CD48", "CD81", "JUN", "CD86")

FeaturePlot(seurat_obj,
            features = Bnaive_genes,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 4
)

VlnPlot(
  seurat_obj,
  features = Bnaive_genes,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 4)
```


```{r fig.width=20, fig.height=15}
Plasma_genes <- c("CD27", "JCHAIN", "XBP1", "CD38", "RF4", "XBP1", "IGHD", "IGHM", "MKI67", "HLA-DRA", "MZB1", "SLAMF7", "SSR4")
# RF4/XBP1/PRDM1 = early PC /PCprec
# SLAMF7/MZB1/SSR4+ Ig = Mature
# when proliferating = Plasmablast

FeaturePlot(seurat_obj,
            features = Plasma_genes,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 4
)

VlnPlot(
  seurat_obj,
  features = Plasma_genes,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 4)
```

### T Lymphocytes

```{r fig.width=30, fig.height=15}
T_genes <- c("CD3D", "CD3E",# T cells
            "CD4", "FOXP3", # CD4+ T cells, Treg
            "CD8A", "CD8B", "NKG7", "GZMK", "GZMB", # CD8+ T cells, cytotoxic
            "SELL", "CCR7", "LEF1", "CD27", "CD69", "HLA-DRB1"
             )

FeaturePlot(seurat_obj,
            features = T_genes,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.25,
            ncol = 5)

VlnPlot(
  seurat_obj,
  features = T_genes,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 5)
```

```{r fig.width=14, fig.height=10}
CD4naive_genes <- c("SELL", "CCR7", "LEF1", "NOSIP", "RGS10", "OXNAD1")

FeaturePlot(seurat_obj,
            features = CD4naive_genes,
            #split.by = "therapy",
            pt.size = 0.5,
            ncol = 2
)

VlnPlot(
  seurat_obj,
  features = CD4naive_genes,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 2)
```

```{r fig.width=25, fig.height=15}
memoryT_genes <- c("KLRB1", "KLRG1", "IL7R", "PRDM1", "PDCD1", "CCR7", "IL3RA", "EGR1", "ICOS", "PASK", "CD69", "FAS", "PD1", "CD27", "CD28")

FeaturePlot(seurat_obj,
            features = memoryT_genes,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.25,
            ncol = 5
)

VlnPlot(
  seurat_obj,
  features = memoryT_genes,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 5)
```


```{r fig.width=14, fig.height=15}
Treg_genes <- c("FOXP3", "CTLA4", "TIGIT", "DUSP4", "IL2RA", "LAG3", 
                "RORC", "IL17A", "IL17F", "CCR6", "CCR4", "CXCR3", "IFNG")

FeaturePlot(seurat_obj,
            features = Treg_genes,
            #split.by = "therapy",
            order = T,
            pt.size = 0.5,
            ncol = 3
)

VlnPlot(
  seurat_obj,
  features = Treg_genes,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 3)
```

```{r fig.width=14, fig.height=10}
cytotoxic_genes <- c("GZMK", "NKG7", "GZMA", "GNLY", "GZMB", "PRF1", "KLRB1", "KLRG1")

FeaturePlot(seurat_obj,
            features = cytotoxic_genes,
            #split.by = "therapy",
            pt.size = 0.5,
            ncol = 3
)

VlnPlot(
  seurat_obj,
  features = cytotoxic_genes,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 3)
```


### Innate cells
```{r fig.width=14, fig.height=10}
MAIT_genes <- c("KLRB1", "NKG7", "GZMK", "SLC4A10", "GZMA", "CXCR6", "NCR3",
                "TRAV1-2", "RORC", "RORA", "CCR6", "IL7R", "IL23R", "EOMES", "STAT4")

FeaturePlot(seurat_obj,
            features = MAIT_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.5,
            ncol = 4
)

VlnPlot(seurat_obj,
            features = MAIT_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0,
            ncol = 4
)
```


```{r fig.width=14, fig.height=10}
gdT_genes <- c("TRDC", "TRGC1", "TRGC2", "TRDV1", "TRDV2", "GZMH", "GZMK", "NKG7", "KLRD1", "KLRG1",  "KLRC1")

FeaturePlot(seurat_obj,
            features = gdT_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.5,
            ncol = 3
)

VlnPlot(seurat_obj,
            features = gdT_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.01,
            ncol = 3
)
```

```{r fig.width=14, fig.height=10}
NKT_genes <- c("CD3E", "CD3D", "CD8B", "NCAM1", "KLRB1", "CD200", "CD1C", "FCGR3A")

FeaturePlot(seurat_obj,
            features = NKT_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.5,
            ncol = 3
)

VlnPlot(seurat_obj,
            features = NKT_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0,
            ncol = 3
)
```

```{r fig.width=20, fig.height=10}
ILC_genes <- c("GATA3", "RORA", "IL7R", 
               "MAF", "RORC", "ZNF683", "TBX21", "CCR6", "IRF4", "EOMES", "IKZF3") 

FeaturePlot(seurat_obj,
            features = ILC_genes,
            label = T,
            #split.by = "tissue",
            order = F,
            pt.size = 0.25,
            ncol = 4
)

VlnPlot(seurat_obj,
            features = ILC_genes,
            #split.by = "tissue",
            #order = T,
            pt.size = 0.25,
            ncol = 4
)
```

```{r fig.width=14, fig.height=10}
NKCD56bright_genes <- c("GPR183", "IL7R", "GZMK", "CD62L", "CD2", "KLRC1", "KLRD1", "NCAM1", "XCL1", "XCL2")

FeaturePlot(seurat_obj,
            features = NKCD56bright_genes,
            #split.by = "tissue",
            pt.size = 0.25,
            ncol = 3
)

VlnPlot(
  seurat_obj,
  features = NKCD56bright_genes,
  cols = color_palette, 
  pt.size = 0.1, 
  ncol = 3)
```


```{r fig.width=25, fig.height=15}
NKCD56dim_genes <- c("FCGR3A", "FCER1G", "GZMB", "GZMH", "CCL3", "CCL4", "CCL5", "NCR3", "CXCR4", 
                     "HIST1H1E", "ZEB2", "ANXA2", "CX3CR1", "ITGB2", "IL32", "CD7", "CFL1", "CST7", "PRF1", "LILRB1", "KIR", "KLRC2", "SH2D1B", "B3GAT1", "FCRL6") 
# High CD16 == FCGR3A, CD18 == ITGB2, PRF1, FCRL6, NCR3,CD57 == B3GAT1
FeaturePlot(seurat_obj,
            features = NKCD56dim_genes,
            #order = TRUE,
            #split.by = "tissue",
            pt.size = 0.25,
            ncol = 5
)

VlnPlot(
  seurat_obj,
  features = NKCD56dim_genes,
  cols = color_palette, 
  pt.size = 0.1, 
  ncol = 5)
```

### Others
```{r fig.width=20, fig.height=10}
platelets <- c("PPBP", "PF4", "GNG11", "ITGA2B", "GP9", "GP1BA", "ITGB3") 

FeaturePlot(seurat_obj,
            features = platelets,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 4
)

VlnPlot(
  seurat_obj,
  features = platelets,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 4)
```

```{r fig.width=20, fig.height=10}
doublets <- c("CD3D", "CD3E", "CD14", "STMN1", "CD79A", "CD79B") 

FeaturePlot(seurat_obj,
            features = doublets,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 4
)

VlnPlot(
  seurat_obj,
  features = doublets,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 4)
```

```{r fig.width=10, fig.height=5}
progenitors <- c("CD34", "SOX4") 

FeaturePlot(seurat_obj,
            features = progenitors,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 2
)

VlnPlot(
  seurat_obj,
  features = progenitors,
  cols = color_palette, 
  pt.size = 0.1, 
  ncol = 2)
```

# Annotation
According to the previous cell-type identities enriched in several gene markers, we will establish the following annotation:

```{r}
# GENERAL
Idents(seurat_obj) <- "celltypes"

seurat_obj <- RenameIdents(
  seurat_obj,
 "0" = "Classical Monocytes",
 "1" = "Naive CD4+ T cells",
 "2" = "NK cells",
 "3" = "Naive CD8+ T cells",
 "4" = "Cytotoxic CD8+ T cells",
 "5" = "Central Memory CD4+ T cells",
 "6" = "Effector Memory CD4+ T cells",
 "7_0" = "Non-classical Monocytes",
 "7_1" = "Intermediate Monocytes",
 "8" = "MAIT",
 "9" = "Transitional Memory CD4+ T cells",
 "10" = "Low quality cells",
 "11" = "Tregs",
 "12" = "Memory B cells", 
 "13" = "DC1/DC2",
 "14" = "Naive B cells",
 "15" = "gdT cells",
 "16" = "Complex Platelets-T cells",
 "17" = "Doublets",
 "18" = "Complex Platelets-Myeloid cells",
 "19" = "Cycling cells",
 "20" = "pDCs",
 "21" = "Progenitor cells"
 )

# Creating new annotation variable
seurat_obj$celltypes <- as.factor(as.character(seurat_obj@active.ident))
seurat_obj$celltypes <- factor(seurat_obj$celltypes, levels = sort(levels(seurat_obj$celltypes)))
```


```{r}
# Exploring number of celltypes
table(seurat_obj$celltypes, seurat_obj$condition)
table(seurat_obj$sample_protocol)
```


## UMAP celltypes
```{r fig.width=16, fig.height=10}
# Clustering overview
gg_umap_clustering_annot <- DimPlot(object = seurat_obj,
        group.by = "celltypes",
        repel = TRUE,
        cols = color_palette
        ) +
  labs(title = "Cell types") +
  guides(color = guide_legend(override.aes = list(size = 5), ncol = 1))
gg_umap_clustering_annot

ggsave(filename = paste0(path_r_figs, "/FIXnCUT_HumanPBMC_clustering_annotation_umap.png"),
       plot = gg_umap_clustering_annot)
```

Now, we will also explore clusters across sample.
```{r fig.width=15, fig.height=5}
DimPlot(
  object = seurat_obj,
  group.by = "celltypes",
  split.by = "sample_protocol",
  label = FALSE,
  cols = color_palette
  )
```

## Cell proportions
```{r fig.width=12, fig.height=6}
gg_cell_num <- dittoBarPlot(seurat_obj, 
                            "celltypes",
                            group.by = "condition",
                            scale = "count",
                            retain.factor.levels = TRUE,
                            legend.show = FALSE,
                            color.panel = color_palette)

gg_cell_prop <- dittoBarPlot(seurat_obj, 
                             "celltypes",
                             group.by = "condition",
                             scale = "percent",
                            retain.factor.levels = TRUE,
                             color.panel = color_palette)

gg_cell_num + gg_cell_prop
```

## Save Seurat object
```{r}
# Save lineage annotations
saveRDS(seurat_obj$celltypes,
        paste0(path_r_objects, "/FIXnCUT_HumanPBMC_cellsnames.rds"))

# Save Seurat object
saveRDS(seurat_obj, paste0(path_r_objects, "/FIXnCUT_HumanPBMC_clustering_annotation.rds"))
#seurat_obj <- readRDS(paste0(path_r_objects, "/FIXnCUT_HumanPBMC_clustering_annotation.rds"))
```



# Remove uninteresting cells
```{r}
Idents(seurat_obj) <- "celltypes"
cells_to_remove <- c("Doublets", "Low quality cells", "Complex Platelets-Myeloid cells", "Complex Platelets-T cells")

seurat_obj_cleaned <- subset(seurat_obj, idents = cells_to_remove, invert = TRUE)
seurat_obj_cleaned$celltypes <- as.factor(as.character(seurat_obj_cleaned$celltypes))

table(seurat_obj_cleaned$celltypes)
length(unique(seurat_obj_cleaned$celltypes))
rm(seurat_obj)
```


## UMAP celltypes
```{r fig.width=16, fig.height=10}
# Clustering overview
gg_umap_clustering_annot <- DimPlot(object = seurat_obj_cleaned,
        group.by = "celltypes",
        repel = TRUE,
        cols = color_palette
        ) +
  labs(title = "Cell types") +
  guides(color = guide_legend(override.aes = list(size = 5), ncol = 1))
gg_umap_clustering_annot

ggsave(filename = paste0(path_r_figs, "/FIXnCUT_HumanPBMC_clustering_annotation_cleaned_umap.png"),
       plot = gg_umap_clustering_annot)
```

Now, we will also explore clusters across sample.
```{r fig.width=15, fig.height=5}
DimPlot(
  object = seurat_obj_cleaned,
  group.by = "celltypes",
  split.by = "sample_protocol",
  label = FALSE,
  cols = color_palette
  )
```

## Cell proportions
```{r fig.width=12, fig.height=6}
gg_cell_num <- dittoBarPlot(seurat_obj_cleaned, 
                            "celltypes",
                            group.by = "condition",
                            scale = "count",
                            retain.factor.levels = TRUE,
                            legend.show = FALSE,
                            color.panel = color_palette)

gg_cell_prop <- dittoBarPlot(seurat_obj_cleaned, 
                             "celltypes",
                             group.by = "condition",
                             scale = "percent",
                            retain.factor.levels = TRUE,
                             color.panel = color_palette)

gg_cell_num + gg_cell_prop
```

## Save Seurat object CLEANED
```{r}
# Save lineage annotations
saveRDS(seurat_obj_cleaned$celltypes,
        paste0(path_r_objects, "/FIXnCUT_HumanPBMC_cellsnames_cleaned.rds"))

# Save Seurat object
saveRDS(seurat_obj_cleaned, paste0(path_r_objects, "/FIXnCUT_HumanPBMC_clustering_annotation_cleaned.rds"))
#seurat_obj <- readRDS(paste0(path_r_objects, "/FIXnCUT_HumanPBMC_clustering_annotation_cleaned.rds"))
```


# Session Info
```{r}
sessionInfo()
```
