---
title: "Annotation to subcluster"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction
After the pre-processing steps done in previous notebooks, we will now start with the downstream data analysis in order to extract biological insights and describe the underlying biology.

## Objective
In this Rmarkdown document, we are going to assign cell type clusters into different cell-types clusters based on the similarity of their gene expression profiles. Then, we will perform again process each object to define cell-type specific clusters with more resolution.

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
library(DT)
library(dittoSeq)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_project_metadata <- here::here("01_cellranger_mapping/data/FIXnCUT_metadata.csv")
path_r_objects <- here::here("03_clustering_annotation/results/R_objects")
path_r_tables <- here::here("03_clustering_annotation/results/tables")
path_r_figs <- here::here("03_clustering_annotation/results/figs")

# Functions
source(here::here("bin/utils.R"))

## Define resolution of choice
resolution <- 0.3
resolution_name <- paste0("RNA_snn_res.", resolution)

# Color palette
set.seed(1234) # for reproducibility
color_palette <- Polychrome::createPalette(35, c("#fc6060", "#74f774", "#7c7cfc"))
names(color_palette) <- NULL
#Polychrome::swatch(color_palette)
```

## Load data
The data used in this Rmarkdown document comes from **SCGTEST_32** dataset.
```{r}
# Load all metadata information
metadata <- read.csv(path_project_metadata)
metadata <- metadata %>% 
  filter(subproject == "SCGTEST_32" & tissue == "Colon")

print("Libraries/Samples metadata")
DT::datatable(metadata, options = list(scrollX = TRUE))

# Load Seurat object
seurat_obj <- readRDS(paste0(path_r_objects, "/scgtest32_m_COLON_healthy_downs_clustering_resolutions.rds"))
DefaultAssay(seurat_obj) <- "RNA"
Idents(seurat_obj) <- "RNA_snn_res.0.3"
seurat_obj

# Load biomarkers
seurat_obj_markers <- readRDS(paste0(path_r_objects, "/biomarkers_scgtest32_m_COLON_healthy_downs_resolution0.3.rds"))
```


# Explore cell clusters
To visualize the cell clusters, there are a few different dimensionality reduction techniques, such as t-SNE and UMAP. These methods require to input a number of PCA dimensions (same number as the one used for the clustering). We will not run UMAP again as it was computed in the previous notebook.
```{r fig.width=14, fig.height=10}
# Clustering overview
gg_umap_clustering <- DimPlot(object = seurat_obj,
        group.by = resolution_name,
        label = TRUE,
        label.size = 6,
        cols = color_palette,
        pt.size=0.5
        ) & NoLegend()
gg_umap_clustering
```

Now, we will also explore clusters across conditions.
```{r fig.width=7, fig.height=10}
# Clusters across tissues
gg_clustering_by_umap <- DimPlot(
  object = seurat_obj,
  group.by = resolution_name,
  split.by = "sample_protocol",
  label = FALSE,
  cols = color_palette,
  ncol = 1
  )

gg_clustering_by_umap
```


## Cell clusters data exploration
```{r}
# Number of cells / cluster
table(seurat_obj$RNA_snn_res.0.3)

# Number of cells per cluster vs confounder variables
table(seurat_obj$library_name, seurat_obj$RNA_snn_res.0.3)
```


## Assessing sources of uninteresting variation
Next, we will explore additional metrics, such as the number of UMIs and genes per cell, mitochondrial / ribosomal gene expression and scrublet doublet scores on the UMAP embedding. 

```{r fig.width=14, fig.height=10}
qc_metrics <-  c("nCount_RNA", "nFeature_RNA", "pct_mt", "pct_rb")

gg_qc_umap <- FeaturePlot(
  seurat_obj,
  reduction = "umap",
  features = qc_metrics,
  cols = c("lightgrey", "darkblue"), 
  order = TRUE,
  ncol = 2)

gg_qc_umap
```

```{r fig.width=10, fig.height=15}
gg_qc_vln <- VlnPlot(
  seurat_obj,
  features = qc_metrics,
  cols = color_palette, 
  pt.size = 0,
  ncol = 1)

gg_qc_vln
```

Finally, we will also check the cell-cycle phase by assessing S-phase and G2M-phase markers signature.
```{r fig.width=10, fig.height=10}
gg_phase_vln <- VlnPlot(
  seurat_obj,
  features = c("S.Score", "G2M.Score"),
  cols = color_palette, 
  pt.size = 0,
  ncol = 1)

gg_phase_vln
```

# Explore cluster biomarkers
```{r}
DefaultAssay(seurat_obj) <- "RNA"
```

## Table
```{r}
# Visualize top markers per cluster
top_all_markers <- seurat_obj_markers %>%
  group_by(cluster) %>%
  slice_max(n = 10, order_by = avg_log2FC)

top_all_markers
DT::datatable(top_all_markers, options = list(scrollX = TRUE))
```

## Dotplot
```{r fig.width=17, fig.height=6}
# Visualize top markers per cluster
top_all_markers <- seurat_obj_markers %>%
  group_by(cluster) %>%
  slice_max(n = 4, order_by = avg_log2FC)

top_genemarkers_list <- unique(top_all_markers$gene)

DotPlot(
  seurat_obj,
  features = top_genemarkers_list,
  dot.scale = 6) +
  RotatedAxis()
```

## FeaturePlot and ViolinPlot
With the cells clustered and based on the previous cell-type markers, we will select the most informative gene markers for each cluster and some well-known cell markers and project them in the UMAP. Basically, we are looking for consistency of expression of the markers across the clusters (e.g. if there are two markers for a cell type and only one of them is expressed in a cluster, we cannot reliably assign that cluster to the cell type).
```{r}
# Selected gene markers
selected_markers <- seurat_obj_markers %>%
  group_by(cluster) %>%
  slice_max(n = 4, order_by = avg_log2FC) %>%
  group_split()
```

```{r fig.width=7, fig.height=5}
# Feature plots
purrr::map(selected_markers, function(cluster) {
  FeaturePlot(seurat_obj,
              features = cluster$gene,
              label = TRUE,
              label.size = 3,
              ncol = 2)
})
```


# Annotation
According to the previous cell-type identities enriched in several gene markers, we will establish the following annotation:

```{r}
seurat_obj <- RenameIdents(
  seurat_obj,
 "0" = "Low quality cells high MT%",
 "1" = "Intestinal Epithelial cells 1",
 "2" = "Low quality cells high RB% 1",
 "3" = "Low quality cells high RB% 2",
 "4" = "Proliferative cells",
 "5" = "Intestinal Epithelial cells 2",
 "6" = "B cells",
 "7" = "Intestinal Epithelial cells Emp1+",
 "8" = "Fibroblasts",
 "9" = "Lymphatic Endothelial cells",
 "10" = "Immune cells ?",
 "11" = "T cells",
 "12" = "Endothelial cells",
 "13" = "Senescent Fibroblasts",
 "14" = "Low quality cells Hba+Hbb+",
 "15" = "Intestinal Epithelial cells 3",
 "16" = "Myeloid cells",
 "17" = "Intestinal Epithelial cells Hormones+"
 )

# Creating new annotation variable
seurat_obj$celltypes_1 <- seurat_obj@active.ident
seurat_obj$celltypes_1 <- factor(seurat_obj$celltypes_1,
                                 levels = sort(levels(seurat_obj$celltypes_1)))
# Exploring number of celltypes
table(seurat_obj$celltypes_1)
```

## UMAP celltypes
```{r fig.width=16, fig.height=10}
# Clustering overview
gg_umap_clustering_annot <- DimPlot(object = seurat_obj,
        group.by = "celltypes_1",
        label = TRUE,
        label.size = 5,
        repel = TRUE,
        cols = color_palette
        ) +
  labs(title = "Cell types - level1")
gg_umap_clustering_annot

ggsave(filename = paste0(path_r_figs, "/scgtest32_m_COLON_healthy_downs_cells1_clustering_annotation_umap.png"),
       plot = gg_umap_clustering_annot)
```

Now, we will also explore clusters across sample.
```{r fig.width=15, fig.height=5}
DimPlot(
  object = seurat_obj,
  group.by = "celltypes_1",
  split.by = "sample_protocol",
  label = FALSE,
  cols = color_palette
  )
```

## Cell proportions
```{r fig.width=12, fig.height=6}
gg_cell_num <- dittoBarPlot(seurat_obj, 
                            "celltypes_1",
                            group.by = "library_name",
                            scale = "count",
                            retain.factor.levels = TRUE,
                            legend.show = FALSE,
                            color.panel = color_palette)

gg_cell_prop <- dittoBarPlot(seurat_obj, 
                             "celltypes_1",
                             group.by = "library_name",
                             scale = "percent",
                            retain.factor.levels = TRUE,
                             color.panel = color_palette)

gg_cell_num + gg_cell_prop
```


# Save Seurat object
```{r}
# Save lineage annotations
saveRDS(seurat_obj$celltypes_1,
        paste0(path_r_objects, "/scgtest32_m_COLON_healthy_downs_cells1_names.rds"))

# Save Seurat object
saveRDS(seurat_obj, paste0(path_r_objects, "/scgtest32_m_COLON_healthy_downs_cells1_clustering_annotation.rds"))
#seurat_obj <- readRDS(paste0(path_r_objects, "/scgtest32_m_COLON_healthy_downs_cells1_clustering_annotation.rds"))
```


# Extra analysis
```{r}
min_new <- 250

seurat_obj@meta.data %>%
  ggplot(aes(x = nFeature_RNA, fill = sample_protocol, color = sample_protocol)) +
    geom_density(alpha = 0.2) + 
    scale_x_log10() +
    labs(x = "Library complexity (Number of Detected Genes)",
         y = "Density") +
    geom_vline(xintercept = min_new, linetype = "dashed", color = "orange") +
    theme(axis.title = element_text(size = 12),
          axis.text = element_text(size = 10),
          legend.position = "bottom") +
    guides(color = FALSE, fill = guide_legend(title = "Condition"))
```



## Cells nFeature_RNA < 250

```{r fig.width=15, fig.height=5}
seurat_obj_sub <- subset(seurat_obj, nFeature_RNA < min_new)
seurat_obj_sub

DimPlot(
  object = seurat_obj_sub,
  group.by = "celltypes_1",
  split.by = "sample_protocol",
  label = FALSE,
  cols = color_palette
  )


seurat_obj_sub$celltypes_1 <- as.character(seurat_obj_sub$celltypes_1)
seurat_obj_sub$celltypes_1 <- as.factor(seurat_obj_sub$celltypes_1)

gg_cell_num <- dittoBarPlot(seurat_obj_sub, 
                            "celltypes_1",
                            group.by = "library_name",
                            scale = "count",
                            retain.factor.levels = TRUE,
                            legend.show = FALSE,
                            color.panel = color_palette)

gg_cell_prop <- dittoBarPlot(seurat_obj_sub, 
                             "celltypes_1",
                             group.by = "library_name",
                             scale = "percent",
                            retain.factor.levels = TRUE,
                             color.panel = color_palette)

gg_cell_num + gg_cell_prop
```




## Removing two exclusive clusters
```{r fig.width=15, fig.height=5}
seurat_obj_sub <- subset(seurat_obj, 
                         celltypes_1 != "Low quality cells high MT%" &
                         celltypes_1 != "Low quality cells high RB% 1" &
                         celltypes_1 != "Low quality cells high RB% 2" &
                        celltypes_1 != "Low quality cells Hba+Hbb+" &
                          celltypes_1 != "Immune cells ?" 
                           )
seurat_obj_sub

DimPlot(
  object = seurat_obj_sub,
  group.by = "celltypes_1",
  split.by = "sample_protocol",
  label = FALSE,
  cols = color_palette
  )


seurat_obj_sub$celltypes_1 <- as.character(seurat_obj_sub$celltypes_1)
seurat_obj_sub$celltypes_1 <- as.factor(seurat_obj_sub$celltypes_1)

gg_cell_num <- dittoBarPlot(seurat_obj_sub, 
                            "celltypes_1",
                            group.by = "library_name",
                            scale = "count",
                            retain.factor.levels = TRUE,
                            legend.show = FALSE,
                            color.panel = color_palette)

gg_cell_prop <- dittoBarPlot(seurat_obj_sub, 
                             "celltypes_1",
                             group.by = "library_name",
                             scale = "percent",
                            retain.factor.levels = TRUE,
                             color.panel = color_palette)

gg_cell_num + gg_cell_prop
```


```{r}
counts_new <- 500

seurat_obj@meta.data %>%
  ggplot(aes(x = nCount_RNA, fill = sample_protocol, color = sample_protocol)) +
    geom_density(alpha = 0.2) + 
    scale_x_log10() +
    labs(x = "Library Size (Total UMIs)",
         y = "Density") +
    geom_vline(xintercept = counts_new, linetype = "dashed", color = "red") +
    theme(axis.title = element_text(size = 12),
          axis.text = element_text(size = 10),
          legend.position = "bottom") +
    guides(color = FALSE, fill = guide_legend(title = "Condition"))
```


## Cells nCount_RNA < 500

```{r fig.width=15, fig.height=5}
seurat_obj_sub <- subset(seurat_obj, nCount_RNA < counts_new)
seurat_obj_sub

DimPlot(
  object = seurat_obj_sub,
  group.by = "celltypes_1",
  split.by = "sample_protocol",
  label = FALSE,
  cols = color_palette
  )


seurat_obj_sub$celltypes_1 <- as.character(seurat_obj_sub$celltypes_1)
seurat_obj_sub$celltypes_1 <- as.factor(seurat_obj_sub$celltypes_1)

gg_cell_num <- dittoBarPlot(seurat_obj_sub, 
                            "celltypes_1",
                            group.by = "library_name",
                            scale = "count",
                            retain.factor.levels = TRUE,
                            legend.show = FALSE,
                            color.panel = color_palette)

gg_cell_prop <- dittoBarPlot(seurat_obj_sub, 
                             "celltypes_1",
                             group.by = "library_name",
                             scale = "percent",
                            retain.factor.levels = TRUE,
                             color.panel = color_palette)

gg_cell_num + gg_cell_prop
```


# Session Info
```{r}
sessionInfo()
```
