---
title: "Annotation"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction
After the pre-processing steps done in previous notebooks, we will now start with the downstream data analysis in order to extract biological insights and describe the underlying biology.

## Objective
In this Rmarkdown document, we are going to assign cell type clusters into different cell-types clusters based on the similarity of their gene expression profiles. Then, we will perform again process each object to define cell-type specific clusters with more resolution.

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
library(DT)
library(dittoSeq)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_project_metadata <- here::here("01_cellranger_mapping/data/FIXnCUT_metadata.csv")
path_r_objects <- here::here("03_clustering_annotation/results/R_objects")
path_r_tables <- here::here("03_clustering_annotation/results/tables")
path_r_figs <- here::here("03_clustering_annotation/results/figs")

# Functions
source(here::here("bin/utils.R"))

## Define resolution of choice
resolution <- 0.4
resolution_name <- paste0("RNA_snn_res.", resolution)

# Color palette
set.seed(1234) # for reproducibility
color_palette <- Polychrome::createPalette(35, c("#fc6060", "#74f774", "#7c7cfc"))
names(color_palette) <- NULL
#Polychrome::swatch(color_palette)
```

## Load data
The data used in this Rmarkdown document comes from **SCGTEST_32** dataset.
```{r}
# Load all metadata information
metadata <- read.csv(path_project_metadata)
metadata <- metadata %>% 
  filter((subproject == "SCGTEST_32")
         & tissue == "Colon")

print("Libraries/Samples metadata")
DT::datatable(metadata, options = list(scrollX = TRUE))

# Load Seurat object
seurat_obj <- readRDS(paste0(path_r_objects, "/FIXnCUT_MouseColon_filtered_post_clustering_resolutions.rds"))
DefaultAssay(seurat_obj) <- "RNA"
Idents(seurat_obj) <- "RNA_snn_res.0.4"
seurat_obj

# Load biomarkers
seurat_obj_markers <- readRDS(paste0(path_r_objects, "/biomarkers_FIXnCUT_MouseColon_filtered_post_resolution0.4.rds"))
```


# Explore cell clusters
To visualize the cell clusters, there are a few different dimensionality reduction techniques, such as t-SNE and UMAP. These methods require to input a number of PCA dimensions (same number as the one used for the clustering). We will not run UMAP again as it was computed in the previous notebook.
```{r fig.width=14, fig.height=10}
# Clustering overview
gg_umap_clustering <- DimPlot(object = seurat_obj,
        group.by = resolution_name,
        label = TRUE,
        label.size = 6,
        cols = color_palette,
        pt.size=0.5
        ) & NoLegend()
gg_umap_clustering
```

Now, we will also explore clusters across conditions.
```{r fig.width=14, fig.height=5}
# Clusters across tissues
gg_clustering_by_umap <- DimPlot(
  object = seurat_obj,
  group.by = resolution_name,
  split.by = "sample_protocol",
  label = FALSE,
  cols = color_palette,
  ncol = 3
  )

gg_clustering_by_umap
```


## Cell clusters data exploration
```{r}
# Number of cells / cluster
table(seurat_obj@active.ident)

# Number of cells per cluster vs confounder variables
table(seurat_obj$library_name, seurat_obj@active.ident)


# Creating new annotation variable
seurat_obj$celltypes_old <- seurat_obj$celltypes
seurat_obj$celltypes <- as.character(seurat_obj@active.ident)
Idents(seurat_obj) <- "celltypes"
```


## Assessing sources of uninteresting variation
Next, we will explore additional metrics, such as the number of UMIs and genes per cell, mitochondrial / ribosomal gene expression and scrublet doublet scores on the UMAP embedding. 

```{r fig.width=14, fig.height=15}
qc_metrics <-  c("nCount_RNA", "nFeature_RNA", "pct_mt", "pct_rb", "S.Score", "G2M.Score")

gg_qc_umap <- FeaturePlot(
  seurat_obj,
  reduction = "umap",
  features = qc_metrics,
  cols = c("lightgrey", "darkblue"), 
  order = TRUE,
  ncol = 2)

gg_qc_umap
```

```{r fig.width=10, fig.height=15}
gg_qc_vln <- VlnPlot(
  seurat_obj,
  features = qc_metrics,
  cols = color_palette, 
  pt.size = 0,
  ncol = 1)

gg_qc_vln
```

Finally, we will also check the cell-cycle phase by assessing S-phase and G2M-phase markers signature.
```{r fig.width=10, fig.height=10}
gg_phase_vln <- VlnPlot(
  seurat_obj,
  features = c("S.Score", "G2M.Score"),
  cols = color_palette, 
  pt.size = 0,
  ncol = 1)

gg_phase_vln
```

```{r fig.width=10, fig.height=5}
VlnPlot(
  seurat_obj,
  features = c("scrublet_doublet_scores"),
  cols = color_palette, 
  pt.size = 0,
  ncol = 1)
```

# Explore cluster biomarkers
```{r}
DefaultAssay(seurat_obj) <- "RNA"
```

## Table
```{r}
# Visualize top markers per cluster
top_all_markers <- seurat_obj_markers %>%
  group_by(cluster) %>%
  slice_max(n = 10, order_by = avg_log2FC)

top_all_markers
DT::datatable(top_all_markers, options = list(scrollX = TRUE))
```

## Dotplot
```{r fig.width=25, fig.height=7}
# Visualize top markers per cluster
top_all_markers <- seurat_obj_markers %>%
  group_by(cluster) %>%
  slice_max(n = 4, order_by = avg_log2FC)

top_genemarkers_list <- unique(top_all_markers$gene)

DotPlot(
  seurat_obj,
  features = top_genemarkers_list,
  dot.scale = 8) +
  RotatedAxis()
```

<!-- ## FeaturePlot and ViolinPlot -->
<!-- With the cells clustered and based on the previous cell-type markers, we will select the most informative gene markers for each cluster and some well-known cell markers and project them in the UMAP. Basically, we are looking for consistency of expression of the markers across the clusters (e.g. if there are two markers for a cell type and only one of them is expressed in a cluster, we cannot reliably assign that cluster to the cell type). -->
<!-- ```{r} -->
<!-- # Selected gene markers -->
<!-- selected_markers <- seurat_obj_markers %>% -->
<!--   group_by(cluster) %>% -->
<!--   slice_max(n = 4, order_by = avg_log2FC) %>% -->
<!--   group_split() -->
<!-- ``` -->

<!-- ```{r fig.width=7, fig.height=5} -->
<!-- # Feature plots -->
<!-- purrr::map(selected_markers, function(cluster) { -->
<!--   FeaturePlot(seurat_obj, -->
<!--               features = cluster$gene, -->
<!--               label = TRUE, -->
<!--               label.size = 3, -->
<!--               ncol = 2) -->
<!-- }) -->
<!-- ``` -->


## Gene expression

```{r fig.width=30, fig.height=15}
# Immune fraction
FeaturePlot(seurat_obj,
            features = c("Ptprc", # CD45+, immune cells
                         "Cd3d", # T cells
                         "Cd4", "Foxp3", # CD4+ T cells, Treg
                         "Cd8a", "Nkg7", # CD8+ T cells, cytotoxic
                         "Klrb1c", # NK cells
                         "Cd19", "Cd79a", # B cells
                         "Ms4a1", "Jchain", # Naive vs Plasma cells
                         "Lyz", "Cd14", "Plac8", "Cd68", # Mono/macro
                         "Itgax", "Mertk", "Siglecf", "Marco", # Alveolar macrophages
                         "S100a8", "S100a9", # Neutrophils
                         "Cd200r3", # Mast cells
                         "H2-Aa", # MHC- antigen presenting
                         "Cd34", "Sox4" # progenitor
                         ),
            order = T,
            ncol = 6)
```

```{r fig.width=25, fig.height=15}
# Non-immune fraction
FeaturePlot(seurat_obj,
            features = c("Epcam", "Krt", "Krt18", # Epithelial cells
                        "Pecam1", "Lyve1", # Endothelial cells
                        "Lypd8", "Lgals4", # Intestinal cells
                        "Dpt", "Ly6a", "Pi16", # Univ Fibroblasts
                        "Clu", "Slpi", # Mesothelial cells
                        "Rgs5", "Acta2", # Pericytes
                        "Bmp4", "Bmp5", # Telocytes
                        "Cfd", "Car3", # Adipocytes
                        "Top2a", "Stmn1", "Mki67"
                         ),
            order = T,
            ncol = 5)
```


### Myeloid

```{r fig.width=20, fig.height=15}
mono_genes <- c("Cd14", "Itgam", "Ccr2", "Sell", "Cd36",
                "Ly6c", "Cxcr1", "Itgal", "Aif1", "Cd68", "Cd163", "Plac8", "Fcgr1", "Apoe", "C1qa") 
# classical CD14/CCR2/Cd11b == ITGAM, non-classical CD16 == FCGR3A, intermediate 
FeaturePlot(seurat_obj,
            features = mono_genes,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 4
)

VlnPlot(
  seurat_obj,
  features = mono_genes,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 4)
```



### B Lymphocytes
```{r fig.width=20, fig.height=15}
Bnaive_genes <- c("Ms4a1", "Cd79a", "Cd79b", 
                  "Cd69", "Fcer2", "Cd1c", "Cd48", "Cd81", "Jun", "Cd86",
                  "Ptprc", "Ebf1", "Cd19", "Bst2" #pre-pro-B cell
                  )

FeaturePlot(seurat_obj,
            features = Bnaive_genes,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 4
)

VlnPlot(
  seurat_obj,
  features = Bnaive_genes,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 4)
```


```{r fig.width=20, fig.height=15}
Plasma_genes <- c("Cd27", "Jchain", "Xbp1", "Cd38", "Rf4", "Ighd", "Ighm", "Mki67", "Mzb1", "Slamf7", "Ssr4")
# RF4/XBP1/PRDM1 = early PC /PCprec
# SLAMF7/MZB1/SSR4+ Ig = Mature
# when proliferating = Plasmablast

FeaturePlot(seurat_obj,
            features = Plasma_genes,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 4
)

VlnPlot(
  seurat_obj,
  features = Plasma_genes,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 4)
```


### T Lymphocytes

```{r fig.width=30, fig.height=15}
T_genes <- c("Cd3d", "Cd3e",# T cells
            "Cd4", "Foxp3", # CD4+ T cells, Treg
            "Cd8a", "Cd8b1", "Nkg7", "Gzmk", "Gzmb", # CD8+ T cells, cytotoxic
            "Sell", "Ccr7", "Lef1", "Cd27", "Cd69"
             )

FeaturePlot(seurat_obj,
            features = T_genes,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.25,
            ncol = 5)

VlnPlot(
  seurat_obj,
  features = T_genes,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 5)
```

### Innate cells

```{r fig.width=15, fig.height=15}
NK_genes <- c("Gpr183", "Il7r", "Gzmb", "Gzmk", "Klrc1", "Klrd1", "Ncam1") 
# High CD16 == FCGR3A, CD18 == ITGB2, PRF1, FCRL6, NCR3,CD57 == B3GAT1
FeaturePlot(seurat_obj,
            features = NK_genes,
            #order = TRUE,
            #split.by = "tissue",
            pt.size = 0.25,
            ncol = 3
)

VlnPlot(
  seurat_obj,
  features = NK_genes,
  cols = color_palette, 
  pt.size = 0.1, 
  ncol = 3)
```


### Colon-specific celltypes
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6022292/

```{r fig.width=20, fig.height=10}
colon_cells <- c("Col1a1", "Clu", "Ascl2", "Emp1", "Lgr5", "Mep1a") 

FeaturePlot(seurat_obj,
            features = colon_cells,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 4
)

VlnPlot(
  seurat_obj,
  features = colon_cells,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 4)
```


```{r fig.width=20, fig.height=15}
CBCLgr5_cells <- c("Mcm7", "Lamc1", "Pvr", "Luzp1", "Itgb1", "Itga2", "Cd44", "Msn", "Lama3",
                   "Lgr5", "Lgr4", "Ascl2", "Olfm4", "Mex3a", "Axin2")

FeaturePlot(seurat_obj,
            features = CBCLgr5_cells,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 5
)

VlnPlot(
  seurat_obj,
  features = CBCLgr5_cells,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 5)
```


```{r fig.width=20, fig.height=5}
enteroendocrine_cells <- c("Chga", "Chgb", "Cpe", "Gfra3") 

FeaturePlot(seurat_obj,
            features = enteroendocrine_cells,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 5
)

VlnPlot(
  seurat_obj,
  features = enteroendocrine_cells,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 5)
```


```{r fig.width=20, fig.height=5}
stem_cells <- c("Tert", "Hopx", "Krt19", "Lrig1", "Mex3a") 

FeaturePlot(seurat_obj,
            features = stem_cells,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 5
)

VlnPlot(
  seurat_obj,
  features = stem_cells,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 5)
```

```{r fig.width=20, fig.height=5}
secretoryprog_cells <- c("Dll1", "Mist1", "Atoh1") 

FeaturePlot(seurat_obj,
            features = secretoryprog_cells,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 5
)

VlnPlot(
  seurat_obj,
  features = secretoryprog_cells,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 5)
```

```{r fig.width=20, fig.height=5}
enterocytes_cells <- c("Alpi",
                       "Mep1a", "Fgf15", "Clec2h", # distal
                       "Lct", "Cbr1", "Ephx2" # proximal
                       ) 

FeaturePlot(seurat_obj,
            features = enterocytes_cells,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 5
)

VlnPlot(
  seurat_obj,
  features = enterocytes_cells,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 5)
```

```{r fig.width=20, fig.height=5}
TA_cells <- c("Stmn1", "Tubb5", "Cenpf", "Nuscup1") 

FeaturePlot(seurat_obj,
            features = TA_cells,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 5
)

VlnPlot(
  seurat_obj,
  features = TA_cells,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 5)
```

```{r fig.width=20, fig.height=5}
goblet_cells <- c("Tff3", "Manf", "Ccl9", "Muc2") 

FeaturePlot(seurat_obj,
            features = goblet_cells,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 5
)

VlnPlot(
  seurat_obj,
  features = goblet_cells,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 5)
```


```{r fig.width=20, fig.height=5}
tuft_cells <- c("Lrmp", "Dclk1", "Cd24a") 

FeaturePlot(seurat_obj,
            features = tuft_cells,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 5
)

VlnPlot(
  seurat_obj,
  features = tuft_cells,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 5)
```

```{r fig.width=20, fig.height=5}
paneth_cells <- c("Lyz1", "Mptx1", "Mptx2", "Defa-r1") 

FeaturePlot(seurat_obj,
            features = paneth_cells,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 5
)

VlnPlot(
  seurat_obj,
  features = paneth_cells,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 5)
```




### Others

```{r fig.width=20, fig.height=20}
endothelial_cells <- c("Pecam1",
                       "Kit", "Gpihbp1", "Itga6", "Peg3", # gCap
                       "Emp2", "Car4", "Igfbp7", "Ednrb", "Inhba", # aCap
                       "Cxcl12", "Fbln5", "Ltbp4", "Eln", "Mgp", "Vwf", "Slc6a2", "Fbln2", "Eln", # Art & Vein
                       "Ccl21a", "Mmrn1", "Fgl2", "Nrp2", "Nts" # Lymph
                       ) 

FeaturePlot(seurat_obj,
            features = endothelial_cells,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 5
)

VlnPlot(
  seurat_obj,
  features = endothelial_cells,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 5)
```


```{r fig.width=20, fig.height=10}
epithelial_cells <- c("Epcam", "Col1a1", "Col13a1", "Col14a1", # fibros 
                      "Postn", "Rgs5", "Acta2", "Crip1", "Saa1", "Car4") 

FeaturePlot(seurat_obj,
            features = epithelial_cells,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 4
)

VlnPlot(
  seurat_obj,
  features = epithelial_cells,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 4)
```

```{r fig.width=20, fig.height=10}
mesothelial_cells <- c("Dcn", "Col1a2", "Msln") 

FeaturePlot(seurat_obj,
            features = mesothelial_cells,
            #split.by = "therapy",
            #order = T,
            pt.size = 0.5,
            ncol = 4
)

VlnPlot(
  seurat_obj,
  features = mesothelial_cells,
  cols = color_palette, 
  pt.size = 0, 
  ncol = 4)
```


```{r fig.width=8, fig.height=5}
# Clustering overview
DimPlot(object = seurat_obj,
        group.by = "celltypes_old",
        repel = TRUE,
        cols = color_palette
        ) +
  labs(title = "Cell types")

DimPlot(object = seurat_obj,
        group.by = "celltypes",
        repel = TRUE,
        cols = color_palette
        ) +
  labs(title = "Cell types")
```

# Annotation
According to the previous cell-type identities enriched in several gene markers, we will establish the following annotation:

```{r}
# GENERAL
Idents(seurat_obj) <- "celltypes"

seurat_obj <- RenameIdents(
  seurat_obj,
 "0" = "Distal Colonocytes", # Mep1a
 "1" = "Proximal Colonocytes", # Cbr1
 "2" = "Transit-amplifying (TA) cells", # Tubb5, Stmn1, Cenpf, Nuscup1
 "3" = "Distal Colonocytes", # Mep1a
 "4" = "Mature B cells", 
 "5" = "Distal Colonocytes Emp1+", #Emp1, Guca2a
 "6" = "Lymphatic Endothelial cells", # Lyve1
 "7" = "Goblet cells", # Tff3, Manf, Ccl9
 "8" = "Immature B cells", 
  "9" = "Myofibroblasts",
 "10" = "Paneth cells", # Mptx1
 "11" = "T-NK cells",
 "12" = "Endothelium", # Fabp4, Pecam1
 "13" = "Myeloid cells", # Cd14, Apoe, C1qa
 "14" = "Stromal cells", # Dcn, 
 "15" = "Tuft cells", # Lrmp, Dclk1, Cd24a
 "16" = "Enteroendocrine (EEC) cells" # Chgb, Cpe, Gfra3
  )

# Creating new annotation variable
seurat_obj$celltypes <- as.factor(as.character(seurat_obj@active.ident))
seurat_obj$celltypes <- factor(seurat_obj$celltypes, levels = sort(levels(seurat_obj$celltypes)))
```


```{r}
# Exploring number of celltypes
table(seurat_obj$celltypes, seurat_obj$library_name)
table(seurat_obj$sample_protocol)
```


## UMAP celltypes
```{r fig.width=16, fig.height=10}
# Clustering overview
gg_umap_clustering_annot <- DimPlot(object = seurat_obj,
        group.by = "celltypes",
        repel = TRUE,
        cols = color_palette
        ) +
  labs(title = "Cell types")
gg_umap_clustering_annot

ggsave(filename = paste0(path_r_figs, "/FIXnCUT_MouseColon_filtered_post_clustering_annotation_umap.png"),
       plot = gg_umap_clustering_annot)
```

Now, we will also explore clusters across sample.
```{r fig.width=15, fig.height=5}
DimPlot(
  object = seurat_obj,
  group.by = "celltypes",
  split.by = "library_name",
  label = FALSE,
  cols = color_palette
  )
```

## Cell proportions
```{r fig.width=12, fig.height=6}
gg_cell_num <- dittoBarPlot(seurat_obj, 
                            "celltypes",
                            group.by = "library_name",
                            scale = "count",
                            retain.factor.levels = TRUE,
                            legend.show = FALSE,
                            color.panel = color_palette)

gg_cell_prop <- dittoBarPlot(seurat_obj, 
                             "celltypes",
                             group.by = "library_name",
                             scale = "percent",
                            retain.factor.levels = TRUE,
                             color.panel = color_palette)

gg_cell_num + gg_cell_prop
```

## Save Seurat object
```{r}
# Save lineage annotations
saveRDS(seurat_obj$celltypes,
        paste0(path_r_objects, "/FIXnCUT_MouseColon_filtered_post_cellsnames.rds"))

# Save Seurat object
saveRDS(seurat_obj, paste0(path_r_objects, "/FIXnCUT_MouseColon_filtered_post_clustering_annotation.rds"))
#seurat_obj <- readRDS(paste0(path_r_objects, "/FIXnCUT_MouseColon_filtered_clustering_lineageannotation.rds"))
```



# Session Info
```{r}
sessionInfo()
```
