---
title: "Annotation (by cell type) from subclusters"
author: "Laura Jiménez Gracia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

# Introduction

## Objective
In this Rmarkdown document, we are going to merge cell type identities annotated by independent cell-lineage analysis into a single one.

# Pre-processing
## Libraries
```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
library(dittoSeq)
library(DT)
```

## Parameters
Here we will define and load objects and functions that will be used throughout the document.
```{r}
# Paths
path_project_metadata <- here::here("01_cellranger_mapping/data/POST_12_metadata.csv")
path_samples_metadata <- here::here("01_cellranger_mapping/data/POST_12_metadata_samples_all.csv")
path_r_objects <- here::here("03_clustering_annotation/results/R_objects")
path_r_tables <- here::here("03_clustering_annotation/results/tables")
path_r_figs <- here::here("03_clustering_annotation/results/figs")

# Functions
source(here::here("bin/utils.R"))

# Parameters
confounder_variables <- c("library_name", "sample_id", "genotype", "treatment")
confounder_names <- c("Library", "Sample", "Genotype", "Treatment")

metadata_variables_confounder <- c("nCount_RNA", "nFeature_RNA", "gem_id", 
                                   "pct_mt", "pct_rb", "Phase",
                                   "library_name", "sample_id", "genotype", "treatment")

# Color palette
set.seed(1234) # for reproducibility
color_palette <- Polychrome::createPalette(60, c("#fc6060", "#74f774", "#7c7cfc"))
names(color_palette) <- NULL
#Polychrome::swatch(color_palette)
```

## Load data
The data used in this Rmarkdown document comes from **POST_12** dataset.
```{r}
# Load all metadata information
metadata <- read.csv(path_project_metadata)

# Merging metadata_all tables
metadata <- metadata %>% filter(type == "cDNA" & fraction == "CD45-")

print("Libraries/Samples metadata")
DT::datatable(metadata, options = list(scrollX = TRUE))

# Load Seurat object
seurat_obj <- readRDS(paste0(path_r_objects, "/post12_cd45neg_cells1_clustering_annotation.rds"))
Idents(seurat_obj) <- "celltypes_1"
seurat_obj
```


## Integrate cell-types identity
```{r}
seurat_obj$celltypes_1 <- as.character(seurat_obj$celltypes_1)
seurat_obj$celltypes_2 <- as.character(seurat_obj$celltypes_1)


celltypes_1_list <- c("Stromal cells", "Epidermal cells")
celltypes_1_filenames <- c("stromal", "epidermal")

# Load & add cell-type annotations
for (i in 1:length(celltypes_1_list)) {
  celltype <- celltypes_1_list[i]
  filename <- celltypes_1_filenames[i]
  file_path <- paste0(path_r_objects, "/post12_cd45neg_", filename, "_annotations.rds")
  # loading annotations
  annot <- readRDS(file_path)
  # adding cell_types labels
  seurat_obj$celltypes_2[seurat_obj$celltypes_1 == celltype] <- as.character(annot$celltypes_2)
}
```


## Remove CD45+ cells
```{r}
seurat_obj <- seurat_obj[, seurat_obj$celltypes_2 != "Immune cells (CD45+)"]
```

## Modify metadata
```{r}
# Correct cell2 to cell1 "name pairing "lineage"
# "Stromal cells cycling" belong to "Stromal cells" (lineage) but clustered with "Epidermal cells" 
seurat_obj$celltypes_1[seurat_obj$celltypes_2 == "Stromal cells cycling"] <- "Stromal cells"

# Reorder cells
seurat_obj$celltypes_1 <- as.factor(seurat_obj$celltypes_1)
seurat_obj$celltypes_1 <- factor(x = seurat_obj$celltypes_1,
                                     levels = c(
                                       "Stromal cells",
                                       "Epidermal cells",
                                       "Endothelial cells",
                                       "Lymphatic Endothelial cells",
                                       "Pericytes",
                                       "Schwann cells"
                                       )
                                 )
                                       
# Reorder cells
seurat_obj$celltypes_2 <- as.factor(seurat_obj$celltypes_2)
seurat_obj$celltypes_2 <- factor(x = seurat_obj$celltypes_2,
                                     levels = c(
                                       "Basal keratinocytes",
                                       "Basal keratinocytes cycling",
                                       "Supra-Basal keratinocytes",
                                       "Granular keratinocytes",
                                       "Cornified keratinocytes",
                                       "Dead keratinocytes",
                                       "Chondrocytes",
                                       "Dermal fibrocytes",
                                       "Reticular fibroblasts",
                                       "Universal fibroblasts",
                                       "Hair dermal sheath",
                                       "Inner ear fibroblasts",
                                       "Vascular smooth muscle",
                                       "Stromal cells cycling",
                                       "Skeletal muscle",
                                       "Endothelial cells",
                                       "Lymphatic Endothelial cells",
                                       "Pericytes",
                                       "Schwann cells"
                                     )
                                 )
```


# Explore data annotation
## Data visualization
### Cell types 1

```{r fig.width=14, fig.height=10}
# Clustering overview
gg_umap_clustering_celltypes_1 <- DimPlot(object = seurat_obj,
        group.by = "celltypes_1",
        label = FALSE,
        cols = color_palette,
        raster = FALSE
        ) +
  guides(color = guide_legend(override.aes = list(size = 5), ncol = 1))
gg_umap_clustering_celltypes_1

ggsave(filename = paste0(path_r_figs, "/post12_cd45neg_cells1_clustering_annotation_umap.png"),
       plot = gg_umap_clustering_celltypes_1)
```

Now, we will also explore lineage clusters across sample
```{r fig.width=25, fig.height=6}
DimPlot(
  object = seurat_obj,
  group.by = "celltypes_1",
  split.by = "sample_id",
  label = FALSE,
  cols = color_palette
  )  +
    theme(legend.position = "bottom") +
  guides(color = guide_legend(override.aes = list(size = 5), ncol = 5))
```

### Cell types 2
```{r fig.width=14, fig.height=10}
# Clustering overview
gg_umap_clustering_celltypes_2 <- DimPlot(
  object = seurat_obj,
        group.by = "celltypes_2",
        label = FALSE,
        cols = color_palette,
        raster = FALSE
        ) +
  guides(color = guide_legend(override.aes = list(size = 5), ncol = 1))
gg_umap_clustering_celltypes_2

ggsave(filename = paste0(path_r_figs, "/post12_cd45neg_cells2_clustering_annotation_umap.png"),
       plot = gg_umap_clustering_celltypes_2)
```

Now, we will also explore lineage clusters across sample
```{r fig.width=25, fig.height=6}
DimPlot(
  object = seurat_obj,
  group.by = "celltypes_2",
  split.by = "sample_id",
  label = FALSE,
  cols = color_palette
  ) + 
  theme(legend.position = "bottom") +
  guides(color = guide_legend(override.aes = list(size = 5), ncol = 6))
```


## Cell proportions
### Cell types 1
```{r fig.width=14, fig.height=5}
gg_cell_num <- dittoBarPlot(seurat_obj, 
                            "celltypes_1",
                            group.by = "sample_id",
                            scale = "count",
                            retain.factor.levels = TRUE,
                            legend.show = FALSE,
                            color.panel = color_palette)

gg_cell_prop <- dittoBarPlot(seurat_obj, 
                             "celltypes_1",
                             group.by = "sample_id",
                             scale = "percent",
                            retain.factor.levels = TRUE,
                             color.panel = color_palette)

gg_cell_num + gg_cell_prop
```

### Cell types 2
```{r fig.width=14, fig.height=7}
gg_cell_num <- dittoBarPlot(seurat_obj, 
                            "celltypes_2",
                            group.by = "sample_id",
                            scale = "count",
                            retain.factor.levels = TRUE,
                            legend.show = FALSE,
                            color.panel = color_palette)

gg_cell_prop <- dittoBarPlot(seurat_obj, 
                             "celltypes_2",
                             group.by = "sample_id",
                             scale = "percent",
                             retain.factor.levels = TRUE,
                             color.panel = color_palette) 

gg_cell_num + gg_cell_prop
```

## Number of cells
```{r}
length(table(seurat_obj$celltypes_1))
as.data.frame(table(seurat_obj$celltypes_1))

length(table(seurat_obj$celltypes_2))
as.data.frame(table(seurat_obj$celltypes_2))
```


```{r}
data.frame(table(seurat_obj$sample_id, seurat_obj$celltypes_1))
data.frame(table(seurat_obj$sample_id, seurat_obj$celltypes_2))
```


# Save data
```{r}
# Save metadata
seurat_obj_metadata_df <- seurat_obj@meta.data
saveRDS(seurat_obj_metadata_df, paste0(path_r_objects, "/post12_cd45neg_alldata_final_metadata.rds"))
openxlsx::write.xlsx(seurat_obj_metadata_df,
                     paste0(path_r_tables, "/post12_cd45neg_alldata_final_metadata.xlsx"))
```

```{r}
# Save annotated Seurat object
saveRDS(seurat_obj, paste0(path_r_objects, "/post12_cd45neg_alldata_final_clustering_annotation.rds"))
#seurat_obj <- readRDS(paste0(path_r_objects, "/post12_cd45neg_alldata_final_clustering_annotation.rds"))
```


# Session Info
```{r}
sessionInfo()
```
